\documentclass[../main.tex]{subfiles}

\begin{document}

\subsection{Basic Definitions}
\noindent

\paragraph{ }
\keyword{$\mathbb{R}$} Set of all real numbers. 

\keyword{$\mathbb{Z}$} Set of all integers. 

\keyword{$\mathbb{C}$} Set of all complex numbers. 

\paragraph{ }
\keyword{\textcolor{blue}{Scalars}}
Scalars are elements from $\mathbb{R}$, $\mathbb{Z}$, or $\mathbb{C}$.
\begin{itemize}
  \item Fundamental building blocks of \textit{vectors} and \textit{matrices}.
\end{itemize}
\paragraph{ }
\keyword{\textcolor{blue}{Vectors}} 
A vector is a one dimensional array of scalars. 
\begin{itemize}
  \item $\fR^n$ is the set of all vectors with real numbers of $m$ rows and $n$ columns.
  \item $[\bx]_{n}$ represents a real vector or size $n$.
\end{itemize}

\paragraph{ }
\keyword{\textcolor{blue}{Matrices}} 
A matrix is a multi-dimensional array of scalars. 
\begin{itemize}
  \item $\fR^{n \times m}$ is the set of all matrices with real numbers of size $n$.
  \item $\fR^{n \times n}$ is the set of all \textit{square} matrices with real numbers of size $n$.
  \item $[\bA]_{(m, n)}$ represents a real matrix with $m$ rows and $n$ columns.
  \item Vectors can be interpreted as matrices
    \begin{itemize}
      \item $\bx \in \fR^{m \times 1}$ represents a \textit{row} vector. $\begin{bmatrix}
          x_1 & \cdots & x_n \\
      \end{bmatrix}$
    \item $\bx \in \fR^{1 \times m}$ represents a \textit{column} vector. $\begin{bmatrix}
        x_1 \\ \vdots \\ x_n 
    \end{bmatrix}$
  \item $\bx \in \fR^{m}$ by default represents a \textit{row} vector (since we are using python for this course, which is a row-major language)
    \end{itemize}
\end{itemize}

\paragraph{ }
\keyword{\textcolor{blue}{\textit{Linear Systems}}} 
Linear systems are equations involving matrices and vectors - 
\begin{equation*}
  \by = \bA \bx, 
\end{equation*}

where, $\bx\in \fR^m, \bx\in \fR^n$  and $\bA \in \fR^{m \times n}$.

\paragraph{ }
\keyword{\textcolor{blue}{\textit{Linear Algebra}}} 

\textit{arithmetic}

\end{document}
