\documentclass[../../main.tex]{subfiles}

\begin{document}

\subsection{Basic Definitions}



\addDef{$\mathbb{R}$} {
    Set of all real numbers. 
}

\addDef{$\mathbb{Z}$} {
    Set of all integers. 
}

\addDef{$\mathbb{C}$} {
    Set of all complex numbers. 
}

\addDef{Scalars} {
    Scalars are elements from $\mathbb{R}$, $\mathbb{Z}$, or $\mathbb{C}$.
    \begin{itemize}
        \item Fundamental building blocks of \textit{vectors} and \textit{matrices}.
    \end{itemize}
}

\addDef{Vectors}{
    A vector is a one dimensional array of scalars. 
\begin{itemize}
  \item $\fR^n$ is the set of all vectors with real numbers of $m$ rows and $n$ columns.
  \item $[\bx]_{n}$ represents a real vector or size $n$.
\end{itemize}
}

\addDef{Matrices}{
    A matrix is a multi-dimensional array of scalars. 
    \begin{equation*}
        \bA \in \fR^{m \times n} = \begin{bmatrix}
            a_{11} & \cdots & a_{1n} \\
            \vdots & \ddots & \vdots \\
            a_{m1} & \cdots & a_{mn}
        \end{bmatrix}.
    \end{equation*}
    \begin{itemize}
      \item $\fR^{n \times m}$ is the set of all matrices with real numbers of size $n$.
      \item $\fR^{n \times n}$ is the set of all \textit{square} matrices with real numbers of size $n$.
      \item Vectors can be interpreted as matrices
        \begin{itemize}
          \item $\bx \in \fR^{m \times 1}$ represents a \textit{row} vector. $\begin{bmatrix}
              x_1 & \cdots & x_n \\
          \end{bmatrix}$
        \item $\bx \in \fR^{1 \times m}$ represents a \textit{column} vector. $\begin{bmatrix}
            x_1 \\ \vdots \\ x_n 
        \end{bmatrix}$
        \end{itemize}
    \end{itemize}
}

\addNote { 
    Matrices can be interpreted as a set of vectors stacked together. 
    \begin{equation*}     
            \bA =  \printmatrow{r}= \printmatcol{c}
    \end{equation*}
    \begin{itemize}
        \item $r_i = \bA[i]$ is a \textit{row vector}\index{row vector} representing the $i^{th}$ row of the matrix $\bA$.
        \item $c_i = \bA[:, i]$ is a \textit{column vector}\index{column vector} representing the $i^{th}$ column of the matrix $\bA$. 
    \end{itemize}
}


\end{document}



%\textit{Note: } $\bx \in \fR^{m}$ by default represents a \textit{row} vector (since we are using python for this course, which is a row-major language)

%\paragraph{ }
%\note{Linear Algebra} 
%
