\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Matrix Arithmetic}

We are all familiar with scalar arithmetic: addition, subtraction, multiplication, and
properties such as associativity and distributivity.  In the following sections we extend
these operations to vectors and matrices, building the algebraic machinery used throughout
scientific computing.

\subsubsection{Addition and Subtraction}

Matrix addition generalizes scalar addition by operating entry-wise.  The key constraint
is dimensional compatibility: two matrices can be added only when they share the same shape.

\addDef{Matrix Sum}{%
Let $\bA, \bB \in \fR^{m \times n}$.  Their sum is defined as
\[
\bA + \bB = \printmatsq{a} + \printmatsq{b} = \printmatsq{(a+b)},
\]
where each entry satisfies $(a+b)_{ij} = a_{ij} + b_{ij}$ for $1 \le i \le m$ and
$1 \le j \le n$.
}
\label{def:matrix-sum}

\addNote{%
For $\bA \in \fR^{m \times n}$ and $\bB \in \fR^{p \times q}$, the sum $\bA + \bB$ is
defined if and only if $m = p$ and $n = q$.
}

\addExcr{%
Refer to Def~\ref{def:matrix-sum}.
\begin{enumerate}
    \item Let $\bA = \begin{pmatrix}1 & 2 \\ 3 & 4\end{pmatrix}$ and
    $\bB = \begin{pmatrix}5 & 6 \\ 7 & 8\end{pmatrix}$.  Compute $\bA + \bB$ and
    $2\bA - \bB$ by hand.
    \item In NumPy, verify your answers using \texttt{np.array}.  Then check whether
    \texttt{np.array([1,2,3])} $+$ \texttt{np.array([1,2])} raises an error and explain why.
\end{enumerate}
}

\textBox{%
\textbf{Commutativity and Associativity of Matrix Addition}\\[2mm]
Let
\[
\bA = \begin{pmatrix} a & b \\ c & d \end{pmatrix} \quad \text{and} \quad
\bB = \begin{pmatrix} e & f \\ g & h \end{pmatrix}.
\]
Then,
\[
\bA + \bB = \begin{pmatrix} a+e & b+f \\ c+g & d+h \end{pmatrix}
\quad \text{and} \quad
\bB + \bA = \begin{pmatrix} e+a & f+b \\ g+c & h+d \end{pmatrix}.
\]
Since addition of real numbers is commutative, $\bA + \bB = \bB + \bA$.  Furthermore,
for any third matrix $\mathbf{C} = \begin{pmatrix} i & j \\ k & l \end{pmatrix}$,
\[
(\bA+\bB)+\mathbf{C} = \begin{pmatrix} (a+e)+i & (b+f)+j \\ (c+g)+k & (d+h)+l \end{pmatrix},
\]
and
\[
\bA+(\bB+\mathbf{C}) = \begin{pmatrix} a+(e+i) & b+(f+j) \\ c+(g+k) & d+(h+l) \end{pmatrix}.
\]
By associativity of real addition, these are identical.  Hence matrix addition is both
commutative and associative.
}

\addThm{$\fR^{m \times n}$ as a Vector Space}{%
The set $\fR^{m \times n}$ equipped with matrix addition and scalar multiplication forms a
real vector space of dimension $mn$.
}
\label{thm:Rmn-vecspace}

\addPrf{%
Each axiom follows from the corresponding property of $\fR$ applied entry-wise.
For \textbf{commutativity}:
\[
    (\bA + \bB)_{ij} = a_{ij} + b_{ij} = b_{ij} + a_{ij} = (\bB + \bA)_{ij}.
\]
The \textbf{additive identity} is $\mathbf{0}$ since $(\bA + \mathbf{0})_{ij} = a_{ij}$.
The \textbf{additive inverse} of $\bA$ is $-\bA$ (negate every entry), satisfying
$(\bA + (-\bA))_{ij} = 0$.

For the \textbf{dimension}: the $mn$ matrices $\bE^{(ij)}$ with a $1$ in position $(i,j)$
and zeros elsewhere are linearly independent, and every $\bA \in \fR^{m \times n}$
decomposes as $\bA = \sum_{i,j} a_{ij}\,\bE^{(ij)}$, so they form a basis of size $mn$.
}

\addDef{Scalar Multiplication}{%
Let $\bA \in \fR^{m \times n}$ and $c \in \fR$.  Then scalar multiplication is defined as
\[
c\bA = c \printmatsq{a} = \printmatsq{c a},
\]
where each entry satisfies $(ca)_{ij} = c \cdot a_{ij}$.
}
\label{def:scalar-mult}

\textBox{%
\textbf{Distributivity of Scalar Multiplication over Matrix Addition}\\[2mm]
Let $\alpha \in \fR$ and
\[
\bA = \begin{pmatrix} a & b \\ c & d \end{pmatrix}, \quad
\bB = \begin{pmatrix} e & f \\ g & h \end{pmatrix}.
\]
Then,
\[
\alpha(\bA + \bB)
= \alpha \begin{pmatrix} a+e & b+f \\ c+g & d+h \end{pmatrix}
= \begin{pmatrix} \alpha(a+e) & \alpha(b+f) \\ \alpha(c+g) & \alpha(d+h) \end{pmatrix}.
\]
Since scalar multiplication distributes over addition for real numbers, this equals
\[
\begin{pmatrix} \alpha a+\alpha e & \alpha b+\alpha f \\ \alpha c+\alpha g & \alpha d+\alpha h \end{pmatrix}
= \alpha\bA + \alpha\bB.
\]
}

\addExcr{%
Refer to Def~\ref{def:scalar-mult} and Thm~\ref{thm:Rmn-vecspace}.
\begin{enumerate}
    \item Compute $3\bA - 2\bB$ for
    $\bA = \begin{pmatrix}1 & 0 \\ -1 & 2\end{pmatrix}$ and
    $\bB = \begin{pmatrix}2 & 1 \\ 0 & -1\end{pmatrix}$.
    \item The standard basis for $\fR^{2 \times 2}$ consists of four matrices $\bE^{(ij)}$.
    Write them out explicitly and verify that $\bA = \sum_{i,j} a_{ij}\bE^{(ij)}$ holds for
    your $\bA$ above.
\end{enumerate}
}

\subsubsection{Dot Product}

Before defining the general matrix product, we introduce the dot product of two vectors.
The matrix product will then be understood entry-wise as a collection of dot products.

\addDef{Dot Product}{%
\index{dot|product}%
For vectors $\ba, \bb \in \fR^{n}$, the \textbf{dot product} (or inner product) is defined as
\begin{equation}
    \ba \cdot \bb = \ba^T \bb = \printrowvec{a} \printcolvec{b} = \sum_{i=1}^{n} a_{i} b_{i}.
\end{equation}
}
\label{def:dot-product}

\addExcr{%
Refer to Def~\ref{def:dot-product}.
\begin{enumerate}
    \item Compute $\ba \cdot \bb$ for $\ba = (1, 2, 3)^T$ and $\bb = (-1, 0, 2)^T$ by hand.
    \item In NumPy, evaluate this using \texttt{np.dot(a, b)} and also with \texttt{a @ b}.
    Confirm both give the same result.
    \item Show that $\ba \cdot \ba \ge 0$ with equality iff $\ba = \bzero$.  What geometric
    quantity does $\sqrt{\ba \cdot \ba}$ represent?
\end{enumerate}
}

\addDef{Matrix Product}{%
Let $\bA \in \mathbb{R}^{m \times n}$ and $\bB \in \mathbb{R}^{n \times p}$.  Their product
$\bA\bB$ is an $m \times p$ matrix defined by
\[
(\bA\bB)_{ij} = \sum_{k=1}^{n} a_{ik} b_{kj},
\]
or equivalently, if $\ba_i$ denotes the $i$-th row of $\bA$ and $\bb_j$ the $j$-th column
of $\bB$:
\[
\bA \bB =
\begin{pmatrix}
\ba_1 \cdot \bb_1 & \cdots & \ba_1 \cdot \bb_p \\
\vdots & \ddots & \vdots \\
\ba_m \cdot \bb_1 & \cdots & \ba_m \cdot \bb_p
\end{pmatrix}.
\]
}
\label{def:matrix-product}

\addNote{%
Matrix multiplication is defined only when the number of columns of $\bA$ equals the number
of rows of $\bB$.  It is associative and distributive over addition, but in general not
commutative: $\bA\bB \neq \bB\bA$ for most matrices.
}

\addLem{Computational Cost of Matrix Multiplication}{%
Computing $\bA\bB$ where $\bA \in \fR^{m \times n}$ and $\bB \in \fR^{n \times p}$ requires
exactly $mnp$ multiplications and $m(n-1)p$ additions using the standard algorithm, giving
$O(mnp)$ floating-point operations (flops).  For two square $n \times n$ matrices this
reduces to $O(n^3)$ flops.
}
\label{lem:matmul-cost}

\addNote{%
The $O(n^3)$ cost of naive matrix multiplication is a central bottleneck in scientific
computing.  Strassen's algorithm reduces this to $O(n^{\log_2 7}) \approx O(n^{2.807})$,
but is rarely used in practice due to numerical stability concerns and cache-unfriendly
memory access.  Modern BLAS/LAPACK implementations approach peak hardware throughput via
blocking strategies, but the asymptotic cost remains $O(n^3)$.  For many applications,
avoiding explicit matrix-matrix products entirely (e.g., exploiting sparsity or factored
structure) is more impactful than algorithmic improvements.
}

\textBox{%
\textbf{Compatibility for Matrix Multiplication}\\[2mm]
Let $\bA$ be a $3 \times 4$ matrix and $\bB$ a $4 \times 2$ matrix.  Since the number of
columns of $\bA$ (4) equals the number of rows of $\bB$ (4), the product $\bA\bB$ is
defined and results in a $3 \times 2$ matrix.  Conversely, if $\bB$ were a $5 \times 2$
matrix, the product would be undefined.
}

\textBox{%
\textbf{Distributive Property of Matrix Multiplication}\\[2mm]
Let
\[
\bA = \begin{pmatrix} a_{11} & a_{12} \\ a_{21} & a_{22} \end{pmatrix}, \quad
\bB = \begin{pmatrix} b_{11} & b_{12} \\ b_{21} & b_{22} \end{pmatrix}, \quad
\mathbf{C} = \begin{pmatrix} c_{11} & c_{12} \\ c_{21} & c_{22} \end{pmatrix}.
\]
Then,
\[
\bA(\bB + \mathbf{C}) = \bA \begin{pmatrix} b_{11}+c_{11} & b_{12}+c_{12} \\ b_{21}+c_{21} & b_{22}+c_{22} \end{pmatrix},
\]
and by computing the entries using Def~\ref{def:matrix-product}, one can verify that
\[
\bA(\bB + \mathbf{C}) = \bA\bB + \bA\mathbf{C}.
\]
}

\textBox{%
\textbf{Non-Commutativity of Matrix Multiplication}\\[2mm]
Consider
\[
\bA = \begin{pmatrix} 1 & 0 \\ 0 & 2 \end{pmatrix} \quad \text{and} \quad
\bB = \begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}.
\]
Then,
\[
\bA\bB = \begin{pmatrix} 0 & 1 \\ 2 & 0 \end{pmatrix}
\quad \text{and} \quad
\bB\bA = \begin{pmatrix} 0 & 2 \\ 1 & 0 \end{pmatrix}.
\]
Since $\bA\bB \neq \bB\bA$, matrix multiplication is not commutative.
}

\addExcr{%
Refer to Def~\ref{def:matrix-product} and Lem~\ref{lem:matmul-cost}.
Let $\bA = \begin{pmatrix} 2 & 1 \\ 0 & 3 \end{pmatrix}$ and
$\bB = \begin{pmatrix} 1 & -1 \\ 2 & 4 \end{pmatrix}$.
\begin{enumerate}
    \item Compute $\bA\bB$ and $\bB\bA$ by hand.  Confirm $\bA\bB \neq \bB\bA$.
    \item Find a nonzero matrix $\mathbf{C} \neq \bI$ that commutes with $\bA$
    (i.e., $\bA\mathbf{C} = \mathbf{C}\bA$).  \textit{Hint: try $\mathbf{C} = \alpha\bA + \beta\bI$.}
    \item Verify associativity: compute $(\bA\bB)\bc$ and $\bA(\bB\bc)$ for $\bc = (1,0)^T$
    and confirm they are equal.
    \item Using Lem~\ref{lem:matmul-cost}, compare the flop count of $(\bA\bB)\bc$ versus
    $\bA(\bB\bc)$ for general $\bA, \bB \in \fR^{n \times n}$ and $\bc \in \fR^n$.  Which
    order is cheaper and by what factor?
\end{enumerate}
}

Many properties of scalar arithmetic extend naturally to vectors and matrices.  In fact,
matrices can be viewed as elements of a higher-dimensional vector space (see
Thm~\ref{thm:Rmn-vecspace}), and matrix multiplication can be interpreted as the
composition of linear transformations.  While the notations for vectors and matrices are
similar, their roles differ:
\begin{itemize}
    \item Vectors typically represent points or directions in space.
    \item Matrices often represent linear maps or transformations between vector spaces.
\end{itemize}
This conceptual distinction becomes crucial when discussing eigenvalues and linear
transformations.

\addCor{Parenthesization Reduces Cost from $O(n^3)$ to $O(n^2)$}{%
For $\bA, \bB \in \fR^{n \times n}$ and $\bx \in \fR^n$, computing $\bA(\bB\bx)$ via two
matrix-vector products costs $O(n^2)$ flops, whereas forming $(\bA\bB)\bx$ via a
matrix-matrix product costs $O(n^3)$ flops.
}
\label{cor:parenthesization}

\addExcr{%
In this exercise, we will prove Cor~\ref{cor:parenthesization}.
Refer to Lem~\ref{lem:matmul-cost}.

Suppose $\bA, \bB \in \fR^{n \times n}$ and $\bx \in \fR^n$.
\begin{enumerate}
    \item \textit{Approach 1:} Compute $\bM = \bA\bB$ first (cost?), then $\bM\bx$ (cost?).
    Write the total flop count.
    \item \textit{Approach 2:} Compute $\by = \bB\bx$ first (cost?), then $\bA\by$ (cost?).
    Write the total flop count.
    \item Conclude the asymptotic improvement factor as $n \to \infty$.  This proves
    Cor~\ref{cor:parenthesization}.
    \item Write NumPy code for both approaches using \texttt{@}.  Does NumPy automatically
    choose the optimal evaluation order for a chain like \texttt{A @ B @ x}?
\end{enumerate}
}


\subsubsection{Matrix Inverse}

We now turn to a fundamental question: when can we ``divide'' by a matrix?  The matrix
inverse plays the role of reciprocal in scalar arithmetic, but it exists only under a
specific condition, and computing it explicitly is almost always the wrong approach in
practice.

\addDef{Matrix Inverse}{%
For a square matrix $\bA \in \fR^{n \times n}$, the \textbf{inverse} of $\bA$, denoted
$\bA^{-1}$, is the unique matrix satisfying
\[
\bA \bA^{-1} = \bA^{-1} \bA = \mathbf{I},
\]
where $\mathbf{I}$ is the $n \times n$ identity matrix.  A matrix that has an inverse is
called \emph{invertible} or \emph{nonsingular}.
}
\label{def:matrix-inverse}

\addNote{%
A square matrix $\bA$ is invertible if and only if $\det(\bA) \neq 0$, which also implies
that $\bA$ has full rank.
}

\textBox{%
\textbf{Uniqueness of the Inverse}\\[2mm]
Suppose $\bA$ is invertible and both $\bB$ and $\mathbf{C}$ satisfy the conditions of
Def~\ref{def:matrix-inverse}.  Then,
\[
\bB = \bB\mathbf{I} = \bB(\bA\mathbf{C}) = (\bB\bA)\mathbf{C} = \mathbf{I}\mathbf{C} = \mathbf{C}.
\]
Thus the inverse of $\bA$ is unique.
}

\addThm{Properties of the Matrix Inverse}{%
Let $\bA, \bB \in \fR^{n \times n}$ be invertible and $\alpha \in \fR \setminus \{0\}$.
Then:
\begin{enumerate}
    \item $(\bA^{-1})^{-1} = \bA$
    \item $(\bA\bB)^{-1} = \bB^{-1}\bA^{-1}$ \quad \textbf{(Reversal Law)}
    \item $(\alpha\bA)^{-1} = \tfrac{1}{\alpha}\bA^{-1}$
    \item $(\bA^T)^{-1} = (\bA^{-1})^T$
\end{enumerate}
}
\label{thm:inverse-props}

\addPrf{%
\textbf{(1):} $\bA^{-1}\bA = \bI$ and $\bA\bA^{-1} = \bI$ show $\bA$ is both a left and
right inverse of $\bA^{-1}$, so $(\bA^{-1})^{-1} = \bA$.

\textbf{(2):} Check the right-inverse condition:
\[
    (\bA\bB)(\bB^{-1}\bA^{-1}) = \bA(\bB\bB^{-1})\bA^{-1} = \bA\bI\bA^{-1}
    = \bA\bA^{-1} = \bI.
\]
Left-inverse: $(\bB^{-1}\bA^{-1})(\bA\bB) = \bB^{-1}(\bA^{-1}\bA)\bB = \bI$.
By uniqueness of the inverse, $(\bA\bB)^{-1} = \bB^{-1}\bA^{-1}$.

\textbf{(3):} $(\alpha\bA)\!\left(\tfrac{1}{\alpha}\bA^{-1}\right) =
\tfrac{\alpha}{\alpha}\bA\bA^{-1} = \bI$.

\textbf{(4):} Transpose both sides of $\bA\bA^{-1} = \bI$ using the Reversal Law
(see the transpose analogue): $(\bA^{-1})^T\bA^T = \bI$, so
$(\bA^T)^{-1} = (\bA^{-1})^T$.
}

\addNote{%
Property (2) is critical in scientific computing.  To solve $(\bA\bB)\bx = \bc$, we should
\emph{not} compute $(\bA\bB)^{-1}$ explicitly (large constant factor, numerical error
amplification).  Instead, use the Reversal Law: first solve $\bA\by = \bc$ for $\by$ via
LU factorization, then solve $\bB\bx = \by$.  The function \texttt{np.linalg.solve} does
exactly this.
}

\addExcr{%
Refer to Def~\ref{def:matrix-inverse} and Thm~\ref{thm:inverse-props}.
\begin{enumerate}
    \item For $\bA = \begin{pmatrix} 2 & 1 \\ 5 & 3 \end{pmatrix}$, compute $\bA^{-1}$
    by hand using the $2 \times 2$ formula
    $\bA^{-1} = \frac{1}{\det\bA}\begin{pmatrix} d & -b \\ -c & a \end{pmatrix}$.
    Verify $\bA\bA^{-1} = \bI$.
    \item Given $\bB = \begin{pmatrix} 1 & 0 \\ -1 & 1 \end{pmatrix}$, use the Reversal Law
    (Thm~\ref{thm:inverse-props}(2)) to compute $(\bA\bB)^{-1}$ without inverting a new
    $2 \times 2$ system.
    \item Explain why \texttt{np.linalg.solve(A, b)} is preferred over
    \texttt{np.linalg.inv(A) @ b} for solving $\bA\bx = \bb$.  Address both numerical
    accuracy and computational cost.
\end{enumerate}
}

\addDef{Involutory Matrix}{%
A square matrix $\bA \in \fR^{n \times n}$ is called \textbf{involutory} if $\bA^2 = \bI$.
}
\label{def:involutory}

\addExcr{%
In this exercise, we will prove that a matrix is involutory (Def~\ref{def:involutory}) if
and only if it is its own inverse.
Refer to Def~\ref{def:matrix-inverse} and Thm~\ref{thm:inverse-props}.
\begin{enumerate}
    \item Suppose $\bA^2 = \bI$.  Multiply both sides of $\bA\bA = \bI$ on the right by
    $\bA^{-1}$ to deduce $\bA^{-1} = \bA$.
    \item Conversely, suppose $\bA^{-1} = \bA$.  Show $\bA^2 = \bI$ directly from
    Def~\ref{def:matrix-inverse}.
    \item Give a concrete $2 \times 2$ example of an involutory matrix other than $\pm\bI$.
    Verify it satisfies $\bA^2 = \bI$.
\end{enumerate}
}

\addExcr{%
\textbf{(Identifying Invertibility)} For each matrix below, determine without computing
the inverse whether it is invertible.  Give a one-sentence justification.
\[
  (a)\ \begin{pmatrix} 1 & 2 \\ 2 & 4 \end{pmatrix} \qquad
  (b)\ \begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix} \qquad
  (c)\ \begin{pmatrix} 3 & 0 & 0 \\ 0 & 5 & 0 \\ 0 & 0 & -1 \end{pmatrix} \qquad
  (d)\ \begin{pmatrix} 1 & 1 & 0 \\ 0 & 1 & 1 \\ 1 & 2 & 1 \end{pmatrix}
\]
}

\end{document}
