\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Chebyshev Polynomials and Polynomial Approximation}
\label{sub:chebyshev}

\textBox{%
In the least squares exercises above, you built Vandermonde matrices (Def~\ref{def:vandermonde})
to fit polynomials to data. That worked fine for low degrees, but something goes wrong
as the degree grows: the matrix becomes catastrophically ill-conditioned, and even
a carefully placed set of interpolation points leads to a polynomial that oscillates
wildly. Both failures have the same root cause: the monomial basis
$\{1, x, x^2, \ldots, x^n\}$ is a poor basis for polynomial approximation.
Chebyshev polynomials fix both problems simultaneously. They are orthogonal, bounded,
and satisfy a minimax optimality property that makes them the best possible basis for
approximating smooth functions on a bounded interval. This subsection demonstrates
each failure mode and shows precisely how Chebyshev polynomials resolve it.
}

\subsubsection{Why the Monomial Basis Fails}

\textBox{%
There are two distinct ways the monomial basis breaks down. The first is numerical:
the Vandermonde matrix built from $\{1, x, x^2, \ldots, x^n\}$ has a condition number
that grows exponentially with degree, regardless of the node placement. The second is
analytical: even if the linear system could be solved exactly, interpolating a smooth
function at equally spaced points leads to a polynomial that diverges near the
endpoints as the degree grows. Understanding both failures motivates every design
choice in the Chebyshev approach.
}

\textBox{%
Recall from Def~\ref{def:vandermonde} that the Vandermonde matrix $\bV$ encodes a
polynomial fitting problem in the monomial basis: given nodes $x_0, \ldots, x_n$ and
values $f(x_0), \ldots, f(x_n)$, the coefficients of the interpolating polynomial
satisfy $\bV\bc = \mathbf{f}$. The matrix is invertible whenever the nodes are distinct
(since its determinant equals $\prod_{i>j}(x_i - x_j)$). The question we focus on
here is not whether a solution exists, but how reliably it can be computed.
}

\addExcr{%
Refer to Def~\ref{def:vandermonde}.
\begin{enumerate}
  \item Build the square Vandermonde matrix for $n+1$ equally spaced nodes on $[-1,1]$
  in NumPy for $n = 5, 10, 15, 20$. Compute $\kappa_2(\bV)$ for each. Plot $\kappa_2(\bV)$
  vs.\ $n$ on a log scale. How fast does the condition number grow?
  \item For $n = 10$, solve $\bV\bc = \mathbf{f}$ with $f(x_i) = \cos(\pi x_i)$
  using \texttt{np.linalg.solve}, then using \texttt{np.linalg.lstsq}. Do you get
  the same coefficients? Compute the residual $\|\bV\bc - \mathbf{f}\|_2$ for both.
  \item If $\mathbf{f}$ has noise at level $10^{-12}$ and $\kappa(\bV) = 10^{10}$,
  what is the worst-case relative error in $\bc$?
\end{enumerate}
}

\addDef{Runge's Phenomenon}{%
\emph{Runge's phenomenon} is the divergence of polynomial interpolants at equally
spaced nodes as the degree grows. The prototypical example is Runge's function
$f(x) = \frac{1}{1 + 25x^2}$ on $[-1, 1]$: although $f$ is infinitely differentiable,
the degree-$n$ interpolant at $n+1$ equally spaced nodes satisfies
\[
  \max_{x \in [-1,1]}|f(x) - p_n(x)| \to \infty \quad \text{as } n \to \infty.
\]
The oscillations are largest near the endpoints $\pm 1$.
}
\label{def:runge}

\addNote{%
The interpolation error at a point $x$ is bounded by
\[
  |f(x) - p_n(x)| \leq \frac{\|f^{(n+1)}\|_\infty}{(n+1)!}\,|\omega_n(x)|,
  \qquad \omega_n(x) = \prod_{k=0}^{n}(x - x_k).
\]
For equally spaced nodes, $\|\omega_n\|_\infty$ grows like $(n/2e)^n$ near the endpoints
(even though $(n+1)!$ also grows, the numerator wins). For Runge's function, the
derivatives $\|f^{(n+1)}\|_\infty$ also blow up, compounding the problem.
}

\addExcr{%
\textbf{(Runge's phenomenon.)}
Refer to Def~\ref{def:runge}.
\begin{enumerate}
  \item In NumPy, interpolate $f(x) = 1/(1+25x^2)$ at $n+1$ equally spaced nodes
  on $[-1,1]$ for $n = 5, 10, 15, 20$ using \texttt{np.polyfit} or by solving the
  Vandermonde system. For each $n$, plot both $f$ and $p_n$ on the same axes.
  Describe what happens near $x = \pm 1$ as $n$ grows.
  \item Compute $\max_{x}|f(x) - p_n(x)|$ (use 500 evaluation points) for
  $n = 5, 10, 15, 20$. Plot the maximum error vs.\ $n$. Does it decrease?
  \item (\textit{The node polynomial.})
  For $n = 10$ equally spaced nodes on $[-1,1]$, plot $|\omega_{10}(x)|$ vs.\ $x$.
  Where is $|\omega_{10}|$ largest? How does this explain the shape of the
  interpolation error in part 1?
\end{enumerate}
}

\subsubsection{Chebyshev Polynomials}

\textBox{%
The failures of the monomial basis come down to two causes: the monomials $x^k$ are
nearly linearly dependent on $[-1,1]$ for large $k$ (leading to ill-conditioning), and
equispaced nodes concentrate the node polynomial near the boundary (leading to Runge's
phenomenon). Chebyshev polynomials are the natural remedy: they are bounded by 1 on
$[-1,1]$, form an orthogonal basis, and their zeros (the Chebyshev nodes) distribute
mass towards the endpoints in exactly the way needed to suppress boundary oscillations.
}

\addDef{Chebyshev Polynomials}{%
The \emph{Chebyshev polynomials of the first kind} $T_k : [-1,1] \to \fR$ are defined by
\[
  T_k(x) = \cos(k\arccos x), \quad k = 0, 1, 2, \ldots
\]
They satisfy the three-term recurrence
\[
  T_0(x) = 1, \quad T_1(x) = x, \quad T_{k+1}(x) = 2x\,T_k(x) - T_{k-1}(x),
\]
and the first few are $T_0 = 1$, $T_1 = x$, $T_2 = 2x^2 - 1$,
$T_3 = 4x^3 - 3x$, $T_4 = 8x^4 - 8x^2 + 1$.
}
\label{def:chebyshev-poly}

\addThm{Key Properties of Chebyshev Polynomials}{%
For all $k \geq 0$ and $x \in [-1,1]$:
\begin{enumerate}
  \item \textbf{Boundedness}: $|T_k(x)| \leq 1$, with $T_k(\cos\theta) = \cos(k\theta)$.
  \item \textbf{Equioscillation}: $T_k$ attains the values $\pm 1$ exactly $k+1$ times
  on $[-1,1]$ (at $x_j = \cos(j\pi/k)$ for $j = 0, \ldots, k$).
  \item \textbf{Zeros}: $T_k$ has $k$ simple roots at
  $\tilde{x}_j = \cos\!\left(\tfrac{(2j-1)\pi}{2k}\right)$ for $j = 1, \ldots, k$.
  \item \textbf{Leading coefficient}: $T_k$ is a polynomial of degree $k$ with
  leading coefficient $2^{k-1}$ (for $k \geq 1$).
\end{enumerate}
}
\label{thm:chebyshev-props}

\addPrf{%
All four properties follow directly from the definition $T_k(x) = \cos(k\arccos x)$.

\medskip
\textbf{(1) and (2):} Setting $x = \cos\theta$ gives $T_k(\cos\theta) = \cos(k\theta)$.
Since $|\cos(k\theta)| \leq 1$ for all $\theta$, we have $|T_k(x)| \leq 1$.
The cosine equals $\pm 1$ at $\theta = j\pi/k$, giving the equioscillation points.

\textbf{(3):} The roots are where $\cos(k\theta) = 0$, i.e., $k\theta = (2j-1)\pi/2$,
giving $\theta = (2j-1)\pi/(2k)$ and $x = \cos((2j-1)\pi/(2k))$.

\textbf{(4):} The recurrence $T_{k+1} = 2x\,T_k - T_{k-1}$ starts from $T_1(x) = x$
(leading coefficient $1 = 2^0$) and $T_2(x) = 2x^2 - 1$ (leading coefficient $2 = 2^1$).
By induction, the recurrence doubles the leading coefficient at each step,
giving $2^{k-1}$ for $T_k$.
}

\addExcr{%
Refer to Def~\ref{def:chebyshev-poly} and Thm~\ref{thm:chebyshev-props}.
\begin{enumerate}
  \item Using the recurrence, compute $T_5(x)$ and $T_6(x)$ by hand. Verify your
  results with \texttt{numpy.polynomial.chebyshev.chebval}.
  \item Plot $T_0, T_1, \ldots, T_5$ on $[-1, 1]$ on the same axes. Confirm that
  each $T_k$ stays within $[-1,1]$ and oscillates exactly $k$ times.
  \item Verify numerically that the roots of $T_5$ are
  $x_j = \cos((2j-1)\pi/10)$ for $j = 1, \ldots, 5$.
  \item Confirm the leading coefficient of $T_k$ is $2^{k-1}$ for $k = 2, 3, 4$ by
  expanding the recurrence and reading off the coefficient of $x^k$.
\end{enumerate}
}

\subsubsection{Orthogonality and Chebyshev Nodes}

\textBox{%
The cosine definition immediately reveals an orthogonality structure. Under the change
of variables $x = \cos\theta$, the Chebyshev polynomials become ordinary cosines, and
their orthogonality follows from the orthogonality of cosines. In the $x$ variable, the
weight function $w(x) = 1/\sqrt{1-x^2}$ appears, which concentrates mass near the
endpoints $\pm 1$. The zeros of $T_n$ (the Chebyshev nodes) are exactly the points
where this endpoint clustering is captured in a finite node set. Using them for
interpolation produces a Vandermonde-type system that is dramatically better conditioned
than the equispaced case.
}

\addThm{Chebyshev Orthogonality}{%
The Chebyshev polynomials are orthogonal on $[-1,1]$ with weight $w(x) = 1/\sqrt{1-x^2}$:
\[
  \int_{-1}^{1} T_j(x)\,T_k(x)\,\frac{dx}{\sqrt{1-x^2}} =
  \begin{cases}
    0     & j \neq k, \\
    \pi/2 & j = k \geq 1, \\
    \pi   & j = k = 0.
  \end{cases}
\]
}
\label{thm:chebyshev-orthogonal}

\addPrf{%
Substitute $x = \cos\theta$, $dx = -\sin\theta\,d\theta$, $\sqrt{1-x^2} = \sin\theta$:
\[
  \int_{-1}^{1} T_j(x)\,T_k(x)\,\frac{dx}{\sqrt{1-x^2}}
  = \int_\pi^0 \cos(j\theta)\cos(k\theta)\,\frac{-\sin\theta\,d\theta}{\sin\theta}
  = \int_0^\pi \cos(j\theta)\cos(k\theta)\,d\theta.
\]
The last integral is the standard cosine orthogonality relation, which equals $0$ for
$j \neq k$, $\pi/2$ for $j = k \geq 1$, and $\pi$ for $j = k = 0$.
}

\addDef{Chebyshev Nodes}{%
The $n$ \emph{Chebyshev nodes of the first kind} on $[-1,1]$ are the zeros of $T_n$:
\[
  x_k = \cos\!\left(\frac{(2k-1)\pi}{2n}\right), \quad k = 1, \ldots, n.
\]
They cluster near the endpoints $\pm 1$ (the density of nodes near $x$ is proportional
to $1/\sqrt{1-x^2}$, exactly matching the weight in the orthogonality relation).
For nodes on a general interval $[a,b]$, shift and scale:
$x_k^{[a,b]} = \frac{a+b}{2} + \frac{b-a}{2}\,x_k$.
}
\label{def:chebyshev-nodes}

\addThm{The Node Polynomial at Chebyshev Nodes}{%
For the $n$ Chebyshev nodes $x_1, \ldots, x_n$, the node polynomial satisfies
\[
  \prod_{k=1}^{n}(x - x_k) = \frac{T_n(x)}{2^{n-1}}.
\]
Its supremum norm on $[-1,1]$ is $1/2^{n-1}$.
}
\label{thm:chebyshev-node-poly}

\addPrf{%
The zeros of $T_n$ are exactly the Chebyshev nodes $x_k$. The leading coefficient of
$T_n$ is $2^{n-1}$ (Thm~\ref{thm:chebyshev-props}(4)), so the monic polynomial with
the same roots is $T_n(x)/2^{n-1}$. Since $|T_n(x)| \leq 1$ on $[-1,1]$, the supremum
norm of the node polynomial is at most $1/2^{n-1}$, and it is attained at the
equioscillation points of $T_n$.
}

\addExcr{%
Refer to Defs~\ref{def:chebyshev-nodes} and~\ref{thm:chebyshev-node-poly}.
\begin{enumerate}
  \item Compute the 10 Chebyshev nodes on $[-1,1]$ and plot them alongside
  10 equally spaced nodes. Describe the qualitative difference in their distribution.
  \item For $n = 10$, plot $|\omega_n(x)|$ for both equispaced nodes and Chebyshev
  nodes on $[-1, 1]$ using 500 evaluation points.
  Compare $\max_x |\omega_n(x)|$ for both. How much smaller is the Chebyshev version?
  \item Now repeat the Runge function experiment (Exercise from
  Def~\ref{def:runge}) but using Chebyshev nodes instead of equispaced nodes.
  Plot the max interpolation error vs.\ $n$ for both node sets on the same axes.
  Does the error decrease with Chebyshev nodes?
\end{enumerate}
}

\addExcr{%
\textbf{(Condition numbers: monomial vs.\ Chebyshev basis.)}
Let $\bV_{\mathrm{mono}}$ be the Vandermonde matrix with $n+1$ Chebyshev nodes and
the monomial basis $V_{ij} = x_i^j$. Let $\bV_{\mathrm{cheb}}$ be the same nodes but
with the Chebyshev basis $V_{ij} = T_j(x_i)$.
\begin{enumerate}
  \item Build both matrices in NumPy for $n = 5, 10, 15, 20$. Compute $\kappa_2$ for
  each. Plot both condition numbers vs.\ $n$ on the same log-scale axes.
  \item Explain why $\bV_{\mathrm{cheb}}$ has a much better condition number, using
  the orthogonality of $\{T_k\}$ (Thm~\ref{thm:chebyshev-orthogonal}).
  \item Using both matrices, solve the interpolation problem for
  $f(x) = 1/(1+25x^2)$ at $n = 20$ Chebyshev nodes. Do both systems give the same
  polynomial? Which system is safer to solve numerically?
\end{enumerate}
}

\subsubsection{The Minimax Property}

\textBox{%
The node polynomial theorem says Chebyshev nodes minimize $\|\omega_n\|_\infty$ over
all choices of $n$ nodes. But Chebyshev polynomials satisfy something even stronger:
$T_n/2^{n-1}$ is the unique monic polynomial of degree $n$ that has the smallest
possible sup-norm on $[-1,1]$, among all monic polynomials of degree $n$. This
minimax property is the algebraic reason Chebyshev nodes and basis are optimal, and
it connects directly to the equioscillation characterization of best polynomial
approximations.
}

\addThm{Chebyshev Minimax Property}{%
Among all monic polynomials of degree $n$, the scaled Chebyshev polynomial
$\tilde{T}_n(x) = T_n(x)/2^{n-1}$ has the smallest sup-norm on $[-1,1]$:
\[
  \|\tilde{T}_n\|_\infty = \frac{1}{2^{n-1}} \leq \|q\|_\infty
\]
for every monic polynomial $q$ of degree $n$.
}
\label{thm:chebyshev-minimax}

\addPrf{%
We argue by contradiction.

\medskip
\textbf{Step 1: Suppose a better monic polynomial exists.}

Assume $q$ is monic of degree $n$ and $\|q\|_\infty < 1/2^{n-1}$.

\textbf{Step 2: Form the difference.}

Let $r = \tilde{T}_n - q$. Since both are monic of degree $n$, their leading coefficients
cancel and $r$ has degree at most $n-1$.

\textbf{Step 3: Count sign changes.}

The polynomial $\tilde{T}_n$ equioscillates between $+1/2^{n-1}$ and $-1/2^{n-1}$ at
$n+1$ points $x_0 > x_1 > \cdots > x_n$ (the extrema of $T_n$). At each $x_j$,
$|\tilde{T}_n(x_j)| = 1/2^{n-1}$ and $|q(x_j)| < 1/2^{n-1}$, so $r(x_j) =
\tilde{T}_n(x_j) - q(x_j)$ has the same sign as $\tilde{T}_n(x_j)$.
Since $\tilde{T}_n$ alternates sign at consecutive extrema, so does $r$.

\textbf{Step 4: Derive a contradiction.}

$r$ changes sign at least $n$ times, so $r$ has at least $n$ roots.
But $\deg(r) \leq n-1$, so $r$ can have at most $n-1$ roots unless $r \equiv 0$.
This is a contradiction, so no such $q$ exists.
}

\addExcr{%
Refer to Thm~\ref{thm:chebyshev-minimax}.
\begin{enumerate}
  \item For $n = 1, 2, 3$, write out $\tilde{T}_n(x) = T_n(x)/2^{n-1}$ explicitly
  and verify $\|\tilde{T}_n\|_\infty = 1/2^{n-1}$ by finding its maximum on $[-1,1]$.
  \item Show that among all polynomials $p$ of degree $\leq n-1$, the best approximation
  to $x^n$ in the sup-norm on $[-1,1]$ is $x^n - \tilde{T}_n(x)$.
  (\textit{Hint: write $x^n = \tilde{T}_n(x) + (x^n - \tilde{T}_n(x))$ and apply the
  minimax theorem.})
  \item In NumPy, generate 10 random monic polynomials of degree 5 with coefficients
  drawn from $\mathcal{N}(0,1)$ (for the lower-degree terms). Compare their
  $\|\cdot\|_\infty$ norms on $[-1,1]$ to $1/2^4 = 1/16$. How many beat $1/16$?
\end{enumerate}
}

\subsubsection{Chebyshev Series and Approximation}

\textBox{%
Just as the Fourier series expands a periodic function in the basis $\{e^{ikx}\}$,
the Chebyshev series expands a function on $[-1,1]$ in the basis $\{T_k\}$.
Under the substitution $x = \cos\theta$, a function $f(x)$ on $[-1,1]$ becomes
$g(\theta) = f(\cos\theta)$ on $[0,\pi]$, and the Chebyshev series of $f$ is exactly
the cosine series of $g$. This is yet another instance of the Fourier change-of-basis
from Section~\ref{sub:fourier}, now applied to the interval $[-1,1]$ via the
substitution $x = \cos\theta$.
}

\addDef{Chebyshev Series}{%
For $f : [-1,1] \to \fR$, the \emph{Chebyshev series} is
\[
  f(x) = \sum_{k=0}^{\infty} {}' c_k T_k(x),
\]
where the prime on the sum means the $k=0$ term is halved, and the coefficients are
\[
  c_k = \frac{2}{\pi}\int_{-1}^{1} f(x)\,T_k(x)\,\frac{dx}{\sqrt{1-x^2}}.
\]
The truncation $p_n(x) = \sum_{k=0}^{n} {}' c_k T_k(x)$ is the degree-$n$ Chebyshev
approximation to $f$.
}
\label{def:chebyshev-series}

\addThm{Exponential Convergence for Analytic Functions}{%
If $f$ is analytic on $[-1,1]$ and can be analytically continued to a Bernstein
ellipse with parameter $\rho > 1$ (an ellipse with foci at $\pm 1$ and sum of
semi-axes equal to $\rho$), then the Chebyshev coefficients satisfy
\[
  |c_k| \leq M\,\rho^{-k}
\]
for some constant $M$. Consequently, the truncation error satisfies
$\|f - p_n\|_\infty = O(\rho^{-n})$ as $n \to \infty$.
}
\label{thm:chebyshev-convergence}

\addNote{%
The exponential convergence rate contrasts sharply with the polynomial convergence of
finite-difference approximations ($O(h^p)$ for order-$p$ methods). For smooth functions,
increasing the polynomial degree by 1 multiplies the error by $1/\rho < 1$. In practice,
just 20 to 30 Chebyshev terms often give errors near machine precision for analytic
functions. This is called \emph{spectral accuracy}.
}

\addExcr{%
Refer to Def~\ref{def:chebyshev-series} and Thm~\ref{thm:chebyshev-convergence}.
\begin{enumerate}
  \item Use \texttt{numpy.polynomial.chebyshev.chebfit(x, f, deg)} to compute the
  degree-$n$ Chebyshev approximation to $f(x) = e^x$ on $[-1,1]$ for
  $n = 2, 4, 8, 16$. Plot $\|f - p_n\|_\infty$ vs.\ $n$ on a log scale.
  Does the error decrease exponentially?
  \item Repeat for $f(x) = |x|$ (not analytic at $x=0$). Compare the convergence
  rate for $e^x$ and $|x|$. What does Thm~\ref{thm:chebyshev-convergence} predict
  for $|x|$, given that it cannot be analytically continued past any ellipse?
  \item (\textit{Comparison of all three methods.})
  For $f(x) = 1/(1+25x^2)$, compute the degree-$n$ approximation error
  $\|f - p_n\|_\infty$ for $n = 5, 10, 15, 20$ using three methods:
  (a) monomial basis with equispaced nodes (Vandermonde solve),
  (b) monomial basis with Chebyshev nodes,
  (c) Chebyshev basis with Chebyshev nodes (\texttt{chebfit}).
  Plot all three error curves on the same log-scale axes.
  Which methods converge? Which diverge? Explain each result using the theory above.
\end{enumerate}
}

\addExcr{%
\textbf{(Chebyshev differentiation.)}
Because the Chebyshev basis diagonalizes the right operations, differentiation in the
Chebyshev basis is a structured recurrence rather than a full matrix-vector product.
\begin{enumerate}
  \item Show that $T_k'(x) = k U_{k-1}(x)$ where $U_{k-1}(x) = \sin(k\arccos x)/\sin(\arccos x)$
  are the Chebyshev polynomials of the second kind. (\textit{Hint: differentiate
  $T_k(x) = \cos(k\arccos x)$ using the chain rule.})
  \item The derivative of $f(x) = \sum_k c_k T_k(x)$ has Chebyshev coefficients
  $c_k^{(1)}$ satisfying the recurrence $c_{k-1}^{(1)} = c_{k+1}^{(1)} + 2k c_k$
  (summing from $k = n$ down to $1$). Use this to differentiate $p(x) = T_3(x) + 2T_2(x)$
  and verify against the exact derivative.
  \item Implement spectral differentiation via \texttt{numpy.polynomial.chebyshev.chebder}
  and compare the accuracy of differentiating $f(x) = \sin(\pi x)$ using (a) second-order
  finite differences with $n$ equispaced points and (b) the Chebyshev spectral derivative
  with $n$ Chebyshev nodes. Plot the error vs.\ $n$ for both. At $n = 20$, which method
  is more accurate?
\end{enumerate}
}

\end{document}
