\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Computational Complexity}
\label{sub:complexity}

Every algorithm in scientific computing carries a cost: the number of arithmetic
operations it performs. Understanding that cost determines whether an algorithm is
practical at scale and lets us compare competing methods objectively. The flop counts
derived here will be referenced throughout the course whenever we compare algorithms.

\subsubsection{Big-O Notation}

\addDef{Big-O Notation}{%
We write $f(n) = O(g(n))$ (read ``$f$ is big-O of $g$'') if there exist constants
$C > 0$ and $n_0 \geq 1$ such that
\[
  |f(n)| \leq C\,|g(n)| \quad \text{for all } n \geq n_0.
\]
Informally, $f$ grows no faster than $g$ asymptotically. Related notations:
\begin{itemize}
  \item $f = \Omega(g)$: $f$ grows at least as fast as $g$ (i.e., $g = O(f)$).
  \item $f = \Theta(g)$: $f$ and $g$ grow at the same rate ($f = O(g)$ and $g = O(f)$).
  \item $f = o(g)$: $f$ grows strictly slower than $g$, i.e., $f(n)/g(n) \to 0$.
\end{itemize}
}
\label{def:big-o}

\addExmpl{%
Common growth rates, ordered from fastest to slowest for large $n$:
\[
  O(n!) \;\gg\; O(2^n) \;\gg\; O(n^3) \;\gg\; O(n^2) \;\gg\; O(n \log n) \;\gg\; O(n) \;\gg\; O(\log n) \;\gg\; O(1).
\]
For $n = 1000$: an $O(n^2)$ algorithm performs $\sim 10^6$ operations; an $O(n^3)$
algorithm performs $\sim 10^9$ --- three orders of magnitude slower. The distinction
between $O(n^2)$ and $O(n^3)$ is the difference between milliseconds and seconds.
}

\addThm{Properties of Big-O}{%
Let $f, g, h$ be positive functions of $n$. Then:
\begin{enumerate}
  \item (Transitivity) If $f = O(g)$ and $g = O(h)$, then $f = O(h)$.
  \item (Sum rule) $O(f) + O(g) = O(\max(f, g))$. In particular, $O(n^3) + O(n^2) = O(n^3)$.
  \item (Product rule) $O(f) \cdot O(g) = O(fg)$.
  \item (Constant factors) $O(cf) = O(f)$ for any constant $c > 0$.
\end{enumerate}
}
\label{thm:big-o-properties}

\addPrf{%
All four follow directly from the definition. For the sum rule: if $|f_1| \leq C_1 g$ and
$|f_2| \leq C_2 h$ for large $n$, then $|f_1 + f_2| \leq C_1 g + C_2 h \leq
(C_1 + C_2)\max(g, h)$. For the product rule: $|f_1 f_2| \leq C_1 C_2 \cdot gh$.
}

\addExcr{%
Refer to Def~\ref{def:big-o} and Thm~\ref{thm:big-o-properties}.
\begin{enumerate}
  \item Show that $3n^2 + 100n + 5 = O(n^2)$ by finding explicit constants $C$ and $n_0$.
  \item Is $n \log n = O(n^2)$? Is $n^2 = O(n \log n)$? Justify both.
  \item Rank the following by asymptotic cost for large $n$:
  (a)~$5n^3 + n$, \enspace (b)~$100n^2$, \enspace (c)~$n^2 \log n$, \enspace (d)~$2^{10} n^2$.
  \item An algorithm has cost $T(n) = 2T(n/2) + n$. Use the master theorem (or unroll
  the recurrence) to show $T(n) = O(n \log n)$.
\end{enumerate}
}

\subsubsection{Flop Counting}

\addDef{Floating-Point Operation (flop)}{%
A \emph{floating-point operation} (\emph{flop}) is a single arithmetic operation
(addition, subtraction, multiplication, or division) on floating-point numbers.
Flop counts give a hardware-independent measure of algorithmic work.
Modern CPUs execute billions of flops per second (GFlop/s); GPUs achieve TFlop/s.
}
\label{def:flop}

\addThm{Basic Operation Costs}{%
For vectors $\bx, \by \in \fR^n$ and matrices $\bA \in \fR^{m \times n}$,
$\bB \in \fR^{n \times p}$:
\begin{itemize}
  \item \textbf{Dot product} $\bx^T\by$: $n$ multiplications $+$ $(n-1)$ additions $= 2n - 1 \approx 2n$ flops.
  \item \textbf{Matrix-vector product} $\bA\bx$: $m$ dot products of length $n$ $= 2mn - m \approx 2mn$ flops.
  \item \textbf{Matrix-matrix product} $\bA\bB$: $p$ matrix-vector products $= 2mnp$ flops; for square $n \times n$ matrices, $2n^3$ flops.
  \item \textbf{Triangular solve} $\bL\bx = \bb$ (lower triangular): $n^2$ flops.
\end{itemize}
}
\label{thm:basic-flops}

\addPrf{%
Each row $i$ of $\bA\bx$ is the dot product $\ba_i^T\bx$, costing $2n - 1$ flops.
Summing over $m$ rows gives $m(2n-1) = 2mn - m$ flops. For $\bA\bB$, column $j$ of
the product is $\bA\bb_j$, a matrix-vector product costing $2mn - m$ flops; summing
over $p$ columns gives $p(2mn - m) \approx 2mnp$. For the triangular solve,
substituting entry $k$ of $\bx$ costs $2k - 1$ flops; summing $k = 1, \ldots, n$
gives $\sum_{k=1}^{n}(2k - 1) = n^2$.
}

\addNote{%
These baseline costs appear throughout the course. For reference:
LU factorization costs $\frac{2}{3}n^3$ flops (Section~\ref{sub:lu-factorization});
Cholesky costs $\frac{1}{3}n^3$ --- half the price for symmetric positive definite
systems; the QR factorization via Householder reflections costs $\frac{4}{3}n^3$ flops
(Lem~\ref{lem:qr-cost}).
}

\addExcr{%
Refer to Def~\ref{def:flop} and Thm~\ref{thm:basic-flops}.
\begin{enumerate}
  \item Verify the exact flop count for $\bA\bx$ ($\bA \in \fR^{m \times n}$) by
  counting each multiply-add in each row.
  \item A computer performs $10^9$ flops/second. Estimate wall-clock times for:
  \begin{itemize}
    \item $\bA\bx$ with $\bA \in \fR^{1000 \times 1000}$.
    \item $\bA\bB$ with $\bA, \bB \in \fR^{1000 \times 1000}$.
    \item $\bA\bB$ with $\bA, \bB \in \fR^{10000 \times 10000}$.
  \end{itemize}
  \item In Python, time \texttt{np.dot(A, B)} for $n \in \{100, 500, 1000, 2000\}$
  with random $n \times n$ matrices. Plot wall-clock time vs.\ $n$ on a log-log scale
  and estimate the empirical exponent. Does it match $O(n^3)$?
  \item Explain why the empirical exponent from the previous part may be less than 3.
  (\textit{Hint: BLAS libraries exploit cache hierarchy and vectorized instructions.})
\end{enumerate}
}

\end{document}
