\documentclass[../readings.tex]{subfiles}


\begin{document}
\subsection{Eigenvalues and Eigenvectors}
\label{subsec:eigenvalues}

Eigenvalues and eigenvectors reveal the fundamental character of a linear transformation:
the special directions along which the matrix acts by pure scaling. This section develops
the theory from the characteristic polynomial through the Spectral Theorem, then turns to
practical algorithms --- power iteration, inverse iteration, and the QR algorithm --- for
computing eigenvalues numerically.

\addDef{Eigenvalues and Eigenvectors}{%
For a matrix $\bA \in \fR^{n \times n}$, a scalar $\lambda \in \fC$ is called an
\emph{eigenvalue} of $\bA$ if there exists a nonzero vector $\bv \in \fC^n$ such that
\[
\bA\bv = \lambda\bv.
\]
The nonzero vector $\bv$ is called an \emph{eigenvector} of $\bA$ corresponding to
$\lambda$. The set of all eigenvalues of $\bA$ is called the \emph{spectrum} of $\bA$,
denoted $\sigma(\bA)$.
}
\label{def:eigenvalue}

\textBox{%
Intuitively, an eigenvector is a vector whose direction remains unchanged when transformed
by $\bA$ --- it may be stretched, compressed, or reversed, but it stays on the same line.
The eigenvalue tells us the factor by which the eigenvector is scaled. Eigenvalues and
eigenvectors reveal the fundamental character of a linear transformation, showing the
directions in which the transformation behaves most simply.
}

\addExcr{%
Refer to Def~\ref{def:eigenvalue}.
\begin{enumerate}
    \item Verify that $\bv = (1, 1)^T$ is an eigenvector of
    $\bA = \begin{pmatrix}3&1\\1&3\end{pmatrix}$. What is the corresponding eigenvalue?
    \item If $\bv$ is an eigenvector of $\bA$ with eigenvalue $\lambda$, show that $c\bv$
    is also an eigenvector for any nonzero scalar $c$.
    \item Can the zero vector be an eigenvector? Why is it explicitly excluded from the
    definition?
\end{enumerate}
}

\addDef{Characteristic Polynomial}{%
For a matrix $\bA \in \fR^{n \times n}$, the \emph{characteristic polynomial} is
\[
p_{\bA}(\lambda) = \det(\bA - \lambda \bI),
\]
where $\bI$ is the $n \times n$ identity matrix. The equation $\det(\bA - \lambda \bI) = 0$
is the \emph{characteristic equation} of $\bA$. The roots of the characteristic polynomial
are precisely the eigenvalues of $\bA$.
}
\label{def:char-poly}

\addExmpl{%
Consider the matrix
$\bA = \begin{pmatrix} 3 & 1 \\ 1 & 3 \end{pmatrix}$.

To find its eigenvalues, we compute the characteristic polynomial:
\begin{align}
p_{\bA}(\lambda)
&= \det\begin{pmatrix} 3-\lambda & 1 \\ 1 & 3-\lambda \end{pmatrix}
= (3-\lambda)^2 - 1
= \lambda^2 - 6\lambda + 8
= (\lambda - 4)(\lambda - 2).
\end{align}

Setting $p_{\bA}(\lambda) = 0$ gives eigenvalues $\lambda_1 = 4$ and $\lambda_2 = 2$.

For $\lambda_1 = 4$: solving $(\bA - 4\bI)\bv = \bzero$ gives $-v_1 + v_2 = 0$, so
$\bv_1 = \frac{1}{\sqrt{2}}(1, 1)^T$ (normalized).

For $\lambda_2 = 2$: solving $(\bA - 2\bI)\bv = \bzero$ gives $v_1 + v_2 = 0$, so
$\bv_2 = \frac{1}{\sqrt{2}}(1, -1)^T$ (normalized).
}

\addExcr{%
Refer to Def~\ref{def:char-poly}.
\begin{enumerate}
    \item Compute the characteristic polynomial and eigenvalues of
    $\bA = \begin{pmatrix}1 & 2 \\ 2 & 1\end{pmatrix}$.
    Verify using trace: $\lambda_1 + \lambda_2 = \text{tr}(\bA)$.
    \item For the $2 \times 2$ rotation matrix
    $\bB = \begin{pmatrix}0 & -1 \\ 1 & 0\end{pmatrix}$, compute the characteristic
    polynomial. Are the eigenvalues real? What does this say geometrically?
    \item In NumPy, use \texttt{np.linalg.eigvals(A)} to find the eigenvalues of each
    matrix above and verify they match your hand computations.
\end{enumerate}
}

\addProp{Eigenvalues of Triangular Matrices}{%
If $\bA$ is upper or lower triangular, its eigenvalues are precisely the diagonal entries
$a_{11}, a_{22}, \ldots, a_{nn}$.
}
\label{prop:triangular-eigenvalues}

\addPrf{%
The characteristic polynomial $\det(\bA - \lambda\bI)$ for a triangular matrix is the
product of its diagonal entries $(a_{ii} - \lambda)$, giving
$p_\bA(\lambda) = \prod_{i=1}^n (a_{ii} - \lambda)$. The roots are $\lambda_i = a_{ii}$.
}

\addProp{Similar Matrices Share Eigenvalues}{%
If $\bB = \bV^{-1}\bA\bV$ for some invertible $\bV$, then $\bA$ and $\bB$ have the same
characteristic polynomial and hence the same eigenvalues.
}
\label{prop:similar-eigenvalues}

\addPrf{%
$\det(\bB - \lambda\bI) = \det(\bV^{-1}\bA\bV - \lambda\bI)
= \det(\bV^{-1}(\bA - \lambda\bI)\bV)
= \det(\bV^{-1})\det(\bA - \lambda\bI)\det(\bV)
= \det(\bA - \lambda\bI)$.
}

\addThm{Key Eigenvalue Properties}{%
For $\bA \in \fR^{n \times n}$ with eigenvalues $\lambda_1, \ldots, \lambda_n$ (counting
algebraic multiplicity):
\begin{enumerate}
    \item $\displaystyle\text{tr}(\bA) = \sum_{i=1}^n \lambda_i$
    \item $\displaystyle\det(\bA) = \prod_{i=1}^n \lambda_i$
    \item $\bA$ is singular if and only if $0 \in \sigma(\bA)$.
    \item If $\bA$ is invertible, the eigenvalues of $\bA^{-1}$ are
    $1/\lambda_1, \ldots, 1/\lambda_n$.
\end{enumerate}
}
\label{thm:eigenvalue-properties}

\addPrf{%
(1) and (2): Both follow from comparing the characteristic polynomial
\[
    p_\bA(\lambda) = \det(\bA - \lambda\bI) = (-1)^n(\lambda - \lambda_1)\cdots(\lambda - \lambda_n)
\]
with the direct expansion of the determinant. Setting $\lambda = 0$ gives
$\det(\bA) = \prod_i \lambda_i$. The coefficient of $\lambda^{n-1}$ gives
$-\text{tr}(\bA) = -\sum_i \lambda_i$.

(3): Direct from (2): $\det(\bA) = 0 \iff \prod_i \lambda_i = 0 \iff$ some $\lambda_i = 0$.

(4): If $\bA\bv = \lambda\bv$ with $\lambda \neq 0$, multiply both sides by
$\frac{1}{\lambda}\bA^{-1}$ to get $\bA^{-1}\bv = \frac{1}{\lambda}\bv$.
}

\addLem{Eigenvectors for Distinct Eigenvalues are Linearly Independent}{%
Let $\lambda_1, \ldots, \lambda_k$ be distinct eigenvalues of $\bA$ with corresponding
eigenvectors $\bv_1, \ldots, \bv_k$. Then $\{\bv_1, \ldots, \bv_k\}$ is linearly
independent.
}
\label{lem:distinct-eigenvectors}

\addPrf{%
By induction on $k$. The case $k=1$ holds since eigenvectors are nonzero. Assume
$\{\bv_1, \ldots, \bv_{k-1}\}$ is independent. Suppose $\sum_{i=1}^k c_i\bv_i = \bzero$.
Apply $(\bA - \lambda_k\bI)$ to both sides:
\[
    \sum_{i=1}^{k-1} c_i(\lambda_i - \lambda_k)\bv_i = \bzero.
\]
By the inductive hypothesis, $c_i(\lambda_i - \lambda_k) = 0$ for each $i < k$. Since
$\lambda_i \neq \lambda_k$, each $c_i = 0$. Substituting back gives $c_k\bv_k = \bzero$,
so $c_k = 0$.
}

\addExcr{%
Refer to Props~\ref{prop:triangular-eigenvalues}--\ref{prop:similar-eigenvalues},
Thm~\ref{thm:eigenvalue-properties}, and Lem~\ref{lem:distinct-eigenvectors}.
\begin{enumerate}
    \item Read off the eigenvalues of $\bT = \begin{pmatrix}4&2&1\\0&-1&3\\0&0&2\end{pmatrix}$
    without any computation. Verify using \texttt{np.linalg.eigvals(T)}.
    \item For $\bA = \begin{pmatrix}5&2\\2&2\end{pmatrix}$, use properties (1) and (2) of
    Thm~\ref{thm:eigenvalue-properties} to check your eigenvalue computation: verify
    $\lambda_1 + \lambda_2 = \text{tr}(\bA)$ and $\lambda_1\lambda_2 = \det(\bA)$.
    \item Let $\bA$ have eigenvalues $2, 3, 5$. What are the eigenvalues of $\bA^{-1}$?
    Of $\bA^2$? Of $\bA - 2\bI$?
    \item If $\bA$ has three distinct eigenvalues, does it follow that its three
    eigenvectors form a basis for $\fR^3$? Cite the relevant lemma.
\end{enumerate}
}

\addThm{Spectral Theorem for Real Symmetric Matrices}{%
Let $\bA \in \fR^{n \times n}$ be symmetric. Then:
\begin{enumerate}
    \item All eigenvalues of $\bA$ are real.
    \item Eigenvectors corresponding to distinct eigenvalues are orthogonal.
    \item $\bA$ is orthogonally diagonalizable: $\bA = \bQ\bD\bQ^T$ where $\bQ$ is
    orthogonal and $\bD = \text{diag}(\lambda_1, \ldots, \lambda_n) \in \fR^{n\times n}$.
\end{enumerate}
}
\label{thm:spectral-theorem}

\addPrf{%
(1): Let $\lambda \in \fC$ be an eigenvalue with eigenvector $\bv \in \fC^n$,
$\bA\bv = \lambda\bv$. Then $\bar{\bv}^T\bA\bv = \lambda\,\bar{\bv}^T\bv$. But also
$\bar{\bv}^T\bA\bv = (\bA\bar{\bv})^T\bv = \overline{(\bA\bv)}^T\bv =
\bar{\lambda}\,\bar{\bv}^T\bv$. Since $\bar{\bv}^T\bv = \|\bv\|_2^2 > 0$, we conclude
$\lambda = \bar{\lambda}$, so $\lambda \in \fR$.

(2): Let $\bA\bv = \lambda\bv$ and $\bA\bw = \mu\bw$ with $\lambda \neq \mu$. Then:
\[
    \lambda\,\bv^T\bw = (\bA\bv)^T\bw = \bv^T\bA^T\bw = \bv^T\bA\bw = \mu\,\bv^T\bw.
\]
Since $\lambda \neq \mu$, we get $\bv^T\bw = 0$.

(3): Existence of a full orthonormal eigenbasis follows from (1), (2), and the fact that the
characteristic polynomial splits over $\fR$ for symmetric matrices (proven via induction on
dimension in a full course).
}

\addDef{Hermitian Matrix}{%
A matrix $\bA \in \fC^{n \times n}$ is \emph{Hermitian} if $\bA = \bA^*$, where
$\bA^* = \bar{\bA}^T$ is the \emph{conjugate transpose}: each entry $a_{ij}$ is
replaced by its complex conjugate $\overline{a_{ji}}$. For real matrices, $\bA^* = \bA^T$,
so symmetric matrices are exactly the real Hermitian matrices.
}
\label{def:hermitian}

\addExcr{%
Refer to Def~\ref{def:hermitian}.
\begin{enumerate}
    \item Verify that $\bA = \begin{pmatrix} 2 & 1+i \\ 1-i & 3 \end{pmatrix}$ is
    Hermitian by computing $\bA^*$.
    \item Show that for any $\bA \in \fC^{n \times n}$, the matrix $\bA + \bA^*$ is
    always Hermitian and $\bA - \bA^*$ is always \emph{skew-Hermitian} ($\bM^* = -\bM$).
    \item Show that the diagonal entries of a Hermitian matrix must be real.
\end{enumerate}
}

\addThm{Spectral Theorem for Hermitian Matrices}{%
Let $\bA \in \fC^{n \times n}$ be Hermitian. Then:
\begin{enumerate}
    \item All eigenvalues of $\bA$ are real.
    \item Eigenvectors for distinct eigenvalues are orthogonal with respect to the
    complex inner product $\langle \bu, \bv \rangle = \bu^*\bv$.
    \item $\bA$ is unitarily diagonalizable: $\bA = \bQ\bD\bQ^*$, where
    $\bQ \in \fC^{n \times n}$ is unitary ($\bQ^*\bQ = \bI$) and
    $\bD = \mathrm{diag}(\lambda_1, \ldots, \lambda_n) \in \fR^{n \times n}$.
\end{enumerate}
}
\label{thm:spectral-theorem-hermitian}

\addPrf{%
The proof mirrors the real symmetric case, replacing transposes with conjugate transposes.

\medskip
\textbf{(1) Eigenvalues are real.}
Let $\bA\bv = \lambda\bv$ with $\bv \neq \bzero$. Compute $\bv^*\bA\bv$ in two ways.
On one hand, $\bv^*\bA\bv = \bv^*(\lambda\bv) = \lambda\|\bv\|_2^2$.
On the other, since $\bA = \bA^*$:
\[
  \bv^*\bA\bv = \bv^*\bA^*\bv = (\bA\bv)^*\bv = (\lambda\bv)^*\bv = \bar{\lambda}\|\bv\|_2^2.
\]
Since $\|\bv\|_2^2 > 0$, we get $\lambda = \bar{\lambda}$, so $\lambda \in \fR$.

\medskip
\textbf{(2) Eigenvectors for distinct eigenvalues are orthogonal.}
Let $\bA\bv = \lambda\bv$ and $\bA\bw = \mu\bw$ with $\lambda \neq \mu$ (both real).
Then:
\[
  \lambda\,\bv^*\bw = (\bA\bv)^*\bw = \bv^*\bA^*\bw = \bv^*\bA\bw = \mu\,\bv^*\bw.
\]
Since $\lambda \neq \mu$, we conclude $\bv^*\bw = 0$.

\medskip
\textbf{(3) Unitary diagonalizability.}
That a complete unitary eigenbasis always exists follows from induction on dimension
(analogous to the real symmetric case) and is covered in a full linear algebra course.
}

\addNote{%
\textbf{Hermitian vs.\ symmetric, unitary vs.\ orthogonal.}
The complex analogs pair up cleanly:
real symmetric $\leftrightarrow$ complex Hermitian (both have real eigenvalues and an orthonormal eigenbasis),
and real orthogonal $\leftrightarrow$ complex unitary (both preserve the norm of every vector).
The DFT matrix, which you will see in the Fourier analysis section, is unitary but
not Hermitian. The discrete Laplacian matrix $\bT$ from the operators section is real
symmetric, a special case of Hermitian.
}

\addExcr{%
Refer to Thm~\ref{thm:spectral-theorem-hermitian}.
\begin{enumerate}
    \item Find the eigenvalues and a unitary eigenbasis for
    $\bA = \begin{pmatrix} 2 & 1+i \\ 1-i & 3 \end{pmatrix}$.
    Verify the eigenvalues are real and the eigenvectors are orthogonal.
    \item In NumPy, \texttt{np.linalg.eigh} is designed for Hermitian matrices.
    Apply it to the matrix above and confirm it returns real eigenvalues.
    What happens if you use \texttt{np.linalg.eig} instead?
    \item Show that $\bA^2$ is also Hermitian whenever $\bA$ is Hermitian.
    What are the eigenvalues of $\bA^2$ in terms of those of $\bA$?
\end{enumerate}
}

\addDef{Diagonalization}{%
A matrix $\bA \in \fR^{n \times n}$ is \emph{diagonalizable} if there exists an invertible
matrix $\bV$ and a diagonal matrix $\mathbf{\Lambda}$ such that
\[
\bA = \bV\mathbf{\Lambda}\bV^{-1},
\]
where the diagonal entries of $\mathbf{\Lambda}$ are the eigenvalues of $\bA$ and the columns of
$\bV$ are the corresponding eigenvectors.
}
\label{def:diagonalization}

\addNote{%
A matrix $\bA$ is diagonalizable if and only if there exist $n$ linearly independent
eigenvectors, which occurs if and only if the geometric multiplicity of each eigenvalue
equals its algebraic multiplicity.
}

\textBox{%
The eigenvalue problem is one of the most fundamental in numerical linear algebra, with
applications ranging from structural engineering (vibration analysis) to quantum mechanics
(energy states), machine learning (principal component analysis), and dynamical systems
(stability analysis). The efficiency and accuracy of eigenvalue algorithms have direct
implications for these applications, making them a central focus in scientific computing.
}

\addExcr{%
Refer to Thm~\ref{thm:spectral-theorem} and Def~\ref{def:diagonalization}.
\begin{enumerate}
    \item For $\bA = \begin{pmatrix}4&1\\1&4\end{pmatrix}$, find eigenvalues and an
    orthonormal eigenbasis $\{\bq_1, \bq_2\}$. Write $\bA = \bQ\bD\bQ^T$ explicitly.
    \item Use the decomposition to compute $\bA^{10}$ efficiently.
    (\textit{Hint: $\bA^k = \bQ\bD^k\bQ^T$, and $\bD^k$ is easy.})
    \item In NumPy, use \texttt{np.linalg.eig(A)} and \texttt{np.linalg.eigh(A)} (the
    latter exploits symmetry). Compare the results and explain why \texttt{eigh} is
    preferred for symmetric matrices.
\end{enumerate}
}

\addExcr{%
For each matrix, find all eigenvalues and a corresponding eigenvector for each.
\begin{enumerate}
    \item $\bA = \begin{pmatrix} 5 & 2 \\ 2 & 2 \end{pmatrix}$ \quad (verify using trace
    and determinant)
    \item $\bB = \begin{pmatrix} 0 & -1 \\ 1 & 0 \end{pmatrix}$ \quad (rotation by 90°;
    are eigenvalues real?)
    \item $\mathbf{C} = \begin{pmatrix} 1 & 1 & 0 \\ 0 & 2 & 0 \\ 0 & 0 & 2 \end{pmatrix}$
    \quad (use Prop~\ref{prop:triangular-eigenvalues})
\end{enumerate}
}

\addExcr{%
\textbf{(Diagonalizability and the Geometric Multiplicity)} The matrix
$\bA = \begin{pmatrix} 2 & 1 \\ 0 & 2 \end{pmatrix}$ has a repeated eigenvalue $\lambda = 2$.
\begin{enumerate}
    \item Find the eigenspace $\ker(\bA - 2\bI)$. What is its dimension (geometric
    multiplicity)?
    \item Is $\bA$ diagonalizable? Why or why not?
    \item Compute $\bA^k$ for general $k$ using the Jordan form $\bA = 2\bI + \mathbf{N}$
    where $\mathbf{N} = \begin{pmatrix}0&1\\0&0\end{pmatrix}$ is nilpotent. (\textit{Hint:
    $\mathbf{N}^2 = \mathbf{0}$, so $(2\bI + \mathbf{N})^k = \sum_{j=0}^{1}\binom{k}{j}
    2^{k-j}\mathbf{N}^{j}$.})
    \item What does \texttt{np.linalg.eig(A)} return for this matrix? Are the eigenvectors
    linearly independent?
\end{enumerate}
}

\subsubsection{Power Iteration and Inverse Iteration}

\textBox{%
The spectral theorem tells us eigenvalues exist and what they look like structurally. Power
iteration gives a practical way to \emph{compute} them. The fundamental idea is that
repeated multiplication by $\bA$ amplifies the dominant eigenvector component --- the one
corresponding to the largest eigenvalue --- while the rest decay away.
}

\addDef{Power Iteration}{%
Given $\bA \in \fR^{n\times n}$ and a starting vector $\bv^{(0)}$ with
$\|\bv^{(0)}\|_2 = 1$, the \emph{power iteration} produces a sequence of unit vectors and
Rayleigh quotients:
\[
  \bw^{(k)} = \bA\bv^{(k-1)}, \quad
  \bv^{(k)} = \frac{\bw^{(k)}}{\|\bw^{(k)}\|_2}, \quad
  \lambda^{(k)} = (\bv^{(k)})^T\bA\bv^{(k)}.
\]
The scalar $\lambda^{(k)}$ is the \emph{Rayleigh quotient} at step $k$ and serves as an
estimate of the dominant eigenvalue.
}
\label{def:power-iteration}

\addThm{Convergence of Power Iteration}{%
Let $\bA \in \fR^{n\times n}$ be diagonalizable with eigenvalues
$|\lambda_1| > |\lambda_2| \geq \cdots \geq |\lambda_n|$ and corresponding eigenvectors
$\{\bv_1, \ldots, \bv_n\}$. Suppose $\bv^{(0)}$ has a nonzero component in the $\bv_1$
direction. Then:
\[
  \bv^{(k)} \to \pm\bv_1 \quad \text{and} \quad \lambda^{(k)} \to \lambda_1
  \quad \text{as } k \to \infty,
\]
with convergence rate $|\lambda_2/\lambda_1|^k$.
}
\label{thm:power-iteration}

\addPrf{%
Write $\bv^{(0)} = c_1\bv_1 + c_2\bv_2 + \cdots + c_n\bv_n$ with $c_1 \neq 0$. Then:
\[
  \bA^k\bv^{(0)} = c_1\lambda_1^k\bv_1 + \cdots + c_n\lambda_n^k\bv_n
  = \lambda_1^k\Bigl(c_1\bv_1 + \sum_{i>1} c_i\Bigl(\frac{\lambda_i}{\lambda_1}\Bigr)^k\bv_i\Bigr).
\]
Since $|\lambda_i/\lambda_1| < 1$ for $i > 1$, the sum decays to zero and
$\bA^k\bv^{(0)}/\|\bA^k\bv^{(0)}\| \to \pm\bv_1$. The Rayleigh quotient satisfies
$\lambda^{(k)} \to \lambda_1$ with error $O(|\lambda_2/\lambda_1|^{2k})$.
}

\addNote{%
Power iteration fails when: (1) $|\lambda_1| = |\lambda_2|$ (e.g., complex conjugate pairs);
(2) $\bv^{(0)}$ has no component along $\bv_1$ (measure zero, but can happen numerically).
Normalization at each step prevents overflow/underflow. The convergence rate $|\lambda_2/\lambda_1|$
can be very slow if the two dominant eigenvalues are close in magnitude.
}

\addDef{Inverse Iteration}{%
\emph{Inverse iteration} applies power iteration to $(\bA - \sigma\bI)^{-1}$ for a shift
$\sigma$ near a target eigenvalue $\mu$. The iteration is:
\[
  \text{Solve } (\bA - \sigma\bI)\bw^{(k)} = \bv^{(k-1)}, \quad
  \bv^{(k)} = \frac{\bw^{(k)}}{\|\bw^{(k)}\|_2}.
\]
The eigenvalue estimate is updated as the Rayleigh quotient
$\lambda^{(k)} = (\bv^{(k)})^T\bA\bv^{(k)}$.
}
\label{def:inverse-iteration}

\addNote{%
When $\sigma$ is close to $\mu$, the matrix $\bA - \sigma\bI$ is nearly singular, but the
solve still yields the correct eigenvector direction. In practice, $\bA - \sigma\bI$ is
factored once (LU or Cholesky), making each inverse iteration step $O(n^2)$ (for dense
$\bA$) instead of $O(n^3)$. The convergence rate is $|(\mu - \sigma)/(\tilde{\mu} - \sigma)|$
where $\tilde{\mu}$ is the nearest eigenvalue to $\sigma$ besides $\mu$.
}

\addDef{Rayleigh Quotient Iteration}{%
\emph{Rayleigh Quotient Iteration} (RQI) applies inverse iteration with the shift
$\sigma_k = \lambda^{(k)}$ updated at each step. Starting from a unit vector $\bv^{(0)}$:
\[
  \sigma_k = (\bv^{(k)})^T\bA\bv^{(k)}, \quad
  \text{solve } (\bA - \sigma_k\bI)\bw^{(k+1)} = \bv^{(k)}, \quad
  \bv^{(k+1)} = \bw^{(k+1)}/\|\bw^{(k+1)}\|.
\]
RQI converges \emph{cubically} for symmetric matrices: once near an eigenvector, the number
of correct digits roughly triples each step.
}
\label{def:rayleigh-quotient-iter}

\addExmpl{%
\textbf{Power iteration on a $3\times3$ symmetric matrix.}

Let $\bA = \begin{pmatrix}4&1&0\\1&3&1\\0&1&2\end{pmatrix}$,
$\bv^{(0)} = (1,0,0)^T$. The eigenvalues are approximately
$\lambda_1 \approx 5.21$, $\lambda_2 \approx 2.79$, $\lambda_3 \approx 1.00$.

Starting power iteration:
\begin{align*}
\bw^{(1)} &= \bA\bv^{(0)} = (4,1,0)^T, \quad
\bv^{(1)} = (4,1,0)^T/\sqrt{17} \approx (0.970, 0.243, 0)^T, \\
\lambda^{(1)} &= (\bv^{(1)})^T\bA\bv^{(1)} \approx 4.12.
\end{align*}
The ratio $|\lambda_2/\lambda_1| \approx 0.535$, so convergence is moderately fast;
after $\sim 10$ iterations the estimate $\lambda^{(k)} \approx 5.21$ to 4 digits.
}

\addExcr{%
Refer to Defs~\ref{def:power-iteration}--\ref{def:rayleigh-quotient-iter} and
Thm~\ref{thm:power-iteration}.
\begin{enumerate}
  \item Implement \texttt{power\_iteration(A, v0, tol, max\_iter)} that returns the
  dominant eigenvalue and eigenvector. Apply it to the tridiagonal $5\times5$ matrix with
  $3$ on the diagonal and $-1$ on the off-diagonals. Plot the Rayleigh quotient error
  vs.\ $k$ on a semi-log scale; verify the slope equals $\log|\lambda_2/\lambda_1|$.
  \item Implement inverse iteration to find the \emph{smallest} eigenvalue of the same
  matrix (use $\sigma = 0$). Compare the convergence rate to power iteration.
  \item For a random $4\times4$ symmetric matrix, apply Rayleigh Quotient Iteration from a
  random starting vector. Plot the error in the eigenvalue estimate vs.\ $k$ on a log
  scale. Can you observe cubic convergence?
  \item What happens to power iteration when $|\lambda_1| = |\lambda_2|$? Construct an
  example (e.g., a rotation matrix scaled to have two equal-magnitude eigenvalues) and
  observe the behavior in NumPy.
\end{enumerate}
}

\subsubsection{Invariant Subspaces}

Eigenvectors are special one-dimensional invariant subspaces. The notion of an invariant
subspace generalizes this: a subspace that $\bA$ maps into itself. Understanding invariant
subspaces is the key to the Schur decomposition and the convergence of the QR algorithm.

\addDef{Invariant Subspace}{%
A subspace $\mathcal{S} \subseteq \fC^n$ is \emph{invariant} under $\bA \in \fC^{n \times n}$
if
\[
  \bA\mathcal{S} \subseteq \mathcal{S}, \qquad
  \text{i.e., } \bA\bv \in \mathcal{S} \text{ for every } \bv \in \mathcal{S}.
\]
}
\label{def:invariant-subspace}

\addExmpl{%
The canonical invariant subspaces of $\bA$:
\begin{itemize}
  \item $\{\bzero\}$ and $\fC^n$ --- trivially invariant for any $\bA$.
  \item $\text{span}\{\bv\}$ for any eigenvector $\bv$ --- since
  $\bA\bv = \lambda\bv \in \text{span}\{\bv\}$.
  \item $\ker(\bA - \lambda\bI)^k$ --- the \emph{generalized eigenspace} for eigenvalue
  $\lambda$ (Jordan block structure).
  \item $\mathcal{R}(\bA)$ --- the column space is invariant since
  $\bA^2\bx = \bA(\bA\bx) \in \mathcal{R}(\bA)$ for all $\bx$.
\end{itemize}
}

\addThm{Invariant Subspace and Block Triangular Form}{%
Let $\mathcal{S} = \text{span}\{\bq_1, \ldots, \bq_k\}$ be a $k$-dimensional invariant
subspace of $\bA$, with $\bQ_1 = [\bq_1, \ldots, \bq_k]$ having orthonormal columns.
Extend to an orthonormal basis $\bQ = [\bQ_1 \mid \bQ_2]$ of $\fC^n$. Then:
\[
  \bQ^*\bA\bQ = \begin{pmatrix} \bA_{11} & \bA_{12} \\ \bzero & \bA_{22} \end{pmatrix},
\]
where $\bA_{11} \in \fC^{k \times k}$. The eigenvalues of $\bA_{11}$ are a subset of the
eigenvalues of $\bA$.
}
\label{thm:invariant-block}

\addPrf{%
Since $\mathcal{S}$ is invariant, $\bA\bQ_1 = \bQ_1\bA_{11}$ for some
$\bA_{11} \in \fC^{k\times k}$. Multiplying on the left by $\bQ^*$:
\[
  \bQ^*\bA\bQ = \bQ^*[\bA\bQ_1 \mid \bA\bQ_2]
  = \begin{pmatrix}\bQ_1^*\bQ_1\bA_{11} & \bQ_1^*\bA\bQ_2 \\
  \bQ_2^*\bQ_1\bA_{11} & \bQ_2^*\bA\bQ_2\end{pmatrix}
  = \begin{pmatrix}\bA_{11} & \bA_{12} \\ \bzero & \bA_{22}\end{pmatrix},
\]
using $\bQ_1^*\bQ_1 = \bI$, $\bQ_2^*\bQ_1 = \bzero$ (orthonormality). The eigenvalues
of the block upper triangular matrix are the union of eigenvalues of $\bA_{11}$ and
$\bA_{22}$, so $\sigma(\bA_{11}) \subseteq \sigma(\bA)$.
}

\addNote{%
This is why invariant subspaces are central to eigenvalue computation. If you can find a
nontrivial invariant subspace $\mathcal{S}$, you \emph{deflate} the problem: reduce finding
all $n$ eigenvalues to finding the eigenvalues of $\bA_{11}$ (dimension $k$) and $\bA_{22}$
(dimension $n-k$) independently. The QR algorithm works precisely by building up an
invariant subspace one eigenvector (or complex-conjugate pair) at a time.
}

\addDef{Eigenspace and Generalized Eigenspace}{%
For $\lambda \in \sigma(\bA)$, the \emph{eigenspace} is $\ker(\bA - \lambda\bI)$ --- the
set of all eigenvectors (plus $\bzero$). Its dimension is the \emph{geometric multiplicity}
of $\lambda$. The \emph{generalized eigenspace} is $\ker(\bA - \lambda\bI)^n$, which
captures all vectors associated with $\lambda$ through the Jordan structure. Both are
invariant subspaces.
}
\label{def:eigenspace}

\addExcr{%
Refer to Def~\ref{def:eigenspace} and Thm~\ref{thm:invariant-block}.
\begin{enumerate}
    \item For $\bA = \begin{pmatrix}3&1\\0&3\end{pmatrix}$, find the eigenspace
    $\ker(\bA - 3\bI)$ and the generalized eigenspace $\ker(\bA - 3\bI)^2$. What is the
    dimension of each?
    \item Identify two nontrivial invariant subspaces of
    $\bB = \begin{pmatrix}3&0\\0&5\end{pmatrix}$ and write the block triangular form
    $\bQ^T\bB\bQ$ for each.
    \item In NumPy, compute the Schur decomposition of a random $4\times4$ matrix and
    verify that $\text{span}\{\bq_1, \bq_2\}$ (first two Schur vectors) is invariant:
    check that $\bA\bQ[:, :2]$ lies in the column space of $\bQ[:, :2]$.
\end{enumerate}
}

\addDef{Stable, Unstable, and Center Subspaces}{%
For a matrix $\bA$ representing a dynamical system $\bx^{(k+1)} = \bA\bx^{(k)}$, the
eigenvalues partition the spectrum into three sets:
\begin{itemize}
  \item $\sigma^- = \{\lambda \in \sigma(\bA) : |\lambda| < 1\}$ --- \emph{stable}: orbits
  decay to zero.
  \item $\sigma^+ = \{\lambda \in \sigma(\bA) : |\lambda| > 1\}$ --- \emph{unstable}: orbits
  grow without bound.
  \item $\sigma^0 = \{\lambda \in \sigma(\bA) : |\lambda| = 1\}$ --- \emph{center}: orbits
  neither grow nor decay.
\end{itemize}
The corresponding generalized eigenspaces span the \emph{stable subspace} $\mathcal{S}^-$,
\emph{unstable subspace} $\mathcal{S}^+$, and \emph{center subspace} $\mathcal{S}^0$ ---
each invariant under $\bA$, and $\fC^n = \mathcal{S}^- \oplus \mathcal{S}^+ \oplus
\mathcal{S}^0$.
}
\label{def:stable-unstable}

\addNote{%
In the continuous-time system $\dot{\bx} = \bA\bx$, stability is governed by the
\emph{real parts} of eigenvalues instead: $\sigma^- = \{\lambda : \text{Re}(\lambda) < 0\}$
(stable), etc. The stable/unstable subspace decomposition is identical in structure --- only
the threshold changes from $|\lambda|=1$ to $\text{Re}(\lambda)=0$.
}

\addExcr{%
\textbf{(Identifying and using invariant subspaces)} Refer to
Defs~\ref{def:invariant-subspace}--\ref{def:stable-unstable} and
Thm~\ref{thm:invariant-block}.
\begin{enumerate}
  \item For $\bA = \begin{pmatrix}3&1&0\\0&3&0\\0&0&5\end{pmatrix}$, identify two
  nontrivial invariant subspaces. For each, write the block triangular form
  $\bQ^T\bA\bQ$ explicitly and read off the eigenvalues.
  \item Prove that if $\mathcal{S}$ and $\mathcal{T}$ are both invariant under $\bA$,
  then so are $\mathcal{S} \cap \mathcal{T}$ and
  $\mathcal{S} + \mathcal{T} = \{\bu + \bw : \bu \in \mathcal{S},\, \bw \in \mathcal{T}\}$.
  \item For a $4\times4$ matrix $\bA$ with eigenvalues $0.5, 0.8, 1.2, 1.5$, identify
  which belong to the stable and unstable subspaces (discrete-time). What are the
  dimensions of each?
  \item In NumPy, compute the Schur decomposition of a random $4\times4$ matrix. Verify
  that the first column of $\bQ$ is an eigenvector and that $\text{span}\{\bq_1, \bq_2\}$
  is invariant: check that $\bA\bQ[:, :2]$ lies in the column space of $\bQ[:, :2]$.
\end{enumerate}
}

\subsubsection{Schur Decomposition}

\textBox{%
Not every matrix is diagonalizable, but every square matrix can be reduced to triangular
form via a unitary similarity. The Schur decomposition is the theoretical foundation for
practical eigenvalue algorithms: the QR algorithm computes the Schur form iteratively,
revealing all eigenvalues on the diagonal of the triangular factor.
}

\addThm{Schur Decomposition}{%
Every matrix $\bA \in \fC^{n \times n}$ has a \emph{Schur decomposition}
\[
  \bA = \bQ\bT\bQ^*,
\]
where $\bQ \in \fC^{n \times n}$ is unitary ($\bQ^*\bQ = \bI$) and $\bT \in \fC^{n \times n}$
is upper triangular. The diagonal entries of $\bT$ are the eigenvalues of $\bA$ (in any
order). If $\bA \in \fR^{n \times n}$ has all real eigenvalues, then $\bQ$ and $\bT$ can be
taken to be real.
}
\label{thm:schur-decomp}

\addPrf{%
By induction on $n$. The base case $n = 1$ is trivial. For the inductive step, let
$\lambda_1$ be any eigenvalue of $\bA$ (it exists over $\fC$) with unit eigenvector
$\bq_1$. Extend $\bq_1$ to an orthonormal basis $\{\bq_1, \bu_2, \ldots, \bu_n\}$ and let
$\bQ_1 = [\bq_1 \mid \bu_2 \mid \cdots \mid \bu_n]$. Then:
\[
  \bQ_1^*\bA\bQ_1 = \begin{pmatrix} \lambda_1 & \star \\ \bzero & \bB \end{pmatrix},
\]
where $\bB \in \fC^{(n-1)\times(n-1)}$. By the inductive hypothesis,
$\bB = \bQ_2\bT'\bQ_2^*$ for some unitary $\bQ_2$ and upper triangular $\bT'$. Setting
$\bQ = \bQ_1\,\mathrm{diag}(1, \bQ_2)$ gives the Schur decomposition with
$\bT = \begin{pmatrix}\lambda_1 & \star \\ \bzero & \bT'\end{pmatrix}$.
}

\addNote{%
The Schur decomposition is a generalization of diagonalization: if $\bA$ is normal
($\bA^*\bA = \bA\bA^*$), then $\bT$ is diagonal (the Spectral Theorem). For non-normal
matrices, off-diagonal entries of $\bT$ are generally nonzero, but the eigenvalues are
still read from the diagonal.
}

\addDef{Real Schur Form}{%
For $\bA \in \fR^{n \times n}$, the \emph{real Schur decomposition} is $\bA = \bQ\bT\bQ^T$
where $\bQ \in \fR^{n \times n}$ is orthogonal and $\bT$ is \emph{quasi-upper triangular}:
block upper triangular with $1\times1$ diagonal blocks (for real eigenvalues) and $2\times2$
diagonal blocks (for complex-conjugate eigenvalue pairs). This avoids complex arithmetic
entirely and is the form computed by \texttt{scipy.linalg.schur(A)}.
}
\label{def:real-schur}

\addDef{Schur Vectors}{%
The columns of $\bQ$ in the Schur decomposition $\bA = \bQ\bT\bQ^*$ are called
\emph{Schur vectors}. Unlike eigenvectors, Schur vectors always exist and form a complete
orthonormal basis, even when $\bA$ is defective. The first $k$ Schur vectors span an
invariant subspace of $\bA$: if $\bQ_k = [\bq_1, \ldots, \bq_k]$, then
$\bA\bQ_k = \bQ_k\bT_{11}$ where $\bT_{11}$ is the leading $k\times k$ block of $\bT$.
}
\label{def:schur-vectors}

\addExcr{%
Refer to Defs~\ref{def:real-schur}--\ref{def:schur-vectors} and
Thm~\ref{thm:schur-decomp}.
\begin{enumerate}
    \item For $\bA = \begin{pmatrix}3&1&0\\0&2&1\\0&0&2\end{pmatrix}$, is $\bA$ already in
    Schur form? What are its Schur vectors? What are the eigenvalues?
    \item For a $2\times2$ matrix with complex-conjugate eigenvalues $a \pm bi$, write out
    the $2\times2$ real Schur form block explicitly in terms of $a$ and $b$.
    \item In Python, compute \texttt{T, Q = scipy.linalg.schur(A)} for a random
    $4\times4$ matrix. Verify: (a) $\bQ\bT\bQ^T = \bA$, (b) $\bQ$ is orthogonal, (c) the
    diagonal of $\bT$ matches \texttt{np.linalg.eigvals(A)}.
\end{enumerate}
}

\addThm{Connection to the QR Algorithm}{%
The basic QR algorithm for computing eigenvalues proceeds by iteratively applying QR
factorizations:
\[
  \bA_0 = \bA, \quad \bA_k = \bR_k\bQ_k, \quad
  \bA_{k+1} = \bQ_k\bR_k = \bQ_k^T\bA_k\bQ_k.
\]
Each $\bA_{k+1}$ is orthogonally similar to $\bA$, so all matrices share the same
eigenvalues. Under mild conditions, the sequence $\bA_k$ converges to the Schur form
$\bT$, with eigenvalues appearing on the diagonal.
}
\label{thm:qr-algorithm}

\addExcr{%
In this exercise, we explore two practical enhancements of the basic QR algorithm
(Thm~\ref{thm:qr-algorithm}): Hessenberg reduction and shifts.
\begin{enumerate}
    \item \textbf{(Hessenberg reduction cost.)}
    An upper Hessenberg matrix has zeros below the first subdiagonal.
    A single QR factorization step on a general $n \times n$ matrix costs $O(n^3)$ flops,
    but on an upper Hessenberg matrix it costs only $O(n^2)$.
    Explain why: how many Givens rotations are needed to reduce an upper Hessenberg matrix
    to triangular form, and what does each rotation cost?
    \item \textbf{(Shifts preserve eigenvalues.)}
    A shifted QR step factors $\bA_k - \sigma_k\bI = \bQ_k\bR_k$, then forms
    $\bA_{k+1} = \bR_k\bQ_k + \sigma_k\bI$.
    Show that $\bA_{k+1}$ is orthogonally similar to $\bA_k$, and hence has the same
    eigenvalues.
    \item \textbf{(Convergence comparison in NumPy.)}
    Let $\bA = \begin{pmatrix}4&1&0\\1&3&1\\0&1&2\end{pmatrix}$.
    \begin{enumerate}
        \item Run 30 basic QR steps (no shifts) and record $|(\bA_k)_{32}|$ at each step.
        \item Repeat with the Rayleigh quotient shift $\sigma_k = (\bA_k)_{nn}$.
        Compare the number of steps required to reach $|(\bA_k)_{32}| < 10^{-10}$.
    \end{enumerate}
    \item Verify using \texttt{np.linalg.eig(A)} that the final diagonal entries of your
    converged $\bA_k$ match the computed eigenvalues.
    What LAPACK routine does \texttt{np.linalg.eig} call internally?
\end{enumerate}
}

\addExcr{%
\textbf{(Schur decomposition in practice)}
\begin{enumerate}
  \item For $\bA = \begin{pmatrix}3&1&0\\0&2&1\\0&0&2\end{pmatrix}$, compute the Schur
  decomposition by hand. Is $\bA$ already in Schur form? What are its Schur vectors?
  \item In Python, compute \texttt{T, Q = scipy.linalg.schur(A)} for a random $4\times4$
  matrix. Verify: (a) $\bQ\bT\bQ^T = \bA$, (b) $\bQ$ is orthogonal, (c) the diagonal of
  $\bT$ matches \texttt{np.linalg.eigvals(A)}.
  \item For the non-diagonalizable matrix $\bA = \begin{pmatrix}2&1\\0&2\end{pmatrix}$,
  find its Schur decomposition. How does it differ from the (impossible) diagonalization?
  \item Implement one step of the basic QR iteration: compute $\bA_1 = \bR_0\bQ_0$ where
  $\bA_0 = \bQ_0\bR_0$ is the QR factorization. Apply 50 steps to a random $3\times3$
  symmetric matrix and verify that the off-diagonal entries converge to zero.
\end{enumerate}
}

\end{document}
