\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Floating Point Arithmetic}

Real numbers cannot be represented exactly on digital computers. This subsection covers
how numbers are stored in the IEEE 754 floating point format, the precision limits of that
representation, and the special values and arithmetic pitfalls that arise in practice.
Understanding these fundamentals is a prerequisite for analyzing any numerical algorithm.

\subsubsection{Floating Point Representation}

Real numbers are stored in computers using a finite-bit format analogous to scientific
notation.  The key parameters are the number of bits for the sign, exponent, and
significand.

\addDef{Floating Point Representation}{%
A normalized floating point number has the form
\[
x = (-1)^s \times (1.f) \times 2^e,
\]
where $s$ is the sign bit, $f$ is the fractional significand (mantissa), $e$ is the
exponent, and the leading 1 is implicit (hidden bit).
The IEEE 754 standard defines two primary formats:
\begin{itemize}
    \item \textbf{Single precision (32-bit)}: 1 sign, 8 exponent, 23 fraction bits.
    \item \textbf{Double precision (64-bit)}: 1 sign, 11 exponent, 52 fraction bits.
\end{itemize}
The exponent field uses a biased representation with bias $= 2^{k-1} - 1$
($k$ = exponent bits).
}
\label{def:fp-repr}

\addExmpl{%
\textbf{16-bit floating point.}\enspace
With 1 sign, 5 exponent (bias $= 15$), and 10 fraction bits, represent $-3.25$:
\begin{enumerate}
    \item Sign $= 1$ (negative).
    \item $3.25_{10} = 11.01_2$.
    \item Normalize: $1.101_2 \times 2^1$.
    \item Biased exponent: $1 + 15 = 16 = 10000_2$.
    \item Fraction bits: $1010000000$.
    \item Final: $1\;10000\;1010000000$.
\end{enumerate}
}

\addExcr{%
Refer to Def~\ref{def:fp-repr}.
\begin{enumerate}
    \item Encode the number $+5.5$ in the 16-bit format above (1 sign, 5 exponent, 10
    fraction bits, bias $= 15$).  Show each step.
    \item How many distinct normalized numbers can be represented with 5 exponent bits and
    10 fraction bits (excluding special values)?
    \textit{Hint: $2 \times (2^5 - 2) \times 2^{10}$.}
\end{enumerate}
}

\addDef{Machine Epsilon}{%
\textbf{Machine epsilon} $\varepsilon_{\text{mach}}$ is the difference between 1 and the
next larger representable floating point number.  For a system with $p$ significand bits
(including the hidden bit):
\[
\varepsilon_{\text{mach}} = 2^{1-p}.
\]
For any normalized floating point number $x$, the gap to the next representable number is
approximately $\varepsilon_{\text{mach}} \cdot |x|$.
}
\label{def:machine-epsilon}

\addExmpl{%
For IEEE double precision: $p = 52 + 1 = 53$ bits, so
$\varepsilon_{\text{mach}} = 2^{-52} \approx 2.22 \times 10^{-16}$.
In Python: \texttt{import sys; sys.float\_info.epsilon}.
}

\addExcr{%
Refer to Def~\ref{def:machine-epsilon}.
\begin{enumerate}
    \item In Python, evaluate \texttt{1.0 + 1e-16} and \texttt{1.0 + 2e-16}.
    Explain both results using $\varepsilon_{\text{mach}}$.
    \item Write a function \texttt{compute\_eps()} that finds machine epsilon empirically
    (without \texttt{sys.float\_info.epsilon} or \texttt{np.finfo}).
    \textit{Hint: start from 1.0, halve repeatedly until $1 + \varepsilon = 1$.}
\end{enumerate}
}

\addDef{Unit Roundoff}{%
The \textbf{unit roundoff} is $u = \varepsilon_{\text{mach}}/2 = 2^{-p}$.
It is the maximum relative error when rounding a real number $x$ to its nearest floating
point representation $\text{fl}(x)$:
\[
\frac{|\text{fl}(x) - x|}{|x|} \leq u.
\]
}
\label{def:unit-roundoff}

\addNote{%
Machine epsilon is the smallest number for which $1 + \varepsilon_{\text{mach}} > 1$ in
floating point.  Unit roundoff is the maximum relative rounding error for any number.
In IEEE round-to-nearest mode, $u = \varepsilon_{\text{mach}}/2$.
}

\addExcr{%
Refer to Def~\ref{def:unit-roundoff}.
\begin{enumerate}
    \item Confirm that $(1.0 + u) - 1.0 = u$ but $(1.0 + u/2) - 1.0 = 0$ in double
    precision.  Explain in terms of the spacing between adjacent floats near 1.
    \item How many significant decimal digits does an IEEE double carry?
    \textit{Hint: $u = 2^{-52} \approx 2.22 \times 10^{-16}$; compute $-\log_{10}(u)$.}
\end{enumerate}
}

\addThm{Fundamental Theorem of Floating Point Arithmetic}{%
For any real $x$ in the normalized range, the nearest floating point number satisfies
\[
    \text{fl}(x) = x(1 + \delta), \quad |\delta| \leq u.
\]
Furthermore, for any floating point $a, b$ and operation $\circ \in \{+,-,\times,\div\}$,
IEEE 754 guarantees:
\[
    \text{fl}(a \circ b) = (a \circ b)(1 + \delta), \quad |\delta| \leq u.
\]
}
\label{thm:fp-fundamental}

\addPrf{%
The real line near $x$ is partitioned into intervals each containing exactly one floating
point number (the left endpoint).  Rounding to nearest halves the worst-case error, giving
$|\delta| \leq u$.  For the second statement, IEEE 754 mandates that each basic operation
is computed as if in infinite precision and then rounded, ensuring the same $|\delta| \leq u$.
}

\addNote{%
The IEEE 754 guarantee that each individual operation incurs at most unit roundoff is the
foundation of all backward error analyses.  After $n$ sequential operations, the accumulated
error grows as $O(nu)$, which is acceptable for double precision when $n \ll 10^{16}$.
}

\addExcr{%
Refer to Thm~\ref{thm:fp-fundamental}.
\begin{enumerate}
    \item Suppose $a = 1.0$ and $b = -0.9999999999999999$ in double precision.
    What is $\text{fl}(a + b)$?  Is the relative error bounded by $u$?
    \item A computation performs $10^6$ floating point additions.  Using the $O(nu)$ bound,
    estimate the worst-case accumulated relative error in double precision.
\end{enumerate}
}

\subsubsection{IEEE 754 Special Values}

Beyond normalized numbers, IEEE 754 defines special bit patterns to handle results outside
the representable range.  Each has a distinct behavior that must be understood when writing
numerical code.

\addDef{Signed Zero}{%
Both $+0$ and $-0$ are represented and compare as equal, but behave differently in
arithmetic: $1/(+0) = +\infty$ and $1/(-0) = -\infty$.
}
\label{def:signed-zero}

\addExcr{%
Refer to Def~\ref{def:signed-zero}.
\begin{enumerate}
    \item In Python, evaluate \texttt{1.0/0.0} and \texttt{1.0/(-0.0)}.
    Are the results $+\infty$ and $-\infty$ respectively, or does Python raise an error?
    \item Evaluate \texttt{(+0.0) == (-0.0)}.  What does this confirm about signed zero?
\end{enumerate}
}

\addDef{Subnormal Numbers}{%
Numbers smaller than the smallest normalized value $2^{e_{\min}}$ are represented with a
leading zero bit, allowing \textbf{gradual underflow} at the cost of reduced precision.
The smallest positive double-precision subnormal is $2^{-1074} \approx 4.94 \times 10^{-324}$.
}
\label{def:subnormal}

\addNote{%
Subnormal numbers fill the gap between 0 and the smallest normalized number
($2^{-1022}$ in double precision), but with fewer significant bits.  A subnormal near
$2^{-1073}$ has only 1 bit of precision, making its relative rounding error much larger
than that of a normalized number.
}

\addExcr{%
Refer to Def~\ref{def:subnormal}.
\begin{enumerate}
    \item What is the smallest positive subnormal number in IEEE single precision
    (1 sign, 8 exponent, 23 fraction bits)?
    \item In Python, evaluate \texttt{5e-324} (the smallest positive double).
    What is \texttt{5e-324 / 2}?  Why?
\end{enumerate}
}

\addDef{Infinities}{%
$+\infty$ and $-\infty$ arise from overflow or division of a nonzero number by zero
(e.g., \texttt{1.0/0.0}).  They propagate correctly through arithmetic:
$x + \infty = \infty$ for any finite $x$, and $\infty \cdot 0 = \text{NaN}$.
}
\label{def:infinities}

\addExcr{%
Refer to Def~\ref{def:infinities}.
\begin{enumerate}
    \item In Python, evaluate \texttt{float('inf') + 1},
    \texttt{float('inf') - float('inf')}, and \texttt{0.0 * float('inf')}.
    Classify each result and explain using IEEE 754 rules.
    \item Is \texttt{float('inf') > 1e308} true?  What does this imply about comparisons
    involving infinity?
\end{enumerate}
}

\addDef{NaN (Not a Number)}{%
A \textbf{NaN} is produced by undefined operations: $0/0$, $\infty - \infty$, $\sqrt{-1}$.
Any arithmetic involving a NaN returns a NaN, and \texttt{nan != nan} is \texttt{True}
in Python (NaN is unequal to everything, including itself).
}
\label{def:nan}

\addExcr{%
Refer to Def~\ref{def:nan}.
\begin{enumerate}
    \item Explain why \texttt{float('nan') == float('nan')} returns \texttt{False}.
    What does this imply for checking whether a computation produced a NaN?
    \item In Python, evaluate \texttt{float('nan') > 0} and \texttt{float('nan') < 0}.
    What are the results, and why?
\end{enumerate}
}

\subsubsection{Floating Point Arithmetic Pitfalls}

Floating point arithmetic differs from real arithmetic in several fundamental ways.
The following properties, while counterintuitive, are direct consequences of finite
precision and must be kept in mind when designing algorithms.

\addNote{%
\textbf{Non-Associativity.}\enspace
$(a \oplus b) \oplus c \neq a \oplus (b \oplus c)$ in general.  Each addition rounds its
result, and changing the order changes which intermediate values are rounded.
}

\addExcr{%
\begin{enumerate}
    \item Using 4-significant-digit decimal arithmetic, compute
    $(10000. + 0.001) - 10000.$ and $10000. + (0.001 - 10000.)$.
    Which order is accurate?  Which demonstrates \emph{absorption}?
    \item In Python, verify with \texttt{a=1e10; b=1.0; c=-1e10}: does
    \texttt{(a+b)+c == a+(b+c)}?  Explain.
\end{enumerate}
}

\addNote{%
\textbf{Non-Distributivity.}\enspace
$a \otimes (b \oplus c) \neq (a \otimes b) \oplus (a \otimes c)$ in general, because
$b \oplus c$ is rounded before multiplication, while $a \otimes b$ and $a \otimes c$
are rounded separately before addition.
}

\addExcr{%
\begin{enumerate}
    \item In Python, compute \texttt{a*(b+c)} vs \texttt{a*b + a*c} for
    \texttt{a = 1e15}, \texttt{b = 1.0}, \texttt{c = -1.0}.
    Explain the discrepancy using the definition above.
    \item True or False: if \texttt{a == b} in floating point, then \texttt{a - b == 0.0}
    always.  What about the converse?
\end{enumerate}
}

\addNote{%
\textbf{Absorption.}\enspace
When $|a| \gg |b|$, the sum $\text{fl}(a + b)$ may equal $a$ exactly, because $b$ falls
below the unit roundoff relative to $a$.  The smaller operand contributes nothing.
}

\addExcr{%
\begin{enumerate}
    \item In Python, evaluate \texttt{1e16 + 1.0 - 1e16}.  What do you get and why?
    \item Give an example where absorption causes a running sum of $n$ small numbers to
    yield zero when added to a large number, even though the true sum is nonzero.
\end{enumerate}
}

\subsection{Quantifying Approximation Error}

Floating point arithmetic introduces errors at every operation.  This subsection develops
the vocabulary and tools for measuring those errors and reasoning about how they
propagate through a computation.  Absolute and relative error quantify the gap between
computed and true values; catastrophic cancellation identifies a common failure mode;
and forward and backward analysis provide complementary frameworks for bounding the
total error in an algorithm.

\subsubsection{Absolute and Relative Errors}

Every floating point operation introduces a small rounding error.  Understanding how these
errors arise and propagate determines whether an algorithm produces trustworthy results.

\addDef{Absolute and Relative Error}{%
When a real number $x$ is approximated by $\hat{x}$, the \textbf{absolute error} is
$|\hat{x} - x|$, and the \textbf{relative error} is
\[
\varepsilon = \frac{|\hat{x} - x|}{|x|},
\]
so that $\hat{x} = x(1 + \varepsilon)$.
}
\label{def:abs-rel-error}

\addExmpl{%
Approximating $\pi \approx 3.14159\ldots$:
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Approximation & Absolute Error & Relative Error \\ \hline
$3.14$ & $0.00159\ldots$ & $0.051\%$ \\
$3.1$ & $0.0416\ldots$ & $1.32\%$ \\
$3$ & $0.1416\ldots$ & $4.51\%$ \\ \hline
\end{tabular}
\end{center}
Relative error is usually more meaningful: a tiny absolute error can still represent a
large relative error when the true value is small.
}

\addExcr{%
Refer to Def~\ref{def:abs-rel-error}.
\begin{enumerate}
    \item Approximate $\sqrt{2} \approx 1.414$.  Compute the absolute and relative errors.
    \item For two computations with relative errors $\varepsilon_1$ and $\varepsilon_2$,
    what is the relative error in their \emph{product}?  (\textit{Hint: use
    $\hat{x}\hat{y} = xy(1+\varepsilon_1)(1+\varepsilon_2) \approx xy(1 + \varepsilon_1
    + \varepsilon_2)$ for small errors.})
\end{enumerate}
}

\addDef{Catastrophic Cancellation}{%
\textbf{Catastrophic cancellation} occurs when subtracting two nearly equal numbers.
If $\hat{x} = x(1+\varepsilon_x)$ and $\hat{y} = y(1+\varepsilon_y)$ with $x \approx y > 0$,
the relative error in $\hat{x} - \hat{y}$ is
\[
\frac{(\hat{x}-\hat{y})-(x-y)}{x-y} = \frac{x\varepsilon_x - y\varepsilon_y}{x - y},
\]
which is unbounded as $x \to y^+$ --- a total loss of significant digits.
}
\label{def:cancellation}

\addThm{Cancellation Error Bound}{%
Let $\hat{x} = x(1+\varepsilon_x)$, $\hat{y} = y(1+\varepsilon_y)$ with
$|\varepsilon_x|, |\varepsilon_y| \leq u$ and $x > y > 0$.  Then
\[
    \left|\frac{(\hat{x}-\hat{y})-(x-y)}{x-y}\right| \leq u\cdot\frac{x+y}{x-y}.
\]
The amplification factor $(x+y)/(x-y) \to \infty$ as $x \to y^+$.
}
\label{thm:cancellation-bound}

\addPrf{%
The error is $x\varepsilon_x - y\varepsilon_y$.  Taking absolute value and dividing by
$x-y$:
\[
    \frac{|x\varepsilon_x - y\varepsilon_y|}{x-y}
    \leq \frac{x|\varepsilon_x| + y|\varepsilon_y|}{x-y}
    \leq u\cdot\frac{x+y}{x-y}.
\]
}

\addExmpl{%
Computing $f(x) = \sqrt{x+1} - \sqrt{x}$ for $x = 10^8$ in double precision:
$\sqrt{10^8+1} \approx 10000.00000005$ and $\sqrt{10^8} = 10000$, so direct subtraction
gives approximately $0$ due to cancellation.  Rationalizing avoids this:
\[
f(x) = \frac{1}{\sqrt{x+1}+\sqrt{x}} \approx \frac{1}{20000} = 5\times 10^{-5}.
\]
Mathematical reformulation eliminates the cancellation without increasing cost.
}

\addExcr{%
\textbf{(Cancellation and Reformulation)}
Refer to Def~\ref{def:cancellation} and Thm~\ref{thm:cancellation-bound}.
For each expression, identify whether catastrophic cancellation can occur and derive a
stable equivalent form.
\begin{enumerate}
    \item $\dfrac{1-\cos x}{x^2}$ for small $x$.
    \textit{Hint: use $1-\cos x = 2\sin^2(x/2)$.}
    \item $\sqrt{x+1} - \sqrt{x}$ for large $x$ (rationalize the numerator).
    \item The smaller root of $x^2 - 10^8 x + 1 = 0$ via the quadratic formula.
    What goes wrong and how do you fix it?
    \item $e^x - 1$ for small $x$.  Why does NumPy provide \texttt{np.expm1(x)}?
    Verify numerically that \texttt{np.expm1(1e-15)} is more accurate than
    \texttt{np.exp(1e-15) - 1}.
\end{enumerate}
}

\subsubsection{Forward and Backward Error Analysis}

Error analysis is the systematic study of how errors enter and propagate through a
computation.  Forward and backward analysis are complementary frameworks; condition number
analysis measures the inherent difficulty of the problem itself, independent of the
algorithm.

\addDef{Forward Error Analysis}{%
\textbf{Forward error analysis} tracks how input errors and per-operation rounding errors
propagate \emph{forward} through the algorithm to bound the output error.
Straightforward but can produce pessimistic overestimates.
}
\label{def:forward-error}

\addExcr{%
Refer to Def~\ref{def:forward-error}.
\begin{enumerate}
    \item For the computation $\hat{z} = \text{fl}(\text{fl}(a+b) + c)$ with
    $a,b,c$ exact, use the model $\text{fl}(x \circ y) = (x \circ y)(1+\delta)$,
    $|\delta| \leq u$, to derive a forward error bound on $|\hat{z} - (a+b+c)|$.
\end{enumerate}
}

\addDef{Backward Error Analysis}{%
\textbf{Backward error analysis} asks: what is the smallest perturbation to the input
that would make the computed output exact?  If this backward error is $O(u\|\bA\|)$, the
algorithm solved a nearby problem exactly --- the best one can hope for in finite precision.
}
\label{def:backward-error}

\addExcr{%
Refer to Def~\ref{def:backward-error}.
\begin{enumerate}
    \item For solving $\bA\bx = \bb$, explain the difference between the forward error
    $\|\hat{\bx} - \bx\|/\|\bx\|$ and the backward error
    $\|\bb - \bA\hat{\bx}\|/(\|\bA\|\|\hat{\bx}\|)$.
    Which is easier to compute after the fact, and why?
    \item Why is backward error analysis generally preferred over forward error analysis
    for evaluating numerical algorithms?
\end{enumerate}
}

\addDef{Interval Arithmetic}{%
In \textbf{interval arithmetic}, each number is replaced by an enclosing interval $[a,b]$,
and every operation maps intervals to intervals in a way that always contains the true
result.  This produces certified error bounds but can suffer from interval explosion for
long computations.
}
\label{def:interval-arith}

\addExcr{%
Refer to Def~\ref{def:interval-arith}.
\begin{enumerate}
    \item Compute $[1.0, 1.1] + [2.0, 2.5]$ and $[1.0, 1.1] \times [2.0, 2.5]$ using
    interval arithmetic rules.  Verify the result contains the true values $1.05 + 2.25$
    and $1.05 \times 2.25$.
\end{enumerate}
}

\addDef{Condition Number Analysis}{%
The \textbf{condition number} of a problem measures its intrinsic sensitivity to small
perturbations in the input, independently of the algorithm.  A problem with condition
number $\kappa$ can lose up to $\log_{10}\kappa$ digits of accuracy even with a perfect
algorithm.  An algorithm is \textbf{backward stable} if its backward error is $O(u)$.
}
\label{def:condition-number-analysis}

\addExcr{%
Refer to Def~\ref{def:condition-number-analysis}.
\begin{enumerate}
    \item An algorithm has backward error $10^{-15}$ applied to a problem with condition
    number $\kappa(\bA) = 10^{8}$.  What is the worst-case forward error?
    Is this acceptable for double-precision computation?
    \item In Python, construct an ill-conditioned matrix using
    \texttt{np.diag([1, 1e-10])}, solve with \texttt{np.linalg.solve}, and compare
    \texttt{np.linalg.cond(A)} with the observed forward error.
\end{enumerate}
}

\addExcr{%
In this exercise, we will prove the backward error formula for linear systems.
Refer to Def~\ref{def:backward-error} and Thm~\ref{thm:fp-fundamental}.

Let $\hat{\bx}$ be the computed solution to $\bA\bx = \bb$ obtained by LU with partial
pivoting.  Define the residual $\br = \bb - \bA\hat{\bx}$.
\begin{enumerate}
    \item Define $\bE = -\br\hat{\bx}^T/\|\hat{\bx}\|^2$.  Show by direct substitution
    that $(\bA + \bE)\hat{\bx} = \bb$, confirming $\hat{\bx}$ is the exact solution to
    a nearby system.
    \item Show that the relative backward error satisfies
    $\|\bE\|_2/\|\bA\|_2 = \|\br\|_2/(\|\bA\|_2\|\hat{\bx}\|_2)$.
    (\textit{Hint: for a rank-1 matrix, $\|\br\bv^T\|_2 = \|\br\|_2\|\bv\|_2$.})
    \item For a well-conditioned system, a small backward error implies a small forward
    error.  Write the bound
    \[
        \frac{\|\hat{\bx}-\bx\|_2}{\|\bx\|_2}
        \leq \kappa(\bA)\cdot\frac{\|\bE\|_2}{\|\bA\|_2},
    \]
    where $\kappa(\bA) = \|\bA\|_2\|\bA^{-1}\|_2$.  If $\kappa(\bA) = 10^{10}$ and the
    backward error is $10^{-16}$, what is the worst-case forward error?
    \item In Python, construct an ill-conditioned matrix using \texttt{np.diag([1, 1e-10])},
    solve with \texttt{np.linalg.solve}, and compare \texttt{np.linalg.cond(A)} with the
    observed forward error.
\end{enumerate}
}

\textBox{%
Floating point arithmetic is not real arithmetic.  Its inherent limitations shape the
design of every numerical algorithm.  By understanding representation, error propagation,
and special cases, we can build algorithms that produce reliable results.  Mathematical
reformulations, careful ordering of operations, and backward error analysis are often more
effective than simply increasing precision.
}

\end{document}
