\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Fourier Analysis as Change of Basis}
\label{sub:fourier}

\textBox{%
The Spectral Theorem says that every symmetric (or Hermitian) operator can be
diagonalized in an orthonormal eigenbasis. For the Laplacian $-d^2/dx^2$, that
eigenbasis is the set of sinusoids. \emph{Fourier analysis is therefore just linear
algebra in an infinite-dimensional space}: expressing a function as a Fourier series
is the same operation as expanding a vector in an eigenbasis. This viewpoint makes
Fourier analysis less magical and reveals exactly why it is so powerful: it
diagonalizes the Laplacian, turning differential equations into simple algebra in
frequency space. The Discrete Fourier Transform (DFT) is the finite-dimensional
version of this, represented as a unitary matrix. Computing it efficiently via the
Fast Fourier Transform (FFT) is one of the most practically impactful algorithms
ever discovered.
}

\subsubsection{The Fourier Basis}

\textBox{%
In Section~\ref{sub:operators}, we found that the eigenfunctions of $-d^2/dx^2$ on
$[0,1]$ are $u_k(x) = \sqrt{2}\sin(k\pi x)$ with eigenvalues $k^2\pi^2$. These
functions are orthonormal in $L^2[0,1]$, and they span the whole space. Expanding any
function $f$ in this basis is the Fourier sine series. On the periodic domain $[0, 2\pi]$,
the natural eigenbasis is the complex exponentials $e^{ikx}$, which give the full
(complex) Fourier series. We work with the periodic setting here because it connects
most directly to the DFT.
}

\addDef{Fourier Series}{%
For $f \in L^2([0, 2\pi])$ (periodic), the \emph{Fourier series} of $f$ is
\[
  f(x) = \sum_{k=-\infty}^{\infty} \hat{f}_k\, e^{ikx},
\]
where the \emph{Fourier coefficients} are
\[
  \hat{f}_k = \frac{1}{2\pi}\int_0^{2\pi} f(x)\,e^{-ikx}\,dx.
\]
The coefficient $\hat{f}_k$ is the inner product of $f$ with the basis function
$e^{ikx}/(2\pi)$, exactly as in a standard change of basis.
}
\label{def:fourier-series}

\addThm{Fourier Basis is Orthonormal}{%
The functions $\phi_k(x) = \frac{1}{\sqrt{2\pi}}e^{ikx}$, $k \in \fZ$, form an
orthonormal set in $L^2([0, 2\pi])$:
\[
  \langle \phi_j, \phi_k \rangle
  = \frac{1}{2\pi}\int_0^{2\pi} e^{ijx}\overline{e^{ikx}}\,dx
  = \frac{1}{2\pi}\int_0^{2\pi} e^{i(j-k)x}\,dx
  = \begin{cases} 1 & j = k, \\ 0 & j \neq k. \end{cases}
\]
Moreover, they form a complete orthonormal basis: every $f \in L^2([0,2\pi])$ can be
recovered exactly from its Fourier coefficients via the Fourier series.
}
\label{thm:fourier-orthonormal}

\addPrf{%
\textbf{Orthonormality.}
For $j \neq k$:
\[
  \frac{1}{2\pi}\int_0^{2\pi} e^{i(j-k)x}\,dx
  = \frac{1}{2\pi} \cdot \frac{e^{i(j-k)x}}{i(j-k)}\Bigg|_0^{2\pi}
  = \frac{e^{2\pi i(j-k)} - 1}{2\pi i(j-k)} = 0,
\]
since $e^{2\pi i m} = 1$ for any integer $m$. For $j = k$, the integrand is 1 and
the integral equals $2\pi$, giving norm 1 after dividing by $2\pi$.

\medskip
\textbf{Completeness.}
Showing that every $L^2$ function is approximated arbitrarily well by finite Fourier
sums is a deeper result (the Riesz-Fischer theorem) that goes beyond this course.
It is the infinite-dimensional analog of the statement that an orthonormal set in
$\fR^n$ that spans $\fR^n$ is a basis.
}

\addNote{%
\textbf{The change-of-basis picture.}
In $\fR^n$, changing from the standard basis $\{\be_j\}$ to an orthonormal basis
$\{\bq_j\}$ is accomplished by the unitary matrix $\bQ$: the $j$-th coordinate of
$\bx$ in the new basis is $\bq_j^T\bx$. The Fourier transform does exactly the same
thing in $L^2$: the $k$-th coordinate of $f$ in the Fourier basis is $\hat{f}_k =
\langle \phi_k, f\rangle$. Just as $\bQ$ diagonalizes a symmetric matrix
($\bA = \bQ\bD\bQ^T$), the Fourier transform diagonalizes the Laplacian:
$-\widehat{f''(x)}_k = k^2\hat{f}_k$.
}

\addExcr{%
Refer to Def~\ref{def:fourier-series} and Thm~\ref{thm:fourier-orthonormal}.
\begin{enumerate}
  \item Compute the Fourier coefficients $\hat{f}_k$ for $f(x) = x$ on $[0, 2\pi]$
  by evaluating the integral directly. For which $k$ are the coefficients nonzero?
  \item Verify the diagonalization claim: if $f(x) = e^{ikx}$, compute $-f''(x)$
  and show that $-\widehat{f''}_k = k^2\hat{f}_k$ without any integration.
  \item Why does the Fourier transform turn the ODE $-u'' = f$ into the algebraic
  equation $k^2\hat{u}_k = \hat{f}_k$? When is this equation not solvable?
\end{enumerate}
}

\subsubsection{The Discrete Fourier Transform}

\textBox{%
The continuous Fourier series works on functions; the DFT works on finite vectors.
Given a vector $\bx \in \fC^n$ (think: $n$ samples of a signal), the DFT produces
another vector $\hat{\bx} \in \fC^n$ (think: frequency amplitudes). The key fact is
that this transformation is a linear map, so it can be written as a matrix. That matrix
turns out to be unitary, which makes the DFT exactly a change of basis in $\fC^n$.
}

\addDef{DFT Matrix}{%
Let $\omega_n = e^{-2\pi i/n}$ be the primitive $n$-th root of unity. The
\emph{DFT matrix} $\mathbf{F} \in \fC^{n \times n}$ has entries
\[
  F_{jk} = \frac{1}{\sqrt{n}}\,\omega_n^{jk}, \quad j, k = 0, 1, \ldots, n-1.
\]
The DFT of $\bx \in \fC^n$ is $\hat{\bx} = \mathbf{F}\bx$, and its $k$-th entry
\[
  \hat{x}_k = \frac{1}{\sqrt{n}}\sum_{j=0}^{n-1} x_j\,e^{-2\pi i jk/n}
\]
is the discrete analog of the Fourier coefficient at frequency $k$.
}
\label{def:dft-matrix}

\addExmpl{%
For $n = 4$, $\omega_4 = e^{-\pi i/2} = -i$. The DFT matrix is
\[
  \mathbf{F} = \frac{1}{2}
  \begin{pmatrix}
    1 &  1 &  1 &  1 \\
    1 & -i & -1 &  i \\
    1 & -1 &  1 & -1 \\
    1 &  i & -1 & -i
  \end{pmatrix}.
\]
Applying $\mathbf{F}$ to $\bx = (1, 0, 0, 0)^T$ gives $\hat{\bx} = \frac{1}{2}(1, 1, 1, 1)^T$:
a constant in all frequency components, just like a delta spike has equal energy at all
frequencies.
}

\addThm{The DFT Matrix is Unitary}{%
$\mathbf{F}^*\mathbf{F} = \bI$, so $\mathbf{F}$ is a unitary matrix (Def~\ref{def:hermitian}).
Equivalently, the columns of $\mathbf{F}$ form an orthonormal basis of $\fC^n$.
The inverse DFT is $\mathbf{F}^{-1} = \mathbf{F}^*$, with entries
$(\mathbf{F}^*)_{jk} = \frac{1}{\sqrt{n}}\omega_n^{-jk} = \frac{1}{\sqrt{n}}e^{2\pi i jk/n}$.
}
\label{thm:dft-unitary}

\addPrf{%
We compute the $(j, k)$ entry of $\mathbf{F}^*\mathbf{F}$.

\medskip
\textbf{Step 1: Write out the entry.}
\[
  (\mathbf{F}^*\mathbf{F})_{jk}
  = \sum_{m=0}^{n-1} \overline{F_{mj}}\,F_{mk}
  = \frac{1}{n}\sum_{m=0}^{n-1} \omega_n^{-mj}\,\omega_n^{mk}
  = \frac{1}{n}\sum_{m=0}^{n-1} \omega_n^{m(k-j)}.
\]

\textbf{Step 2: Evaluate the geometric sum.}

If $j = k$, each term equals 1, so the sum is $n$ and the entry is $1$.

If $j \neq k$, let $r = \omega_n^{k-j} \neq 1$ (since $k - j \not\equiv 0 \pmod{n}$).
The sum is $\sum_{m=0}^{n-1}r^m = \frac{r^n - 1}{r - 1}$. But $r^n = \omega_n^{n(k-j)} =
e^{-2\pi i(k-j)} = 1$, so the numerator is zero and the sum is $0$.

\textbf{Step 3: Conclude.}

Therefore $(\mathbf{F}^*\mathbf{F})_{jk} = \delta_{jk}$, confirming $\mathbf{F}^*\mathbf{F} = \bI$.
}

\addThm{Parseval's Theorem}{%
For any $\bx \in \fC^n$ with DFT $\hat{\bx} = \mathbf{F}\bx$:
\[
  \|\hat{\bx}\|_2 = \|\bx\|_2.
\]
The DFT preserves the $\ell^2$ norm: total energy is the same in the time domain and the
frequency domain.
}
\label{thm:parseval}

\addPrf{%
Since $\mathbf{F}$ is unitary, it is an isometry (Thm~\ref{thm:orthogonal-isometry}):
\[
  \|\hat{\bx}\|_2^2 = \|\mathbf{F}\bx\|_2^2 = (\mathbf{F}\bx)^*(\mathbf{F}\bx) = \bx^*\mathbf{F}^*\mathbf{F}\bx = \bx^*\bx = \|\bx\|_2^2.
\]
}

\addExcr{%
Refer to Defs~\ref{def:dft-matrix}--\ref{thm:dft-unitary} and Thm~\ref{thm:parseval}.
\begin{enumerate}
  \item For $n = 4$, write out $\mathbf{F}$ explicitly (using $\omega_4 = -i$).
  Verify $\mathbf{F}^*\mathbf{F} = \bI$ by hand for the first two columns.
  \item In NumPy, \texttt{np.fft.fft(x) / np.sqrt(n)} computes the normalized DFT.
  Verify Parseval's theorem numerically for a random $\bx \in \fC^{10}$ by comparing
  \texttt{np.linalg.norm(x)} and \texttt{np.linalg.norm(np.fft.fft(x) / np.sqrt(n))}.
  \item Compute the DFT of the signal $x_j = \cos(2\pi j / n)$ for $n = 16$.
  Which frequency bins $\hat{x}_k$ are nonzero? Explain why.
\end{enumerate}
}

\subsubsection{Circulant Matrices and the Convolution Theorem}

\textBox{%
The DFT is more than just a change of basis for signals: it simultaneously diagonalizes
an entire class of matrices. Circulant matrices arise naturally whenever a system has
translation symmetry (same local rule everywhere), which includes the periodic discrete
Laplacian and discrete convolution operators. Diagonalizing them with the DFT reduces
matrix-vector multiplication from $O(n^2)$ to $O(n \log n)$.
}

\addDef{Circulant Matrix}{%
A matrix $\mathbf{C} \in \fC^{n \times n}$ is \emph{circulant} if each row is a cyclic
right-shift of the previous row. It is completely determined by its first column
$\bc = (c_0, c_1, \ldots, c_{n-1})^T$:
\[
  \mathbf{C} =
  \begin{pmatrix}
    c_0     & c_{n-1} & c_{n-2} & \cdots & c_1 \\
    c_1     & c_0     & c_{n-1} & \cdots & c_2 \\
    c_2     & c_1     & c_0     & \cdots & c_3 \\
    \vdots  &         & \ddots  &        & \vdots \\
    c_{n-1} & c_{n-2} & \cdots  & c_1    & c_0
  \end{pmatrix}.
\]
The product $\mathbf{C}\bx$ computes the \emph{circular convolution} $\bc * \bx$.
}
\label{def:circulant}

\addThm{DFT Diagonalizes Every Circulant Matrix}{%
Every circulant matrix $\mathbf{C}$ with first column $\bc$ satisfies
\[
  \mathbf{C} = \mathbf{F}^*\,\mathrm{diag}(\hat{\bc})\,\mathbf{F},
\]
where $\hat{\bc} = \mathbf{F}\bc$ is the DFT of $\bc$. The eigenvalues of $\mathbf{C}$
are exactly the entries of $\hat{\bc}$ (the DFT coefficients of the first column), and
the eigenvectors are the columns of $\mathbf{F}^*$.
}
\label{thm:circulant-diag}

\addCor{Convolution Theorem}{%
For any $\bx, \by \in \fC^n$, the circular convolution satisfies
\[
  \mathbf{F}(\bx * \by) = (\mathbf{F}\bx) \odot (\mathbf{F}\by),
\]
where $\odot$ denotes entry-wise (Hadamard) product. Circular convolution in the
time domain is pointwise multiplication in the frequency domain.
}
\label{cor:convolution}

\addPrf{%
\textbf{Step 1: Express convolution as a matrix-vector product.}
Circular convolution $\bz = \bx * \by$ means $z_j = \sum_{k=0}^{n-1} x_k y_{(j-k)\bmod n}$.
This is exactly $\bz = \mathbf{C}_{\bx}\by$ where $\mathbf{C}_{\bx}$ is the circulant
matrix with first column $\bx$.

\textbf{Step 2: Diagonalize.}
By Thm~\ref{thm:circulant-diag}, $\mathbf{C}_{\bx} = \mathbf{F}^*\mathrm{diag}(\hat{\bx})\mathbf{F}$, so:
\[
  \bz = \mathbf{C}_{\bx}\by = \mathbf{F}^*\mathrm{diag}(\hat{\bx})\mathbf{F}\by
      = \mathbf{F}^*(\hat{\bx} \odot \hat{\by}).
\]

\textbf{Step 3: Apply $\mathbf{F}$ to both sides.}
\[
  \hat{\bz} = \mathbf{F}\bz = \mathbf{F}\mathbf{F}^*(\hat{\bx} \odot \hat{\by})
  = \hat{\bx} \odot \hat{\by},
\]
using $\mathbf{F}\mathbf{F}^* = \bI$.
}

\addNote{%
\textbf{Application to the Poisson equation.}
The 1-D discrete Laplacian with \emph{periodic} boundary conditions is circulant.
Its eigenvalues are the DFT of the first column $\bc = \frac{1}{h^2}(2, -1, 0, \ldots, 0, -1)^T$.
This gives an $O(n \log n)$ solver for the periodic Poisson equation:
\begin{enumerate}
  \item Compute $\hat{\mathbf{f}} = \mathbf{F}\mathbf{f}$ via FFT.
  \item Divide entry-wise: $\hat{u}_k = \hat{f}_k / \hat{c}_k$.
  \item Invert: $\mathbf{u} = \mathbf{F}^*\hat{\mathbf{u}}$ via inverse FFT.
\end{enumerate}
Compare this to the $O(n^3)$ dense solve or $O(n^{3/2})$ sparse solve.
On a 2-D grid of size $n \times n$ ($N = n^2$ unknowns), the FFT-based solver costs
$O(N \log N)$, making it the method of choice for large periodic problems.
}

\addExcr{%
Refer to Def~\ref{def:circulant} and Thm~\ref{thm:circulant-diag}.
\begin{enumerate}
  \item Build the $5 \times 5$ circulant matrix with first column
  $\bc = (2, -1, 0, 0, -1)^T$ (the periodic 1-D Laplacian for $n = 5$).
  Compute its eigenvalues directly and via $\hat{\bc} = \mathbf{F}\bc$ using
  \texttt{np.fft.fft(c)}. Verify they match.
  \item Implement circular convolution two ways: (a) directly as a matrix-vector
  product using \texttt{scipy.linalg.circulant}, and (b) using \texttt{np.fft.fft}
  and pointwise multiplication. Verify they agree for random $\bx, \by \in \fR^{16}$.
  \item Implement the FFT-based Poisson solver described in the note above for
  $n = 100$ with $f(x) = \sin(2\pi x)$. The exact solution is
  $u(x) = \frac{1}{4\pi^2}\sin(2\pi x)$. Verify your solver recovers it.
\end{enumerate}
}

\subsubsection{The Fast Fourier Transform}

\textBox{%
Computing $\hat{\bx} = \mathbf{F}\bx$ directly costs $O(n^2)$ flops (one row of $\mathbf{F}$
dotted with $\bx$ is $O(n)$, and there are $n$ rows). For $n = 10^6$, this is $10^{12}$
operations: unusable. The Fast Fourier Transform (FFT) reduces this to $O(n \log n)$
by exploiting a recursive structure in the DFT matrix. For $n = 10^6$, that is $2 \times 10^7$
operations: entirely practical. This gap is why FFT appears in audio processing, image
compression (JPEG), wireless communications, and numerical PDE solvers.
}

\addThm{Cooley-Tukey FFT Complexity}{%
When $n = 2^p$ is a power of two, the DFT $\mathbf{F}\bx$ can be computed in
$O(n \log n)$ flops using the Cooley-Tukey algorithm.
}
\label{thm:fft-complexity}

\addPrf{%
The key identity splits the $n$-point DFT into two $(n/2)$-point DFTs.

\medskip
\textbf{Step 1: Split $\bx$ into even and odd indexed entries.}

Let $\bx^{(e)} = (x_0, x_2, \ldots, x_{n-2})$ and $\bx^{(o)} = (x_1, x_3, \ldots, x_{n-1})$.

\textbf{Step 2: Rewrite the DFT sum.}

The $k$-th DFT output is
\[
  \hat{x}_k = \sum_{j=0}^{n-1} x_j \omega_n^{jk}
  = \sum_{j=0}^{n/2-1} x_{2j}\,\omega_n^{2jk} + \omega_n^k\sum_{j=0}^{n/2-1} x_{2j+1}\,\omega_n^{2jk}.
\]

\textbf{Step 3: Recognize the sub-DFTs.}

Since $\omega_n^2 = e^{-4\pi i/n} = \omega_{n/2}$, both sums are $(n/2)$-point DFTs:
\[
  \hat{x}_k = \hat{x}^{(e)}_k + \omega_n^k\,\hat{x}^{(o)}_k.
\]
For $k \geq n/2$, a similar relation holds using $\omega_n^{k + n/2} = -\omega_n^k$.

\textbf{Step 4: Count the cost.}

Let $T(n)$ be the number of flops. The recursion is $T(n) = 2T(n/2) + O(n)$,
which has solution $T(n) = O(n \log n)$.
}

\addNote{%
\textbf{FFT in NumPy.}
\texttt{np.fft.fft(x)} computes the DFT using an FFT algorithm.
\texttt{np.fft.ifft(x)} computes the inverse DFT ($\mathbf{F}^*\bx$).
For real signals, \texttt{np.fft.rfft} exploits the symmetry $\hat{x}_{n-k} = \overline{\hat{x}_k}$
to halve the work. The 2-D FFT (\texttt{np.fft.fft2}) applies the 1-D FFT along each
axis and is the key subroutine for image processing and 2-D PDE solvers.
}

\addExcr{%
Refer to Thm~\ref{thm:fft-complexity}.
\begin{enumerate}
  \item In Python, time \texttt{np.fft.fft(x)} vs.\ a naive DFT implemented as
  \texttt{F @ x} (build $\mathbf{F}$ explicitly) for $n \in \{64, 256, 1024, 4096\}$.
  Plot both times vs.\ $n$ on a log-log scale. At what size does the FFT become
  noticeably faster?
  \item Verify the splitting identity in the proof: for $n = 8$ and a random $\bx$,
  check that $\hat{x}_k = \hat{x}^{(e)}_k + \omega_8^k\hat{x}^{(o)}_k$ for
  $k = 0, 1, 2, 3$ (where $\hat{\bx}^{(e)}, \hat{\bx}^{(o)}$ are 4-point DFTs
  of the even and odd subsequences).
  \item (\textit{Spectral differentiation.})
  Given samples $x_j = f(2\pi j/n)$ for $j = 0, \ldots, n-1$ of a smooth periodic
  function $f$, one can approximate $f'$ by: (a) FFT to get $\hat{\bx}$, (b) multiply
  $\hat{x}_k$ by $ik$, (c) IFFT. Implement this for $f(x) = \sin(x) + \cos(3x)$
  and verify you recover $f'(x) = \cos(x) - 3\sin(3x)$. Compare the accuracy to
  finite differences as $n$ grows.
\end{enumerate}
}

\end{document}
