\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Spectral Graph Theory and PageRank}
\label{sub:spectral_graph_theory}

Encoding graph topology into matrix spectrum for connectivity, clustering, and ranking.

\subsubsection{Graph Matrices}

\addDef{Adjacency and Degree Matrices}{%
For a graph $G = (V, E)$ with $n$ vertices:
\begin{itemize}
    \item \textbf{Adjacency $\bA$:}\enspace $A_{ij} = w_{ij}$ if $(i,j) \in E$, else $0$.
    \item \textbf{Degree $\bD$:}\enspace Diagonal matrix with $D_{ii} = d_i = \sum_{j} w_{ij}$.
\end{itemize}
}

\addDef{Graph Laplacian}{%
Defined as \textbf{$\bL = \bD - \bA$}.
\begin{itemize}
    \item \textbf{Discrete Derivative:}\enspace $(\bL \boldsymbol{f})_i = \sum_{j \sim i} w_{ij} (f_i - f_j)$.
    \item \textbf{Quadratic Form:}\enspace $\bx^T \bL \bx = \sum_{(i,j) \in E} w_{ij} (x_i - x_j)^2$.
\end{itemize}
}

\addThm{Spectral Properties}{%
For a connected graph with $0 = \lambda_1 \leq \lambda_2 \leq \dots \leq \lambda_n$:
\begin{enumerate}
    \item $\lambda_1 = 0$ with eigenvector $\mathbf{1}$.
    \item **Multiplicity of 0:** Equals the number of connected components.
    \item **Fiedler Value ($\lambda_2$):** Measures algebraic connectivity. $\lambda_2 > 0$ iff graph is connected.
\end{enumerate}
}

\addDef{Normalized Laplacians}{%
\begin{itemize}
    \item \textbf{Symmetric:}\enspace $\bL_{\text{sym}} = \bD^{-1/2} \bL \bD^{-1/2}$. (Eigenvalues in $[0, 2]$).
    \item \textbf{Random Walk:}\enspace $\bL_{\text{rw}} = \bD^{-1} \bL = \bI - \bP$.
\end{itemize}
}

\subsubsection{Spectral Clustering}

\addThm{Spectral Relaxation}{%
Minimizing a graph cut (RatioCut) is NP-hard. Spectral methods relax this to an eigenvector problem:
\begin{enumerate}
    \item Compute $k$ smallest eigenvectors of $\bL$ (starting from $\bv_2$).
    \item Embed vertices in $\fR^k$ using these eigenvectors.
    \item Cluster embedding via $k$-means.
\end{enumerate}
}

\subsubsection{Random Walks and PageRank}

\addDef{Random Walk Matrix}{%
Transition matrix $\bP = \bD^{-1} \bA$.
\begin{itemize}
    \item $P_{ij} = w_{ij}/d_j$ is the probability of moving from $j$ to $i$.
    \item \textbf{Stationary Distribution:}\enspace $\boldsymbol{\pi}$ such that $\bP\boldsymbol{\pi} = \boldsymbol{\pi}$. For undirected graphs, $\pi_i = d_i / \sum d_j$.
\end{itemize}
}

\addDef{PageRank and Google Matrix}{%
Models a random surfer on a directed web graph with teleportation:
\[
    \mathbf{G} = d \bS + \frac{1-d}{N} \mathbf{J}
\]
\begin{itemize}
    \item $d$ is the damping factor (standard: 0.85).
    \item $\bS$ is the link matrix (stochastic, handles dangling nodes).
    \item $\mathbf{J}$ is the all-ones matrix (teleportation).
\end{itemize}
\textbf{Computation:}\enspace PageRank vector $\mathbf{r}$ is the dominant eigenvector of $\mathbf{G}$, computed via power iteration.
}

\subsubsection{Exercises}

\addExcr{%
\begin{enumerate}
    \item Construct $\bL$ for a 3-vertex path graph. Find its eigenvalues by hand.
    \item Show that $\bL \mathbf{1} = \bzero$ for any graph.
    \item Prove $\bL$ is positive semi-definite using the quadratic form.
    \item Implement PageRank power iteration for a 10-node random graph.
\end{enumerate}
}

\end{document}
