\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Jacobi Eigenvalue Algorithm}
\label{sub:jacobi-theory}

Iterative algorithm for symmetric matrices using plane rotations to zero off-diagonal entries.

\subsubsection{Givens Rotations}

\addDef{Givens Rotation}{%
$G(p, q, \theta)$ is an identity matrix with a $2 \times 2$ rotation block at $(p, p), (p, q), (q, p), (q, q)$.
\begin{itemize}
    \item \textbf{Similarity Transform:}\enspace $\bA' = G^T \bA G$.
    \item \textbf{Effect:}\enspace Modifies only rows/columns $p$ and $q$. Preserves eigenvalues ($\bA'$ is orthogonally similar to $\bA$).
\end{itemize}
}
\label{def:givens-rotation}

\addThm{Optimal Rotation Angle}{%
To set $A'_{pq} = 0$, let $\tau = (A_{qq} - A_{pp}) / (2A_{pq})$. The optimal tangent $t = \tan\theta$ is:
\[
    t = \frac{\text{sign}(\tau)}{|\tau| + \sqrt{1 + \tau^2}}, \quad c = \frac{1}{\sqrt{1+t^2}}, \quad s = ct.
\]
Choosing the smaller root $|t| \leq 1$ ensures numerical stability and minimal perturbation.
}
\label{thm:jacobi-angle}

\subsubsection{Convergence and Cost}

\addThm{Monotone Convergence}{%
Let $\sigma^2_{\text{off}}(\bA) = \sum_{p < q} A_{pq}^2$. Each rotation reduces this norm:
\[
    \sigma^2_{\text{off}}(\bA') = \sigma^2_{\text{off}}(\bA) - A_{pq}^2.
\]
\textbf{Convergence:}\enspace Off-diagonals decrease monotonically. Convergence is ultimately \textbf{quadratic} (errors square each sweep).
}
\label{thm:jacobi-convergence}

\addNote{%
\textbf{Performance vs.\ Accuracy:}\enspace
\begin{itemize}
    \item \textbf{Cost:}\enspace $O(n^3)$ per sweep. Typically 5--15 sweeps for convergence. More expensive than QR.
    \item \textbf{Accuracy:}\enspace Unlike QR, Jacobi achieves \textbf{high relative accuracy} on small eigenvalues.
\end{itemize}
}

\addExcr{%
\begin{enumerate}
    \item Apply one Jacobi rotation to zero $A_{12}$ in $\bA = \begin{pmatrix}2 & 1 \\ 1 & 3\end{pmatrix}$.
    \item Show that after zeroing $A_{pq}$, the new diagonal entries $A'_{pp}$ and $A'_{qq}$ are the eigenvalues of the $2 \times 2$ submatrix at $\{p, q\}$.
    \item Perform one full cyclic sweep on $\bA = \begin{pmatrix}1&1&0\\1&2&1\\0&1&3\end{pmatrix}$ by hand.
\end{enumerate}
}

\end{document}
