\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Jacobi Iteration}
\label{subsec:jacobi-iteration}

\subsubsection{Jacobi Iteration}

\addDef{Splitting of $\bA$}{%
To derive a stationary iteration from $\bA\bx = \bb$, we decompose $\bA$ as
\[
  \bA = \bD + \bR, \qquad \bD = \mathrm{diag}(a_{11},\ldots,a_{nn}), \quad \bR = \bA - \bD.
\]
Assuming all diagonal entries are nonzero ($a_{ii} \neq 0$ for all $i$), we can rearrange $(\bD + \bR)\bx = \bb$ to obtain the fixed-point equation $\bx = -\bD^{-1}\bR\bx + \bD^{-1}\bb$.
}

\addDef{Jacobi Iteration}{%
The \emph{Jacobi iteration} for $\bA\bx = \bb$ is the stationary scheme
\[
  \bx^{(k+1)} = \bT_J\bx^{(k)} + \bD^{-1}\bb, \qquad \bT_J = -\bD^{-1}\bR,
\]
or in component form,
\[
  x_i^{(k+1)} = \frac{1}{a_{ii}}\Bigl(b_i - \sum_{j \neq i} a_{ij}x_j^{(k)}\Bigr), \quad i = 1,\ldots,n.
\]
Each component $x_i^{(k+1)}$ depends only on values from the \emph{previous} iterate $\bx^{(k)}$, so all $n$ components may be updated simultaneously (or in any order).
}

\newpage
\subsubsection{Convergence of Jacobi Iteration}

\addThm{Convergence of Jacobi Iteration}{%
The Jacobi iteration converges to the solution of $\bA\bx = \bb$ for every initial vector $\bx^{(0)}$ if and only if $\rho(\bT_J) < 1$.
}

\addPrf{%
This is a direct application of the general convergence theorem for stationary iterations: convergence for all $\bx^{(0)}$ is equivalent to $\rho(\bT) < 1$ for the iteration matrix. Here $\bT_J = -\bD^{-1}\bR = \bI - \bD^{-1}\bA$, so the fixed-point equation $\bx^* = \bT_J\bx^* + \bD^{-1}\bb$ simplifies to $\bD^{-1}\bA\bx^* = \bD^{-1}\bb$, i.e., $\bA\bx^* = \bb$.
}

\addThm{Jacobi Convergence for Strictly Diagonally Dominant Matrices}{%
If $\bA$ is strictly diagonally dominant, meaning
\[
  |a_{ii}| > \sum_{j \neq i} |a_{ij}|, \quad \text{for all } i = 1, \ldots, n,
\]
then $\rho(\bT_J) < 1$ and the Jacobi iteration converges for every $\bx^{(0)}$.
}

\addPrf{%
Let $\lambda$ be any eigenvalue of $\bT_J = -\bD^{-1}\bR$ with eigenvector $\bv \neq \bzero$. Let index $i^*$ satisfy $|v_{i^*}| = \|\bv\|_\infty > 0$. The eigenvalue equation $\bT_J\bv = \lambda\bv$ gives, for row $i^*$:
\[
  \lambda v_{i^*} = -\frac{1}{a_{i^*i^*}}\sum_{j \neq i^*} a_{i^*j}\,v_j.
\]
Taking absolute values and dividing by $|v_{i^*}|$:
\[
  |\lambda| \leq \frac{1}{|a_{i^*i^*}|} \sum_{j \neq i^*} |a_{i^*j}| \cdot \frac{|v_j|}{|v_{i^*}|} \leq \frac{\sum_{j \neq i^*} |a_{i^*j}|}{|a_{i^*i^*}|} < 1,
\]
where the last inequality is strict diagonal dominance. Since $\lambda$ was arbitrary, $\rho(\bT_J) < 1$.
}

\addExmpl{%
\textbf{Effect of matrix structure on convergence.}

\textit{Case 1 — Strictly diagonally dominant:}
$\bA_1 = \begin{pmatrix} 5 & 1 & 1 \\ 1 & 5 & 1 \\ 1 & 1 & 5 \end{pmatrix}$.
Each row satisfies $|a_{ii}| = 5 > 2 = \sum_{j\neq i}|a_{ij}|$, so the theorem guarantees convergence. One can compute $\rho(\bT_J) = 2/5 = 0.4$.

\textit{Case 2 — Diagonal dominance violated in one row:}
$\bA_2 = \begin{pmatrix} 2 & 1 & 3 \\ 1 & 3 & 1 \\ 2 & 2 & 2 \end{pmatrix}$.
Row 3 violates strict diagonal dominance. The theorem gives no guarantee; $\rho(\bT_J)$ must be checked numerically.

\textit{Case 3 — Divergent case:}
$\bA_3 = \begin{pmatrix} 1 & 3 & 1 \\ 1 & 2 & 1 \\ 1 & 1 & 2 \end{pmatrix}$.
Here $\rho(\bT_J) > 1$ and the Jacobi iteration diverges for every $\bx^{(0)} \neq \bx^*$.
}

\subsubsection{Stopping Criteria}

\addDef{Residual-Based Stopping Criterion}{%
Stop when the norm of the residual falls below a tolerance:
\[
  \|\bA\bx^{(k)} - \bb\| < \epsilon_{\mathrm{res}}.
\]
This directly measures how well $\bx^{(k)}$ satisfies the linear system.
}

\addDef{Increment-Based Stopping Criterion}{%
Stop when the change between successive iterates is small:
\[
  \|\bx^{(k+1)} - \bx^{(k)}\| < \epsilon_{\mathrm{inc}}.
\]
This is cheap (no matrix–vector product needed), but measures the rate of change of the iteration rather than the accuracy of the solution directly.
}

\addDef{Relative Residual Criterion}{%
Stop when the residual is small relative to the right-hand side:
\[
  \frac{\|\bA\bx^{(k)} - \bb\|}{\|\bb\|} < \epsilon_{\mathrm{rel}}.
\]
This is scale-invariant: multiplying $\bA$ and $\bb$ by a constant does not alter the stopping condition.
}

\addNote{%
A \emph{maximum iteration count} is always imposed as a safeguard against slow convergence or divergence. Typical practice is to set $k_{\max}$ based on the problem size and report non-convergence if the tolerance is not met within the budget.
}

\addThm{A Priori Error Bound for Jacobi Iteration}{%
Let $\rho = \rho(\bT_J) < 1$. For all $k \geq 1$,
\[
  \|\bx^{(k)} - \bx^*\| \leq \frac{\rho^k}{1-\rho}\,\|\bx^{(1)} - \bx^{(0)}\|.
\]
}

\addPrf{%
The basic contraction estimate gives $\|\bx^{(k)} - \bx^*\| \leq \rho^k\|\bx^{(0)} - \bx^*\|$. We bound $\|\bx^{(0)} - \bx^*\|$ via the telescoping sum:
\[
  \|\bx^{(0)} - \bx^*\| \leq \sum_{j=0}^{\infty}\|\bx^{(j+1)} - \bx^{(j)}\|.
\]
Since $\|\bx^{(j+1)} - \bx^{(j)}\| = \|\bT_J^j(\bx^{(1)}-\bx^{(0)})\| \leq \rho^j\|\bx^{(1)}-\bx^{(0)}\|$, the geometric series gives $\|\bx^{(0)} - \bx^*\| \leq \|\bx^{(1)}-\bx^{(0)}\|/(1-\rho)$. Combining yields the stated bound.
}

\addExcr{%
Implement the Jacobi method in Python.
\begin{enumerate}
  \item Write \texttt{jacobi(A, b, x0, tol, max\_iter)} that returns the iterate at convergence (relative residual below \texttt{tol}) and the number of iterations used.
  \item Test on $\bA_1$ from the example with $\bb = (7,7,7)^T$. Compute the exact solution by hand. How many iterations are needed for $\epsilon_{\mathrm{rel}} = 10^{-8}$?
  \item Plot $\|\bA\bx^{(k)} - \bb\|$ versus $k$ on a semi-log scale. What is the observed slope? Compare to $\log_{10}\rho(\bT_J)$.
\end{enumerate}
}

\addExcr{%
\textbf{(A priori vs.\ observed convergence)} For the system with $\bA_1$ and $\bb = (7,7,7)^T$:
\begin{enumerate}
  \item Evaluate the a priori bound for $k = 20$ iterations. How many times larger is the bound than the actual error $\|\bx^{(20)} - \bx^*\|$?
  \item Consider the $5\times5$ tridiagonal matrix with $4$ on the diagonal and $-1$ on the off-diagonals. Verify that it is strictly diagonally dominant, compute $\rho(\bT_J)$ numerically, and determine how many iterations are needed to achieve $\|\bx^{(k)}-\bx^*\| < 10^{-10}$.
\end{enumerate}
}

\end{document}
