\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Lanczos Algorithm}
\label{sub:lanczos-theory}

\textBox{%
Iterative method for finding extreme eigenpairs of large, symmetric matrices without explicit $O(n^2)$ formation.
}

\subsubsection{Krylov Subspaces}

\addDef{Krylov Subspace}{%
$\mathcal{K}_k(\bA, \bv) = \text{span}\{\bv,\, \bA\bv,\, \bA^2\bv,\, \dots,\, \bA^{k-1}\bv\}$.
\textbf{Intuition:}\enspace $\bA^j\bv$ amplifies components along dominant eigenvectors, concentrating information about the spectral edges.
}
\label{def:krylov-subspace}

\subsubsection{The Lanczos Recurrence}

\addThm{Lanczos Relation}{%
The process builds an orthonormal basis $V_k = [\bv_1, \dots, \bv_k]$ for $\mathcal{K}_k$ satisfying:
\begin{align*}
    \bA V_k = V_k T_k + \beta_k \bv_{k+1} \be_k^T.
\end{align*}
\begin{itemize}
    \item \textbf{$T_k$:}\enspace Symmetric \textbf{tridiagonal} matrix.
    \item \textbf{$\alpha_j = \bv_j^T \bA \bv_j$:}\enspace Rayleigh quotients (diagonal).
    \item \textbf{$\beta_j = \|\text{residual}\|$:}\enspace Orthogonalization coefficients (off-diagonal).
\end{itemize}
}
\label{thm:lanczos-relation}

\addNote{%
\textbf{Performance Advantage:}\enspace
Unlike Gram-Schmidt, which orthogonalizes against all previous vectors, Lanczos only requires the \textbf{two previous vectors} (three-term recurrence). This makes it $O(k \cdot \text{nnz})$ for sparse matrices.
}

\subsubsection{Ritz Values and Vectors}

Approximations to the true eigenpairs from the Krylov subspace.

\addDef{Ritz Pair}{%
\begin{itemize}
    \item \textbf{Ritz Value ($\theta_i$):}\enspace Eigenvalue of $T_k$.
    \item \textbf{Ritz Vector ($\tilde{\bu}_i$):}\enspace $V_k \bs_i$, where $\bs_i$ is an eigenvector of $T_k$.
\end{itemize}
}
\label{def:ritz-pair}

\addThm{Residual Estimate}{%
The residual of the $i$-th Ritz pair $(\theta_i, \tilde{\bu}_i)$ satisfies:
\begin{align*}
    \|\bA\tilde{\bu}_i - \theta_i \tilde{\bu}_i\|_2 = \beta_k |s_i(k)|.
\end{align*}
\textbf{Significance:}\enspace Convergence of a Ritz pair can be monitored for $O(k)$ cost without forming the full $n$-vector $\tilde{\bu}_i$.
}
\label{thm:ritz-residual}

\addPrf{%
Starting from the Lanczos relation $\bA V_k = V_k T_k + \beta_k \bv_{k+1} \be_k^T$, multiply by the $i$-th eigenvector $\bs_i$ of $T_k$ on the right:
\begin{align*}
    \bA V_k \bs_i &= V_k T_k \bs_i + \beta_k \bv_{k+1} \be_k^T \bs_i \\
    \bA \tilde{\bu}_i &= V_k (\theta_i \bs_i) + \beta_k \bv_{k+1} s_i(k).
\end{align*}
where $s_i(k)$ is the $k$-th entry of $\bs_i$. Rearranging:
\begin{align*}
    \bA \tilde{\bu}_i - \theta_i \tilde{\bu}_i = \beta_k \bv_{k+1} s_i(k).
\end{align*}
Taking the $L_2$ norm and using the fact that $\|\bv_{k+1}\|_2 = 1$:
\begin{align*}
    \|\bA \tilde{\bu}_i - \theta_i \tilde{\bu}_i\|_2 &= |\beta_k s_i(k)| \|\bv_{k+1}\|_2 \\
    &= \beta_k |s_i(k)|.
\end{align*}
}

\addNote{%
\textbf{Ghost Eigenvalues:}\enspace
In floating-point, the three-term recurrence loses orthogonality, leading to ``ghost'' (spurious) copies of eigenvalues. \textbf{Practical Rule:}\enspace Use full or selective reorthogonalization to maintain basis integrity.
}

\addNote{%
\textbf{Connection to Conjugate Gradient:}\enspace
The Lanczos algorithm applied to a symmetric system $\bA\bx=\bb$ is mathematically equivalent to the Conjugate Gradient method. Both algorithms build the same Krylov subspace and produce identical iterates in exact arithmetic.
}

\addExcr{%
\begin{enumerate}
    \item Carry out 2 steps of Lanczos for $\bA = \text{diag}(5, 3, 1)$ starting with $\bv_1 = \frac{1}{\sqrt{3}}\mathbf{1}$.
    \item Show that $T_k = V_k^T \bA V_k$ (Galerkin Projection).
    \item Implement Lanczos on a $50 \times 50$ Hilbert matrix. Count spurious eigenvalues without reorthogonalization.
\end{enumerate}
}

\end{document}
