\documentclass[../readings.tex]{subfiles}

\begin{document}


\subsection{Linearity}

Linearity is a fundamental concept in numerical linear algebra that underpins many of the algorithms and methods discussed in this course. At its core, linearity describes mathematical structures that behave in a predictable way under addition and scalar multiplication.

\textBox{
A mapping or operation $L$ is considered linear if it satisfies two key properties:
\begin{itemize}
    \item \textbf{Additivity}: $L(\mathbf{x} + \mathbf{y}) = L(\mathbf{x}) + L(\mathbf{y})$ for all inputs $\mathbf{x}$ and $\mathbf{y}$
    \item \textbf{Homogeneity}: $L(\alpha\mathbf{x}) = \alpha L(\mathbf{x})$ for all inputs $\mathbf{x}$ and all scalars $\alpha$
\end{itemize}

These properties can be combined into the principle of superposition: $L(\alpha\mathbf{x} + \beta\mathbf{y}) = \alpha L(\mathbf{x}) + \beta L(\mathbf{y})$ for all inputs $\mathbf{x}$ and $\mathbf{y}$, and all scalars $\alpha$ and $\beta$.
}

\textBox{
Linear mappings preserve vector operations, which allows us to decompose complex problems into simpler ones and then combine the solutions. This property is what makes linear algebra so powerful for a wide range of applications, from solving systems of equations to approximating functions, analyzing data, and modeling physical phenomena.
}

\textBox{
In numerical linear algebra, we often represent linear mappings as matrices. The multiplication of a matrix $\mathbf{A}$ by a vector $\mathbf{x}$ is a linear operation, making matrices the perfect tool for expressing and analyzing linear transformations. The computational challenges that arise in numerical linear algebra—such as solving linear systems, finding eigenvalues, or computing matrix decompositions—all fundamentally involve manipulating these linear structures efficiently and accurately.
}

\subsubsection{Linear Systems}
\addDef{Linear Systems \index{linear systems}}{%
A \textbf{system of linear equations} is a collection of equations in which each equation is a linear combination of the unknown variables. Such a system can be written in the form
\begin{eqnarray*}
    a_{11}x_{1} + a_{12}x_{2} + &\cdots& + a_{1n}x_{n} = b_{1}, \\
    a_{21}x_{1} + a_{22}x_{2} + &\cdots& + a_{2n}x_{n} = b_{2}, \\
    &\vdots& \\
    a_{m1}x_{1} + a_{m2}x_{2} + &\cdots& + a_{mn}x_{n} = b_{m},
\end{eqnarray*}
where each \(a_{ij}\) (with \(1 \le i \le m\) and \(1 \le j \le n\)) is a known constant and \(x_{1},\dots,x_{n}\) are the unknowns. In matrix-vector form, this system is succinctly expressed as
\[
\bA \bx = \bb,
\]
where 
\[
\bA = \begin{pmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{m1} & a_{m2} & \cdots & a_{mn}
\end{pmatrix}, \quad
\bx = \begin{pmatrix}
x_{1} \\
x_{2} \\
\vdots \\
x_{n}
\end{pmatrix}, \quad
\bb = \begin{pmatrix}
b_{1} \\
b_{2} \\
\vdots \\
b_{m}
\end{pmatrix}.
\]
}

\textBox{%
The mapping \(T: \fR^n \to \fR^m\) defined by \(T(\bx) = \bA\bx\) is a linear transformation. That is, for any vectors \(\bx,\by \in \fR^n\) and scalar \(c \in \fR\), we have:
\[
T(\bx + \by) = \bA(\bx+\by) = \bA\bx + \bA\by = T(\bx) + T(\by),
\]
\[
T(c\bx) = \bA(c\bx) = c\,\bA\bx = c\,T(\bx).
\]
These properties, known as additivity and homogeneity, are the defining features of linear maps.
}

\addThm{Matrix-Vector Multiplication is Linear}{%
The map $T: \fR^n \to \fR^m$ defined by $T(\bx) = \bA\bx$ is a linear transformation. That is, for all $\bx, \by \in \fR^n$ and $\alpha \in \fR$:
\[
    T(\bx + \by) = T(\bx) + T(\by), \qquad T(\alpha\bx) = \alpha T(\bx).
\]
Equivalently, $T(\alpha\bx + \beta\by) = \alpha T(\bx) + \beta T(\by)$ for all $\alpha, \beta \in \fR$.
}

\addPrf{%
By distributivity of matrix multiplication: $\bA(\bx + \by) = \bA\bx + \bA\by = T(\bx) + T(\by)$.
By scalar compatibility: $\bA(\alpha\bx) = \alpha(\bA\bx) = \alpha T(\bx)$.
}

\addDef{Existence and Uniqueness \index{linear systems!existence and uniqueness}}{%
For a square system (i.e., when \(m=n\)), the linear system
\[
\bA\bx = \bb
\]
has a unique solution if and only if the coefficient matrix \(\bA\) is invertible, which is equivalent to \(\det(\bA) \neq 0\). If \(\bA\) is singular (i.e., noninvertible, \(\det(\bA)=0\)), then the system either has no solution or has infinitely many solutions. In the latter case, the set of solutions forms an affine subspace of \(\fR^n\).
}

\addThm{Equivalences for Invertibility}{%
Let $\bA \in \fR^{n \times n}$. The following are equivalent:
\begin{enumerate}
    \item $\bA$ is invertible.
    \item $\bA\bx = \bb$ has a unique solution for every $\bb \in \fR^n$.
    \item $\bA\bx = \bzero$ has only the trivial solution $\bx = \bzero$.
    \item $\text{rank}(\bA) = n$.
    \item The columns of $\bA$ are linearly independent.
\end{enumerate}
}

\addPrf{%
$(1)\Rightarrow(2)$: Set $\bx = \bA^{-1}\bb$; then $\bA\bx = \bb$. If $\bA\bx_1 = \bA\bx_2$, then $\bA(\bx_1 - \bx_2) = \bzero$, so $\bx_1 - \bx_2 = \bA^{-1}\bzero = \bzero$.

$(2)\Rightarrow(3)$: Applying (2) with $\bb = \bzero$ forces the unique solution to be $\bx = \bzero$.

$(3)\Rightarrow(4)$: Null space of $\bA$ is $\{\bzero\}$, so nullity$(\bA) = 0$. By the Rank-Nullity Theorem, rank$(\bA) = n$.

$(4)\Rightarrow(1)$: A square matrix with rank $n$ has $n$ linearly independent columns that span $\fR^n$, which is the condition for invertibility.
}

\addDef{Linear Combination}{%
A \textbf{linear combination} of vectors \(\{ \bx_1, \dots, \bx_n \}\) is any vector \(\by\) that can be written as
\[
\by = \alpha_1 \bx_1 + \alpha_2 \bx_2 + \cdots + \alpha_n \bx_n,
\]
where \(\alpha_1, \dots, \alpha_n \in \mathbb{R}\).
}

\addNote{%
View each vector as an ingredient; a linear combination mixes these ingredients in various proportions to produce new vectors.
}

\addDef{Linear Dependence}{%
A set of vectors \(\{ \bx_1, \dots, \bx_n \}\) is \textbf{linearly dependent} if there exist scalars \(\alpha_1, \dots, \alpha_n\), not all zero, such that
\[
\alpha_1 \bx_1 + \alpha_2 \bx_2 + \cdots + \alpha_n \bx_n = \mathbf{0}.
\]
If the only solution is \(\alpha_1 = \cdots = \alpha_n = 0\), the vectors are \textbf{linearly independent}.
}

\addNote{%
Linear dependence indicates redundancy; one or more vectors in the set can be expressed as a combination of the others.
}


\addDef{Basis}{%
A \textbf{basis} for a vector space \(\mathcal{V}\) is a set of vectors that is both linearly independent and spanning. Every vector in \(\mathcal{V}\) can be uniquely expressed as a linear combination of the basis vectors. The number of basis vectors is the \textbf{dimension} of \(\mathcal{V}\).
}



\subsubsection{Rank}

\addDef{Rank}{%
The \textbf{rank} of a matrix \(\bA\) is defined as the maximum number of linearly independent rows (or equivalently, columns) of \(\bA\). Equivalently, it is the dimension of the row space (or column space) of \(\bA\). For an \(m \times n\) matrix, the rank is always less than or equal to \(\min(m,n)\).
}

\addNote{%
Intuitively, the rank measures the amount of “information” contained in the matrix. It indicates the maximum number of independent directions or components that the matrix can represent. For instance, if a matrix has a rank of \(r\), then there are \(r\) rows (or columns) that contribute unique information, while the remaining rows (or columns) can be expressed as linear combinations of these.
}

\addNote{%
For an \(m \times n\) matrix, since there are \(m\) rows and \(n\) columns, the rank cannot exceed either \(m\) or \(n\). In particular, if \(\bA\) is square (i.e., \(m=n\)) and full rank (i.e., rank = \(n\)), then \(\bA\) is invertible.
}

\addDef{Full Rank}{%
A matrix is said to be \textbf{full rank} if its rank is as large as possible, that is, if \(\text{rank}(\bA) = \min(m,n)\). In the case of a square matrix, full rank implies invertibility.
}

\addExmpl{%
\textbf{Example: Determining the Rank of a Matrix}\\[2mm]
Consider the matrix
\[
\bA = \begin{pmatrix}
1 & 2 & 3 \\
2 & 4 & 6 \\
1 & 1 & 1
\end{pmatrix}.
\]
Observe that the second row is exactly twice the first row, suggesting a dependency. To determine the rank, we perform elementary row operations to obtain the row-echelon form:

\textbf{Step 1:} Leave the first row unchanged:
\[
R_1 = (1,2,3).
\]

\textbf{Step 2:} Replace the second row by subtracting \(2R_1\):
\[
R_2 \rightarrow R_2 - 2R_1: \quad (2,4,6) - 2(1,2,3) = (0,0,0).
\]

\textbf{Step 3:} Replace the third row by subtracting \(R_1\):
\[
R_3 \rightarrow R_3 - R_1: \quad (1,1,1) - (1,2,3) = (0,-1,-2).
\]

The matrix becomes:
\[
\bA' = \begin{pmatrix}
1 & 2 & 3 \\
0 & 0 & 0 \\
0 & -1 & -2
\end{pmatrix}.
\]
Swap \(R_2\) and \(R_3\) to bring the nonzero row up:
\[
\bA'' = \begin{pmatrix}
1 & 2 & 3 \\
0 & -1 & -2 \\
0 & 0 & 0
\end{pmatrix}.
\]
Finally, multiply the second row by \(-1\):
\[
\bA''' = \begin{pmatrix}
1 & 2 & 3 \\
0 & 1 & 2 \\
0 & 0 & 0
\end{pmatrix}.
\]

Since there are two nonzero rows (pivot rows), the rank of \(\bA\) is \(2\).
}

\addDef{Rank-Nullity Theorem}{%
For any \(m \times n\) matrix \(\bA\),
\[
\text{rank}(\bA) + \text{nullity}(\bA) = n,
\]
where the nullity of \(\bA\) is the dimension of its null space. This relationship is fundamental in understanding the solution structure of the homogeneous system \(\bA\bx = \mathbf{0}\).
}

\addThm{Rank-Nullity Theorem}{%
For any $\bA \in \fR^{m \times n}$,
\[
    \text{rank}(\bA) + \text{nullity}(\bA) = n,
\]
where $\text{nullity}(\bA) = \dim(\ker(\bA))$ and $\ker(\bA) = \{\bx \in \fR^n : \bA\bx = \bzero\}$.
}

\addNote{%
For a square $n \times n$ system $\bA\bx = \bb$, the Rank-Nullity Theorem gives a complete picture:
\begin{itemize}
    \item $\text{rank}(\bA) = n$: the null space is $\{\bzero\}$, so $\bA\bx = \bb$ has a unique solution.
    \item $\text{rank}(\bA) = r < n$: if a solution exists, there are infinitely many — the solution set is an affine subspace of dimension $n - r$.
\end{itemize}
In NumPy, \texttt{np.linalg.matrix\_rank(A)} computes the rank numerically (via SVD).
}

\addNote{%
In many applications such as data compression, signal processing, and machine learning, the rank of a matrix is used to identify redundancy in data. Low-rank approximations can help reduce dimensionality while preserving essential information.
}

\subsubsection{Subspaces and the Four Fundamental Subspaces}

\addDef{Subspace}{%
A nonempty subset $\mathcal{S} \subseteq \fR^n$ is a \emph{subspace} if it is closed under addition and scalar multiplication:
\begin{enumerate}
  \item $\bzero \in \mathcal{S}$,
  \item $\bu, \bv \in \mathcal{S} \Rightarrow \bu + \bv \in \mathcal{S}$,
  \item $\bu \in \mathcal{S},\, \alpha \in \fR \Rightarrow \alpha\bu \in \mathcal{S}$.
\end{enumerate}
Equivalently, $\mathcal{S}$ is a subspace if and only if it is closed under all linear combinations: $\alpha\bu + \beta\bv \in \mathcal{S}$ for all $\bu, \bv \in \mathcal{S}$ and $\alpha, \beta \in \fR$.
}

\addNote{%
Every subspace of $\fR^n$ must contain $\bzero$ (set $\alpha = 0$ in condition 3). Conversely, a set that does not contain $\bzero$ cannot be a subspace. Geometrically, subspaces are flat (no curvature) and always pass through the origin — lines and planes through the origin in $\fR^3$ are subspaces; lines and planes that do not pass through the origin are \emph{affine subspaces} (translates of subspaces).
}

\addDef{The Four Fundamental Subspaces}{%
For any matrix $\bA \in \fR^{m \times n}$ of rank $r$, four subspaces arise naturally:
\begin{enumerate}
  \item \textbf{Column space} (range): $\mathcal{R}(\bA) = \{\bA\bx : \bx \in \fR^n\} \subseteq \fR^m$, \quad $\dim = r$.
  \item \textbf{Null space} (kernel): $\mathcal{N}(\bA) = \{\bx \in \fR^n : \bA\bx = \bzero\} \subseteq \fR^n$, \quad $\dim = n - r$.
  \item \textbf{Row space}: $\mathcal{R}(\bA^T) = \{\bA^T\by : \by \in \fR^m\} \subseteq \fR^n$, \quad $\dim = r$.
  \item \textbf{Left null space}: $\mathcal{N}(\bA^T) = \{\by \in \fR^m : \bA^T\by = \bzero\} \subseteq \fR^m$, \quad $\dim = m - r$.
\end{enumerate}
The dimensions are given by the Rank-Nullity Theorem: $r + (n-r) = n$ and $r + (m-r) = m$.
}

\addThm{Fundamental Theorem of Linear Algebra}{%
The four fundamental subspaces of $\bA \in \fR^{m \times n}$ satisfy two orthogonality relations:
\[
  \mathcal{R}(\bA^T) \perp \mathcal{N}(\bA) \quad \text{in } \fR^n, \qquad \mathcal{R}(\bA) \perp \mathcal{N}(\bA^T) \quad \text{in } \fR^m.
\]
Moreover, these pairs are \emph{orthogonal complements}: $\mathcal{R}(\bA^T) \oplus \mathcal{N}(\bA) = \fR^n$ and $\mathcal{R}(\bA) \oplus \mathcal{N}(\bA^T) = \fR^m$.
}

\addPrf{%
For the first relation: if $\bv \in \mathcal{R}(\bA^T)$, then $\bv = \bA^T\by$ for some $\by$. For any $\bx \in \mathcal{N}(\bA)$, $\bA\bx = \bzero$, so
\[
  \bv^T\bx = (\bA^T\by)^T\bx = \by^T(\bA\bx) = \by^T\bzero = 0.
\]
Hence every row space vector is orthogonal to every null space vector. The complementarity (their dimensions sum to $n$, and they have only $\bzero$ in common) follows from the Rank-Nullity Theorem.
The second relation follows by applying the first to $\bA^T$.
}

\addNote{%
The orthogonal complement picture has an immediate payoff for solving $\bA\bx = \bb$:
\begin{itemize}
  \item \textbf{Solvability}: $\bA\bx = \bb$ has a solution if and only if $\bb \in \mathcal{R}(\bA)$, equivalently $\bb \perp \mathcal{N}(\bA^T)$ (i.e., $\by^T\bb = 0$ for all $\by$ with $\bA^T\by = \bzero$).
  \item \textbf{Solution structure}: If $\hat{\bx}$ is any particular solution, the full solution set is $\hat{\bx} + \mathcal{N}(\bA)$ — a translate of the null space.
  \item \textbf{Uniqueness}: The solution is unique iff $\mathcal{N}(\bA) = \{\bzero\}$, i.e., $\bA$ has full column rank.
\end{itemize}
}

\addExmpl{%
Let $\bA = \begin{pmatrix}1 & 2 & 3 \\ 2 & 4 & 6\end{pmatrix}$. Row 2 is twice row 1, so $\text{rank}(\bA) = 1$.
\begin{itemize}
  \item $\mathcal{R}(\bA) = \text{span}\{(1,2)^T\} \subset \fR^2$ — a line through the origin (dimension 1).
  \item $\mathcal{N}(\bA^T) = \text{span}\{(2,-1)^T\} \subset \fR^2$ — its orthogonal complement (dimension 1). Check: $(1,2)\cdot(2,-1)=0$. $\checkmark$
  \item $\mathcal{R}(\bA^T) = \text{span}\{(1,2,3)^T\} \subset \fR^3$ — a line (dimension 1).
  \item $\mathcal{N}(\bA) \subset \fR^3$ — a plane (dimension 2), spanned by $(-2,1,0)^T$ and $(-3,0,1)^T$. Every vector in this plane is orthogonal to $(1,2,3)^T$.
\end{itemize}
The system $\bA\bx = \bb$ has a solution only when $\bb = \alpha(1,2)^T$ for some scalar $\alpha$.
}

\textBox{%
These four subspaces reappear in every major topic of these notes. In least squares (\S\ref{subsec:least-squares}), the residual $\bb - \bA\hat{\bx}$ lands in $\mathcal{N}(\bA^T)$ — orthogonal to $\mathcal{R}(\bA)$. In SVD, the right singular vectors split into a basis for $\mathcal{R}(\bA^T)$ and $\mathcal{N}(\bA)$; the left singular vectors split into $\mathcal{R}(\bA)$ and $\mathcal{N}(\bA^T)$. In eigenvalue problems, the concept generalizes to \emph{invariant subspaces} — subspaces that are mapped into themselves under $\bA$.
}

\addDef{Linear Span}{%
The \textbf{linear span} of a set of vectors \(\{ \bx_1, \dots, \bx_n \}\), denoted \(\text{span}\{\bx_1, \dots, \bx_n\}\), is the set of all linear combinations of these vectors:
\[
\text{span}\{\bx_1, \dots, \bx_n\} = \{ \alpha_1 \bx_1 + \cdots + \alpha_n \bx_n : \alpha_i \in \mathbb{R} \}.
\]
This is the smallest subspace of the vector space that contains the given vectors.
}

\addExmpl{%
\textbf{ Basis for \(\text{span}\{(1,2), (3,4)\}\) in \(\mathbb{R}^2\)}\\[2mm]
Consider the vectors \(\bv_1 = (1,2)\) and \(\bv_2 = (3,4)\). To check their linear independence, assume there exist scalars \(\alpha\) and \(\beta\) such that
\[
\alpha (1,2) + \beta (3,4) = (0,0).
\]
This yields the system:
\[
\alpha + 3\beta = 0, \quad 2\alpha + 4\beta = 0.
\]
Multiplying the first equation by 2 gives \(2\alpha + 6\beta = 0\). Subtracting the second equation, we have
\[
(2\alpha + 6\beta) - (2\alpha + 4\beta) = 2\beta = 0 \quad \Rightarrow \quad \beta = 0.
\]
Substituting back, \(\alpha = 0\). Since the only solution is \(\alpha=\beta=0\), the vectors are linearly independent. Hence, \(\{(1,2), (3,4)\}\) forms a basis for \(\text{span}\{(1,2), (3,4)\}\), which in this case is all of \(\mathbb{R}^2\).
}

\addExcr{%
Consider the system $\bA\bx = \bb$ with
\[
    \bA = \begin{pmatrix} 1 & 2 & -1 \\ 2 & 5 & 0 \\ -1 & 1 & 6 \end{pmatrix}, \quad \bb = \begin{pmatrix} 1 \\ 4 \\ 3 \end{pmatrix}.
\]
\begin{enumerate}
    \item Determine the rank of $\bA$ by row-reducing the augmented matrix $[\bA \mid \bb]$.
    \item How many solutions does this system have? Justify using the Rank-Nullity Theorem.
    \item Find the solution(s).
    \item In NumPy, solve using \texttt{np.linalg.solve(A, b)} and check the residual \texttt{np.linalg.norm(b - A @ x)}. What residual do you expect for a consistent system?
\end{enumerate}
}

\addExcr{%
\textbf{(Linear Independence and the Null Space)} Let $\bA \in \fR^{m \times n}$ with $m < n$ (a wide matrix).
\begin{enumerate}
    \item Argue that the columns of $\bA$ must be linearly dependent. (\textit{Hint: use Rank-Nullity with $\text{rank}(\bA) \leq m < n$.})
    \item What does this imply about the existence and uniqueness of solutions to $\bA\bx = \bb$?
    \item Give a $2 \times 3$ example where $\bA\bx = \bb$ has infinitely many solutions, and find the general solution as a particular solution plus the null space.
    \item Can a wide system $\bA\bx = \bb$ ever have no solution? Explain.
\end{enumerate}
}

\addExcr{%
\textbf{(Superposition Principle)} Suppose $\bx_1$ solves $\bA\bx = \bb_1$ and $\bx_2$ solves $\bA\bx = \bb_2$.
\begin{enumerate}
    \item Show that $\alpha\bx_1 + \beta\bx_2$ solves $\bA\bx = \alpha\bb_1 + \beta\bb_2$.
    \item Use this principle to solve $\bA\bx = 3\bb_1 - 2\bb_2$ without any new computation.
    \item Why is this principle computationally valuable when solving $\bA\bX = \bB$ for a matrix right-hand side $\bB$? How does \texttt{np.linalg.solve(A, B)} exploit this?
\end{enumerate}
}

\end{document}
