\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{LU Factorization}
\label{sub:lu-factorization}

\textBox{%
The LU factorization expresses a square matrix \(\bA \in \mathbb{R}^{n \times n}\) as the product 
\[
\bA = \bL\,\bU,
\]
where \(\bL\) is a lower triangular matrix with ones on the diagonal (a unit lower triangular matrix) and \(\bU\) is an upper triangular matrix. This factorization simplifies solving linear systems \(\bA\bx=\bb\) by allowing the solution to be obtained via forward substitution (for \(\bL\by=\bb\)) followed by backward substitution (for \(\bU\bx=\by\)).
}

\subsubsection{LU Factoriazation and Gaussian Elimination}
\addDef{LU Factorization}{%
An \textbf{LU Factorization} of a matrix \(\bA \in \mathbb{R}^{n \times n}\) is a decomposition of the form
\[
\bA = \bL\,\bU,
\]
with
\[
\bL = \begin{pmatrix}
1 & 0 & \cdots & 0\\[1mm]
\ell_{21} & 1 & \cdots & 0\\[1mm]
\vdots & \vdots & \ddots & \vdots\\[1mm]
\ell_{n1} & \ell_{n2} & \cdots & 1
\end{pmatrix} \quad \text{and} \quad
\bU = \begin{pmatrix}
u_{11} & u_{12} & \cdots & u_{1n}\\[1mm]
0 & u_{22} & \cdots & u_{2n}\\[1mm]
\vdots & \vdots & \ddots & \vdots\\[1mm]
0 & 0 & \cdots & u_{nn}
\end{pmatrix}.
\]
}

\addDef{Gaussian elimination}{%
The procedure for LU factorization closely follows Gaussian elimination. For example, given
\[
\bA = \begin{pmatrix}
a_{11} & a_{12} & \cdots & a_{1n}\\[1mm]
a_{21} & a_{22} & \cdots & a_{2n}\\[1mm]
\vdots & \vdots & \ddots & \vdots\\[1mm]
a_{n1} & a_{n2} & \cdots & a_{nn}
\end{pmatrix},
\]
one begins by eliminating the entries below \(a_{11}\). For \(i=2,\dots,n\), set
\[
\ell_{i1} = \frac{a_{i1}}{a_{11}},
\]
and update row \(i\) by replacing it with
\[
\text{row}_i \leftarrow \text{row}_i - \ell_{i1}\,\text{row}_1.
\]
This results in zeros below \(a_{11}\) and new entries
\[
a_{ij}^{(1)} = a_{ij} - \ell_{i1}\,a_{1j}, \quad j=2,\dots,n.
\]
Repeating this process on the \((n-1)\times(n-1)\) submatrix produces the full LU decomposition.
}

\addThm{LU Factorization Existence and Uniqueness}{%
A matrix $\bA \in \fR^{n \times n}$ has a unique LU factorization (without pivoting) if and only if all leading principal submatrices $\bA_k \in \fR^{k \times k}$ (the top-left $k \times k$ blocks) are nonsingular for $k = 1, \ldots, n-1$.
}

\addPrf{%
At step $k$ of Gaussian elimination, the pivot is the $(k,k)$ entry of the current reduced matrix, which equals $\det(\bA_k)/\det(\bA_{k-1})$ (an inductive result). This pivot is nonzero iff $\det(\bA_k) \neq 0$. If all $n-1$ pivots are nonzero, elimination completes and produces a unique $\bL, \bU$ since each multiplier $\ell_{ik} = a_{ik}/a_{kk}$ and each updated entry are determined uniquely.
}

\addThm{LU Factorization Cost}{%
Gaussian elimination to compute the LU factorization of $\bA \in \fR^{n \times n}$ costs
\[
    \frac{2}{3}n^3 + O(n^2) \text{ flops.}
\]
}

\addPrf{%
At elimination step $k$ ($k = 1, \ldots, n-1$), we compute $n-k$ multipliers ($n-k$ divisions) and update an $(n-k)\times(n-k)$ submatrix ($2(n-k)^2$ flops for multiply-add pairs). The dominant cost is:
\[
    \sum_{k=1}^{n-1} 2(n-k)^2 = 2\sum_{j=1}^{n-1} j^2 = 2\cdot\frac{(n-1)n(2n-1)}{6} = \frac{2n^3}{3} + O(n^2).
\]
}

\addNote{%
The $\frac{2}{3}n^3$ constant is exactly twice the $\frac{1}{3}n^3$ cost for a triangular solve. So the factorization dominates: it costs $O(n^3)$ once, but each solve thereafter is only $O(n^2)$.
}

\addExmpl{%
To illustrate, consider the \(3 \times 3\) matrix
\[
\bA = \begin{pmatrix}
2 & 3 & 1 \\
4 & 7 & 3 \\
-2 & -3 & 1
\end{pmatrix}.
\]
We wish to decompose \(\bA\) as \(\bA = \bL\,\bU\) with
\[
\bL = \begin{pmatrix}
1 & 0 & 0 \\
\ell_{21} & 1 & 0 \\
\ell_{31} & \ell_{32} & 1
\end{pmatrix}, \quad
\bU = \begin{pmatrix}
u_{11} & u_{12} & u_{13} \\
0 & u_{22} & u_{23} \\
0 & 0 & u_{33}
\end{pmatrix}.
\]
Begin by taking the first row of \(\bA\) as the first row of \(\bU\):
\[
u_{11} = 2,\quad u_{12} = 3,\quad u_{13} = 1.
\]
The multipliers for the first column are then
\[
\ell_{21} = \frac{4}{2} = 2,\quad \ell_{31} = \frac{-2}{2} = -1.
\]
After eliminating the first column, update the remaining rows. For row 2:
\[
u_{22} = 7 - 2\cdot 3 = 1,\quad u_{23} = 3 - 2\cdot 1 = 1.
\]
For row 3, the updated entries become
\[
\tilde{a}_{32} = -3 - (-1)\cdot 3 = 0,\quad \tilde{a}_{33} = 1 - (-1)\cdot 1 = 2.
\]
Thus, \(\ell_{32} = 0\) and \(u_{33} = 2\). Consequently, we have
\[
\bL = \begin{pmatrix}
1 & 0 & 0 \\
2 & 1 & 0 \\
-1 & 0 & 1
\end{pmatrix},\quad
\bU = \begin{pmatrix}
2 & 3 & 1 \\
0 & 1 & 1 \\
0 & 0 & 2
\end{pmatrix}.
\]
A direct multiplication verifies that \(\bL\,\bU = \bA\).
}

\addExcr{%
Compute the LU factorization (without pivoting) of
\[
    \bA = \begin{pmatrix} 1 & 2 & 1 \\ 3 & 8 & 1 \\ 0 & 4 & 1 \end{pmatrix}.
\]
\begin{enumerate}
    \item Carry out the elimination steps explicitly: compute multipliers $\ell_{21}, \ell_{31}$ at step 1, update the submatrix, then compute $\ell_{32}$ at step 2.
    \item Write $\bL$ and $\bU$ explicitly. Verify $\bL\bU = \bA$.
    \item Use the factorization to solve $\bA\bx = (5, 15, 4)^T$ via forward then backward substitution.
    \item In Python, verify with \texttt{from scipy.linalg import lu} and \texttt{lu(A)}. Why does SciPy always return a permutation $\bP$ even if no row swaps are needed?
\end{enumerate}
}

\subsubsection{Pivoting and LUP Decomposition}

\addDef{Pivoting and LUP Decomposition}{%
When the matrix \(\bA\) has zeros or small entries in prospective pivot positions, it is necessary to perform row interchanges to maintain numerical stability. These interchanges are recorded by a permutation matrix \(\mathbf{P}\), leading to the factorization
\[
\mathbf{P}\bA = \bL\,\bU,
\]
which is known as the \emph{LUP Decomposition}. Here, \(\bL\) is a unit lower triangular matrix and \(\bU\) is an upper triangular matrix.
}

\textBox{%
Partial pivoting selects the largest available pivot (in absolute value) by interchanging (permuting) rows. This minimizes the risk of division by a small number and reduces the growth of rounding errors. Although complete pivoting, which interchanges both rows and columns, can further improve stability, it is generally more computationally expensive.
}

\textBox{%
The permutation matrix \(\mathbf{P}\) is orthogonal in the sense that \(\mathbf{P}^T\mathbf{P} = \bI\). In fact, for any permutation matrix \(\mathbf{P}\), one has \(\mathbf{P}^{-1} = \mathbf{P}^T\), reflecting that \(\mathbf{P}\) merely rearranges the rows of the identity matrix.
}

\addExmpl{%
For example, consider the matrix
\[
\bA = \begin{pmatrix}
0 & 2 & 1 \\
2 & 1 & 0 \\
1 & 0 & 2
\end{pmatrix}.
\]
Since the (1,1) entry is zero, we must permute the rows to secure a nonzero (and preferably large) pivot in the (1,1) position. Observing the first column, the largest absolute value is \(2\) (in row 2). Therefore, we swap row 1 with row 2 using the permutation matrix
\[
\mathbf{P} = \begin{pmatrix}
0 & 1 & 0 \\
1 & 0 & 0 \\
0 & 0 & 1
\end{pmatrix}.
\]
Then,
\[
\mathbf{P}\bA = \begin{pmatrix}
2 & 1 & 0 \\
0 & 2 & 1 \\
1 & 0 & 2
\end{pmatrix}.
\]
}

\textBox{%
Next, we perform the LU factorization on \(\mathbf{P}\bA\). Write
\[
\mathbf{P}\bA = \bL\,\bU,
\]
with
\[
\bL = \begin{pmatrix}
1 & 0 & 0 \\
\ell_{21} & 1 & 0 \\
\ell_{31} & \ell_{32} & 1
\end{pmatrix} \quad \text{and} \quad
\bU = \begin{pmatrix}
u_{11} & u_{12} & u_{13} \\
0 & u_{22} & u_{23} \\
0 & 0 & u_{33}
\end{pmatrix}.
\]
Assign the first row of \(\bU\) from \(\mathbf{P}\bA\):
\[
u_{11} = 2,\quad u_{12} = 1,\quad u_{13} = 0.
\]
Then, the multipliers for the first column are
\[
\ell_{21} = \frac{0}{2} = 0,\quad \ell_{31} = \frac{1}{2} = 0.5.
\]
Subtracting \(0.5\) times the first row from the third row updates it to
\[
\begin{pmatrix} 1 & 0 & 2 \end{pmatrix} - 0.5\,\begin{pmatrix} 2 & 1 & 0 \end{pmatrix} = \begin{pmatrix} 0 & -0.5 & 2 \end{pmatrix}.
\]
For the \(2\times2\) submatrix (rows 2 and 3), set
\[
u_{22} = 2,\quad u_{23} = 1,
\]
and compute the multiplier
\[
\ell_{32} = \frac{-0.5}{2} = -0.25.
\]
Finally, update the (3,3) entry:
\[
u_{33} = 2 - (-0.25)(1) = 2.25.
\]
Thus, we obtain
\[
\bL = \begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0.5 & -0.25 & 1
\end{pmatrix},\quad
\bU = \begin{pmatrix}
2 & 1 & 0 \\
0 & 2 & 1 \\
0 & 0 & 2.25
\end{pmatrix}.
\]
A multiplication of \(\bL\) and \(\bU\) confirms that \(\bL\,\bU = \mathbf{P}\bA\).
}

\subsubsection{Solving Linear Systems using LU Factorization}

\textBox{%
Once a square matrix \(\bA \in \fR^{n \times n}\) has been factorized as \(\bA = \bL\bU\), solving the linear system \(\bA\bx = \bb\) reduces to two simpler steps: forward substitution and backward substitution.
}

\addDef{LU Solve}{%
Given the LU factorization, we solve the system \(\bA\bx = \bL \bU \bx =\bb\) by first solving
\[
\bL\by = \bb,
\]
and then solving
\[
\bU\bx = \by.
\]}

\textBox{
Since \(\bL\) is a unit lower triangular matrix, we use forward substitution. Specifically, for \(i=1,2,\dots,n\),
\[
y_i = b_i - \sum_{j=1}^{i-1} \ell_{ij} \, y_j.
\]
After obtaining \(\by\), we solve the upper triangular system using backward substitution. For \(i=n, n-1, \dots, 1\),
\[
x_i = \frac{1}{u_{ii}} \left( y_i - \sum_{j=i+1}^{n} u_{ij} \, x_j \right).
\]
}

\addNote{%
The forward substitution process requires roughly
\[
\sum_{i=1}^{n} (i-1) = \frac{n(n-1)}{2} = O(n^2)
\]
operations, and backward substitution has a similar computational cost. Thus, once the LU factorization is available, solving \(\bA\bx = \bb\) takes \(O(n^2)\) operations.
}

\addNote{%
It is important to note that while solving the system after the factorization is efficient, computing the LU factorization itself generally requires \(O(n^3)\) operations. Therefore, if multiple right-hand sides \(\bb\) are to be solved with the same matrix \(\bA\), it is advantageous to factorize \(\bA\) once and reuse the factors \(\bL\) and \(\bU\) for each subsequent system (very common in iterative methods), reducing the per-system cost to \(O(n^2)\).
}

\addThm{Total Cost of LU-Based Solve}{%
Solving $k$ linear systems $\bA\bx_i = \bb_i$ ($i = 1, \ldots, k$) with the same $\bA \in \fR^{n \times n}$ via LU factorization costs
\[
    \frac{2}{3}n^3 + 2kn^2 + O(n) \text{ flops,}
\]
versus $k \cdot \left(\frac{2}{3}n^3 + O(n^2)\right)$ if the matrix is re-factorized for each solve.
}

\addExcr{%
\begin{enumerate}
    \item Derive the exact flop count for forward substitution (computing $\by$ from $\bL\by = \bb$): count the multiplications and additions at each step $i = 1, \ldots, n$. Show the total is $n^2 - n$ arithmetic operations.
    \item Suppose $\bA$ is $1000 \times 1000$ and you need to solve for 500 right-hand sides. How many flops does the LU-then-solve approach save compared to calling \texttt{np.linalg.solve} 500 times? (Assume each call re-factorizes.)
    \item Give a concrete $2 \times 2$ matrix where Gaussian elimination without pivoting fails (division by zero) but LUP succeeds. Write out $\bP$, $\bL$, $\bU$.
    \item In Python, use \texttt{scipy.linalg.lu\_factor(A)} once and \texttt{scipy.linalg.lu\_solve} for each right-hand side. Write the code for 3 right-hand sides and verify each residual is near machine precision.
\end{enumerate}
}

\addExcr{%
\textbf{(Stability of Gaussian Elimination)} Partial pivoting selects the largest pivot at each step.
\begin{enumerate}
    \item The \textbf{growth factor} $\rho$ of LU with partial pivoting is $\rho = \max_{i,j,k}|u_{ij}^{(k)}| / \max_{i,j}|a_{ij}|$, where $u_{ij}^{(k)}$ are the entries at elimination step $k$. For partial pivoting, $\rho \leq 2^{n-1}$. What is this for $n = 50$?
    \item In practice $\rho$ is almost always $O(n)$ or smaller. Does this make LU with partial pivoting \emph{backward stable} in theory or just in practice?
    \item For a matrix where partial pivoting performs poorly (the theoretical worst case), complete pivoting—interchanging both rows and columns—is guaranteed to have $\rho = O(n^{2/3})$ but is rarely used. What is the computational reason?
\end{enumerate}
}

\end{document}
