\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Mathematical Notation and Proof Techniques}
\label{sub:math-notation}

These notes assume familiarity with calculus and basic proof writing, but the conventions
below establish a shared vocabulary used throughout. Skimming this subsection now will
prevent confusion when proofs appear later.

\subsubsection{Logic and Set Notation}

\addDef{Logical Quantifiers}{%
The \emph{universal quantifier} $\forall$ means ``for all''; the \emph{existential quantifier}
$\exists$ means ``there exists''. For example, ``$\forall \varepsilon > 0,\;\exists \delta > 0$
such that\ldots'' reads as ``for every positive $\varepsilon$ there is a positive $\delta$
such that\ldots''. The negation of $\forall x,\, P(x)$ is $\exists x,\, \neg P(x)$, and vice versa.
}
\label{def:quantifiers}

\addDef{Implication and Equivalence}{%
$P \Rightarrow Q$ (``$P$ implies $Q$'') means: whenever $P$ is true, $Q$ is also true.
$P \Leftrightarrow Q$ (``$P$ if and only if $Q$'', abbreviated \emph{iff}) means
$P \Rightarrow Q$ and $Q \Rightarrow P$.
The \emph{contrapositive} of $P \Rightarrow Q$ is $\neg Q \Rightarrow \neg P$, which is
logically equivalent and often easier to prove.
}
\label{def:implication}

\addDef{Set Notation}{%
A \emph{set} is a collection of distinct objects. Common notation:
\begin{itemize}
  \item $x \in S$: $x$ is an element of $S$; \quad $x \notin S$: $x$ is not.
  \item $S \subseteq T$: $S$ is a subset of $T$ (every element of $S$ is in $T$).
  \item $S \cup T$: union; \quad $S \cap T$: intersection; \quad $S \setminus T$: set difference.
  \item $\{x \in S : P(x)\}$: the set of elements in $S$ satisfying property $P$.
\end{itemize}
Common number sets used throughout these notes: $\fZ$ (integers), $\fR$ (reals),
$\fC$ (complex numbers), $\fR^n$ (real $n$-vectors), $\fR^{m \times n}$
(real $m \times n$ matrices).
}
\label{def:set-notation}

\addExcr{%
\begin{enumerate}
  \item Write the statement ``every invertible matrix has a nonzero determinant'' using
  quantifiers and implication.
  \item State the contrapositive of: ``if $\bA$ is invertible, then $\bA\bx = \bzero$
  implies $\bx = \bzero$.''
  \item Let $S = \{1, 2, 3, 4\}$ and $T = \{3, 4, 5, 6\}$. Compute $S \cup T$,
  $S \cap T$, and $S \setminus T$.
\end{enumerate}
}

\subsubsection{Proof Techniques}

\textBox{%
Nearly every claim in these notes is accompanied by a proof. There are three main
strategies used here; recognizing which applies is the first step in writing your own proofs.
}

\addDef{Direct Proof}{%
A \emph{direct proof} of $P \Rightarrow Q$ assumes $P$ is true and derives $Q$ through a
chain of valid logical steps. Most proofs in these notes are direct.
}
\label{def:direct-proof}

\addDef{Proof by Contradiction}{%
To prove statement $P$: assume $\neg P$ and derive a logical contradiction
($R$ and $\neg R$ simultaneously, for some statement $R$). This shows $\neg P$ is
impossible, so $P$ must hold.
}
\label{def:proof-contradiction}

\addDef{Proof by Induction}{%
To prove $P(n)$ holds for all integers $n \geq n_0$:
\begin{enumerate}
  \item \emph{Base case}: verify $P(n_0)$ directly.
  \item \emph{Inductive step}: assume $P(k)$ (the \emph{inductive hypothesis}) and prove $P(k+1)$.
\end{enumerate}
Mathematical induction appears whenever an algorithm or formula is defined recursively.
}
\label{def:proof-induction}

\addExcr{%
\begin{enumerate}
  \item Use a direct proof to show: if $\bA$ and $\bB$ are symmetric
  ($\bA^T = \bA$, $\bB^T = \bB$), then $\bA + \bB$ is symmetric.
  \item Use proof by contradiction to show: if $\bA\bx = \bzero$ implies $\bx = \bzero$,
  then $\bA$ has no zero eigenvalue.
  \item By induction, prove $\displaystyle\sum_{k=1}^{n} k = \frac{n(n+1)}{2}$.
\end{enumerate}
}

\subsubsection{Notation Conventions}

\textBox{%
The following conventions are used consistently throughout these notes.
}

\addNote{%
\textbf{Scalars, Vectors, and Matrices.}\enspace
Scalars are lowercase italic: $a, b, c, x$.
Vectors are lowercase bold: $\bx, \by, \ba$.
Matrices are uppercase bold: $\bA, \bB, \bQ$.
The $i$-th entry of $\bx$ is $x_i$; the $(i,j)$ entry of $\bA$ is $a_{ij}$ or $(\bA)_{ij}$.
Indexing is 1-based throughout (unlike Python).
}

\addNote{%
\textbf{Transpose and Inverse.}\enspace
$\bA^T$ is the transpose of $\bA$; $\bA^{-1}$ is the matrix inverse (exists only for
square, invertible $\bA$). The notation $\bA^{-T}$ means $(\bA^{-1})^T = (\bA^T)^{-1}$.
$\bA^k$ denotes $\bA$ multiplied by itself $k$ times; $\bA^0 = \bI$.
}

\addNote{%
\textbf{Sums and Products.}\enspace
\[
  \sum_{i=1}^{n} a_i = a_1 + a_2 + \cdots + a_n, \qquad
  \prod_{i=1}^{n} a_i = a_1 a_2 \cdots a_n.
\]
When the index set is clear from context we write $\sum_i a_i$.
}

\addNote{%
\textbf{Functions.}\enspace
$f : \fR^n \to \fR^m$ means $f$ maps $n$-vectors to $m$-vectors.
$f \circ g$ denotes function composition: $(f \circ g)(\bx) = f(g(\bx))$.
A function $f : \fR^{m \times n} \to \fR$ maps matrices to scalars (e.g., a norm or the determinant).
}

\end{document}
