\documentclass[../readings.tex]{subfiles}

\begin{document}
\subsection{Vector and Matrix Norms}
\label{sub:matrix-norms}

\subsubsection{Vector Norms}

\textBox{%
Before defining norms on matrices, we establish the vector norm foundations on which they rest. A vector norm assigns a non-negative scalar ``length'' to every vector, and the key families each capture a different geometric intuition.
}

\addDef{Vector Norm}{%
A function $\|\cdot\| : \fR^n \to \fR$ is a \emph{vector norm} if for all $\bx, \by \in \fR^n$ and $c \in \fR$:
\begin{enumerate}
  \item $\|\bx\| \geq 0$ and $\|\bx\| = 0 \Leftrightarrow \bx = \bzero$ \quad (positive definiteness)
  \item $\|c\bx\| = |c|\|\bx\|$ \quad (absolute homogeneity)
  \item $\|\bx + \by\| \leq \|\bx\| + \|\by\|$ \quad (triangle inequality)
\end{enumerate}
}

\addDef{$\ell^p$ Norms}{%
For $\bx \in \fR^n$ and $p \geq 1$, the \emph{$\ell^p$ norm} is
\[
  \|\bx\|_p = \left(\sum_{i=1}^n |x_i|^p\right)^{1/p}.
\]
The three most common cases are:
\begin{itemize}
  \item $\|\bx\|_1 = \sum_{i=1}^n |x_i|$ \quad (\emph{Manhattan norm} — sum of absolute values)
  \item $\|\bx\|_2 = \sqrt{\sum_{i=1}^n x_i^2} = \sqrt{\bx^T\bx}$ \quad (\emph{Euclidean norm})
  \item $\|\bx\|_\infty = \max_{1\leq i\leq n}|x_i|$ \quad (\emph{max norm} — limit as $p \to \infty$)
\end{itemize}
}

\addThm{Cauchy-Schwarz Inequality}{%
For any $\bx, \by \in \fR^n$:
\[
  |\bx^T\by| \leq \|\bx\|_2\,\|\by\|_2,
\]
with equality if and only if $\bx$ and $\by$ are linearly dependent.
}

\addPrf{%
If either vector is zero the inequality holds trivially. Otherwise, for any $t \in \fR$:
\[
  0 \leq \|\bx - t\by\|_2^2 = \|\bx\|_2^2 - 2t\,\bx^T\by + t^2\|\by\|_2^2.
\]
This is a quadratic in $t$ that is always non-negative, so its discriminant must satisfy $4(\bx^T\by)^2 - 4\|\bx\|_2^2\|\by\|_2^2 \leq 0$, giving $|\bx^T\by| \leq \|\bx\|_2\|\by\|_2$. Equality holds iff $\bx = t\by$ for the optimal $t = \bx^T\by/\|\by\|_2^2$.
}

\addLem{Young's Inequality}{%
For $a, b \geq 0$ and conjugate exponents $p, q \geq 1$ with $\frac{1}{p}+\frac{1}{q}=1$,
\[
  ab \leq \frac{a^p}{p} + \frac{b^q}{q}.
\]
Equality holds if and only if $a^p = b^q$.
}

\addPrf{%
If $a = 0$ or $b = 0$ the inequality is immediate. Otherwise, the function $f(t) = t/p - t^{1/p}/p + 1/q$ satisfies $f(1) = 0$ and $f'(t) = (1 - t^{1/p-1})/p$, so $f$ has a global minimum at $t = 1$ with $f \geq 0$. Setting $t = a^p/b^q$ and multiplying through by $b^q$ gives the result.

Alternatively: the weighted AM-GM inequality $\alpha u + \beta v \geq u^\alpha v^\beta$ (for $\alpha + \beta = 1$, $u,v > 0$) applied with $\alpha = 1/p$, $\beta = 1/q$, $u = a^p$, $v = b^q$ gives $a^p/p + b^q/q \geq (a^p)^{1/p}(b^q)^{1/q} = ab$.
}

\addThm{Hölder's Inequality}{%
For $p, q \geq 1$ with $\frac{1}{p} + \frac{1}{q} = 1$ (conjugate exponents), and any $\bx, \by \in \fR^n$:
\[
  |\bx^T\by| \leq \|\bx\|_p\,\|\by\|_q.
\]
Cauchy-Schwarz is the special case $p = q = 2$.
}

\addPrf{%
If $\bx = \bzero$ or $\by = \bzero$ the result is immediate. Otherwise, set $a_i = |x_i|/\|\bx\|_p$ and $b_i = |y_i|/\|\by\|_q$. By Young's inequality applied to each $i$:
\[
  a_i b_i \leq \frac{a_i^p}{p} + \frac{b_i^q}{q}.
\]
Summing over $i$ and using $\sum a_i^p = 1$, $\sum b_i^q = 1$ gives $\sum a_i b_i \leq 1/p + 1/q = 1$. Multiplying by $\|\bx\|_p\|\by\|_q$ yields $|\bx^T\by| \leq \sum|x_i||y_i| \leq \|\bx\|_p\|\by\|_q$.
}

\addThm{Equivalence of Vector Norms}{%
All norms on $\fR^n$ are equivalent: for any two norms $\|\cdot\|_\alpha$ and $\|\cdot\|_\beta$, there exist constants $c_1, c_2 > 0$ such that
\[
  c_1\|\bx\|_\beta \leq \|\bx\|_\alpha \leq c_2\|\bx\|_\beta \quad \text{for all } \bx \in \fR^n.
\]
In particular, for $\bx \in \fR^n$:
\[
  \|\bx\|_2 \leq \|\bx\|_1 \leq \sqrt{n}\,\|\bx\|_2, \qquad \|\bx\|_\infty \leq \|\bx\|_2 \leq \sqrt{n}\,\|\bx\|_\infty.
\]
}

\addPrf{%
Existence of $c_1, c_2$ follows from the compactness of the unit sphere $\{\bx : \|\bx\|_\beta = 1\}$ in $\fR^n$: any norm is a continuous function, so it attains its bounds on a compact set. The explicit inequalities follow from Cauchy-Schwarz: $\|\bx\|_1 = \sum|x_i| \cdot 1 \leq \|\bx\|_2\|\mathbf{1}\|_2 = \sqrt{n}\|\bx\|_2$, and $\|\bx\|_2^2 = \sum x_i^2 \leq \|\bx\|_\infty \sum |x_i| = \|\bx\|_\infty\|\bx\|_1 \leq n\|\bx\|_\infty^2$.
}

\addNote{%
Norm equivalence means convergence in one norm implies convergence in all norms. In practice, the choice of norm affects constants but not qualitative behavior. The $\ell^2$ norm is preferred for geometric intuition; the $\ell^\infty$ norm is often used in error analysis; the $\ell^1$ norm appears in sparse recovery and optimization.
}

\addExcr{%
\begin{enumerate}
  \item For $\bx = (3, -4, 0)^T$, compute $\|\bx\|_1$, $\|\bx\|_2$, $\|\bx\|_\infty$. Verify all three equivalence inequalities hold.
  \item Prove the triangle inequality for $\|\cdot\|_2$ using Cauchy-Schwarz: expand $\|\bx+\by\|_2^2 = (\bx+\by)^T(\bx+\by)$.
  \item In NumPy, \texttt{np.linalg.norm(x, ord=p)} computes $\ell^p$ norms. For a random $\bx \in \fR^{10}$, verify numerically that $\|\bx\|_\infty \leq \|\bx\|_2 \leq \sqrt{10}\|\bx\|_\infty$.
  \item Show that $\|\bx\|_\infty = \lim_{p\to\infty}\|\bx\|_p$. (\textit{Hint: factor out the maximum entry.})
\end{enumerate}
}

\subsubsection{Matrix Norms}

\textBox{%
Matrix norms assign a scalar ``size'' to a matrix. They are fundamental in numerical analysis: they measure approximation errors in low-rank truncation, bound how much a matrix amplifies vectors, and define the condition number. Every norm inequality we use in convergence analysis ultimately rests on a handful of key theorems proved here.
}

\addDef{Matrix Norm}{%
A function $\|\cdot\| : \fR^{m\times n} \to \fR$ is a \emph{matrix norm} if for all $\bA, \bB \in \fR^{m\times n}$ and all $c \in \fR$:
(i)~$\|\bA\| \geq 0$ and $\|\bA\|=0 \Leftrightarrow \bA = \mathbf{0}$ \quad (non-negativity and definiteness);
(ii)~$\|c\bA\| = |c|\|\bA\|$ \quad (absolute homogeneity);
(iii)~$\|\bA+\bB\| \leq \|\bA\| + \|\bB\|$ \quad (triangle inequality).
A square-matrix norm is additionally \emph{submultiplicative} if $\|\bA\bB\| \leq \|\bA\|\|\bB\|$.
}

\addDef{Induced (Operator) Norm}{%
Given vector norms $\|\cdot\|_u$ on $\fR^m$ and $\|\cdot\|_v$ on $\fR^n$, the \emph{induced} (or \emph{operator}) norm of $\bA \in \fR^{m\times n}$ is
\[
  \|\bA\|_{u,v} = \max_{\bx \neq \bzero} \frac{\|\bA\bx\|_u}{\|\bx\|_v} = \max_{\|\bx\|_v=1}\|\bA\bx\|_u.
\]
It measures the maximum stretching factor applied by $\bA$.
}

\addThm{Induced Norms are Submultiplicative}{%
Every induced norm satisfies $\|\bA\bB\| \leq \|\bA\|\|\bB\|$.
}

\addPrf{%
For any $\bx \neq \bzero$,
\[
  \|\bA\bB\bx\| \leq \|\bA\|\,\|\bB\bx\| \leq \|\bA\|\,\|\bB\|\,\|\bx\|.
\]
Dividing by $\|\bx\|$ and taking the supremum over all $\bx \neq \bzero$ gives $\|\bA\bB\| \leq \|\bA\|\|\bB\|$.
}

\addDef{$L_1$ Norm (Maximum Absolute Column Sum)}{%
The matrix $L_1$ norm, induced by the vector $\ell^1$ norm, is
\[
  \|\bA\|_1 = \max_{1 \leq j \leq n} \sum_{i=1}^m |a_{ij}|.
\]
It equals the maximum absolute column sum of $\bA$.
}

\addDef{$L_\infty$-Norm (Maximum Absolute Row Sum)}{%
The matrix $L_\infty$ norm, induced by the vector $\ell^\infty$ norm, is
\[
  \|\bA\|_\infty = \max_{1 \leq i \leq m} \sum_{j=1}^n |a_{ij}|.
\]
It equals the maximum absolute row sum of $\bA$.
}

\addDef{Spectral Norm}{%
The matrix $L_2$ norm (spectral norm), induced by the vector $\ell^2$ norm, is
\[
  \|\bA\|_2 = \sigma_{\max}(\bA) = \sqrt{\lambda_{\max}(\bA^T\bA)}.
\]
It equals the largest singular value and measures the maximum Euclidean stretching factor of $\bA$.
}

\addThm{Spectral Norm via Singular Values}{%
$\|\bA\|_2 = \sigma_{\max}(\bA)$.
}

\addPrf{%
Write the SVD as $\bA = \bU\bsigma\bV^T$. For $\|\bx\|_2 = 1$, let $\by = \bV^T\bx$ (which also satisfies $\|\by\|_2 = 1$, since $\bV$ is orthogonal). Then
\[
  \|\bA\bx\|_2 = \|\bU\bsigma\bV^T\bx\|_2 = \|\bsigma\by\|_2 \leq \sigma_1\|\by\|_2 = \sigma_1,
\]
with equality when $\by = \mathbf{e}_1$, i.e., $\bx = \bv_1$ (the first right singular vector). Hence $\|\bA\|_2 = \sigma_1 = \sigma_{\max}(\bA)$.
}

\addDef{Frobenius Norm}{%
The \emph{Frobenius norm} treats $\bA$ as a vector in $\fR^{mn}$:
\[
  \|\bA\|_F = \sqrt{\sum_{i=1}^m \sum_{j=1}^n a_{ij}^2}.
\]
}

\addThm{Frobenius Norm via Trace and Singular Values}{%
$\|\bA\|_F = \sqrt{\mathrm{Tr}(\bA^T\bA)} = \sqrt{\mathrm{Tr}(\bA\bA^T)} = \sqrt{\sum_{k} \sigma_k^2}$.
}

\addPrf{%
By definition, $\|\bA\|_F^2 = \sum_{i,j} a_{ij}^2 = \sum_j (\bA^T\bA)_{jj} = \mathrm{Tr}(\bA^T\bA)$. The matrix $\bA^T\bA$ is symmetric positive semidefinite with eigenvalues $\sigma_1^2 \geq \cdots \geq \sigma_n^2 \geq 0$ (the squared singular values of $\bA$), so $\mathrm{Tr}(\bA^T\bA) = \sum_k \sigma_k^2$. The identity $\mathrm{Tr}(\bA^T\bA) = \mathrm{Tr}(\bA\bA^T)$ follows since both equal $\sum_{i,j} a_{ij}^2$.
}

\addNote{%
The Frobenius norm is submultiplicative and is consistent with the vector $\ell^2$ norm: $\|\bA\bx\|_2 \leq \|\bA\|_F\|\bx\|_2$. It is \emph{not} induced by any vector norm, but satisfies $\|\bA\|_2 \leq \|\bA\|_F \leq \sqrt{r}\,\|\bA\|_2$ where $r = \mathrm{rank}(\bA)$.
}

\addDef{Max Norm}{%
The \emph{max norm} is $\|\bA\|_{\max} = \max_{i,j}|a_{ij}|$. It is a valid matrix norm but is \emph{not} submultiplicative in general ($\|\bA\bB\|_{\max}$ can exceed $\|\bA\|_{\max}\|\bB\|_{\max}$) and is not an induced norm.
}

\addThm{Norm Equivalence in Finite Dimensions}{%
All matrix norms on $\fR^{m\times n}$ are equivalent: for any two norms $\|\cdot\|_\alpha$ and $\|\cdot\|_\beta$, there exist constants $c_1, c_2 > 0$ such that $c_1\|\bA\|_\beta \leq \|\bA\|_\alpha \leq c_2\|\bA\|_\beta$ for all $\bA$.
}

\addPrf{%
Since $\fR^{m\times n}$ is a finite-dimensional vector space (of dimension $mn$), any norm on it is equivalent to the standard Euclidean norm on $\fR^{mn}$ (which equals the Frobenius norm). Specifically, the unit ball of any norm is compact, so the continuous function $\bA \mapsto \|\bA\|_\alpha / \|\bA\|_\beta$ attains its supremum and infimum on the unit sphere $\{\bA : \|\bA\|_\beta = 1\}$, yielding the constants $c_1$ and $c_2$.
}

\addProp{Consistency with Vector Norms}{%
An induced norm $\|\bA\|$ satisfies $\|\bA\bx\|_v \leq \|\bA\|\|\bx\|_v$ for all $\bx$ by definition. The Frobenius norm also satisfies $\|\bA\bx\|_2 \leq \|\bA\|_F\|\bx\|_2$.
}

\addPrf{%
For the Frobenius norm, use Cauchy-Schwarz: $(\bA\bx)_i = \sum_j a_{ij}x_j$, so $|(\bA\bx)_i|^2 \leq \sum_j a_{ij}^2 \cdot \|\bx\|_2^2$. Summing over $i$ gives $\|\bA\bx\|_2^2 \leq \|\bA\|_F^2\|\bx\|_2^2$.
}

\addNote{%
Matrix norms appear in the condition number $\kappa(\bA) = \|\bA\|\|\bA^{-1}\|$. For the 2-norm this gives $\kappa_2(\bA) = \sigma_{\max}/\sigma_{\min}$. For the Frobenius norm, $\|\bA\|_F$ prominently features in the Eckart-Young-Mirsky theorem for optimal low-rank approximation.
}

\addExmpl{%
Let $\bA = \begin{pmatrix} 1 & -2 \\ -3 & 4 \end{pmatrix}$.
\begin{itemize}
  \item $\|\bA\|_1 = \max(1+3,\, 2+4) = 6$ (maximum column sum).
  \item $\|\bA\|_\infty = \max(1+2,\, 3+4) = 7$ (maximum row sum).
  \item $\|\bA\|_F = \sqrt{1+4+9+16} = \sqrt{30} \approx 5.48$.
  \item $\|\bA\|_2 = \sigma_{\max}$: eigenvalues of $\bA^T\bA$ satisfy $\lambda^2 - 30\lambda + 4 = 0$, giving $\lambda_{\max} = 15 + \sqrt{221}$, so $\|\bA\|_2 \approx 5.46$.
\end{itemize}
Note $\|\bA\|_2 \leq \|\bA\|_F$ as guaranteed by the theorem.
}

\addExcr{%
\begin{enumerate}
  \item Prove that $\|\bA\|_2 \leq \|\bA\|_F \leq \sqrt{r}\,\|\bA\|_2$ where $r = \mathrm{rank}(\bA)$. \textit{Hint: use the singular value representation of both norms.}
  \item For $n\times n$ matrices, prove $\|\bA\|_2 \leq \sqrt{n}\|\bA\|_\infty$ and $\|\bA\|_\infty \leq \sqrt{n}\|\bA\|_2$.
  \item Using NumPy, verify these inequalities for 1000 random $5\times 5$ matrices drawn from \texttt{np.random.randn}. Plot histograms of the ratios $\|\bA\|_F/\|\bA\|_2$ and $\|\bA\|_\infty/\|\bA\|_2$.
\end{enumerate}
}

\addExcr{%
\textbf{(Max norm is not submultiplicative)} Find a pair of $2\times 2$ matrices $\bA, \bB$ such that $\|\bA\bB\|_{\max} > \|\bA\|_{\max}\|\bB\|_{\max}$. Conclude that the max norm cannot be an induced norm.
}

\end{document}
