\documentclass[../readings.tex]{subfiles}

\begin{document}
\subsection{Numerical Methods for ODEs}
\label{subsec:ode-methods}

\textBox{%
Solve the \emph{initial value problem} (IVP): find $y : [t_0, T] \to \fR^d$ satisfying
\[
\frac{dy}{dt} = f(t, y), \qquad y(t_0) = y_0.
\]
Numerical methods advance the solution step by step using a fixed or adaptive step size $h$.
The fundamental trade-offs are \emph{accuracy} (local truncation error per step),
\emph{stability} (whether errors grow or decay), and \emph{cost} (function evaluations per step).
}

\subsubsection{Euler's Method}

\addDef{Forward Euler}{%
The simplest one-step method replaces $dy/dt$ by a forward difference:
\[
y_{n+1} = y_n + h f(t_n, y_n).
\]
}

\addThm{Euler Method Error}{%
The \emph{local truncation error} (error per step) is $O(h^2)$: for smooth $y$,
\[
y(t_{n+1}) - y_{n+1} = \frac{h^2}{2} y''(t_n) + O(h^3).
\]
The \emph{global error} over $[t_0, T]$ is $O(h)$: halving $h$ halves the total error.
}

\addPrf{%
Taylor-expand the exact solution: $y(t_{n+1}) = y(t_n) + hy'(t_n) + \frac{h^2}{2}y''(t_n) + O(h^3)$.
The Euler step uses $y'(t_n) = f(t_n, y_n)$, so the local error is $\frac{h^2}{2}y''(t_n) + O(h^3)$.
Over $N = (T - t_0)/h$ steps, the global error is bounded by $N \cdot O(h^2) = O(h)$.
}

\subsubsection{Classical Runge-Kutta (RK4)}

\addDef{Classical Fourth-Order Runge-Kutta}{%
The \emph{RK4} method evaluates $f$ at four intermediate stages:
\begin{align}
k_1 &= f(t_n,\, y_n), \qquad k_2 = f\!\left(t_n + \tfrac{h}{2},\, y_n + \tfrac{h}{2}k_1\right), \\
k_3 &= f\!\left(t_n + \tfrac{h}{2},\, y_n + \tfrac{h}{2}k_2\right), \qquad k_4 = f(t_n + h,\, y_n + hk_3), \\
y_{n+1} &= y_n + \frac{h}{6}(k_1 + 2k_2 + 2k_3 + k_4).
\end{align}
}

\addThm{RK4 Accuracy}{%
The local truncation error of RK4 is $O(h^5)$ and the global error is $O(h^4)$. The four
stages are chosen so that the Taylor expansion of the numerical update matches $y(t_{n+1})$
through terms of order $h^4$. Halving $h$ reduces the error by a factor of 16.
}

\subsubsection{Adaptive Methods: RK45 / Dormand-Prince}

\textBox{%
Fixed step-size methods waste computation in smooth regions and are inaccurate in rapidly
varying regions. \emph{Adaptive} methods adjust $h$ to maintain a target local error tolerance.
}

\addDef{Embedded Runge-Kutta Methods}{%
An \emph{embedded} pair uses the same stage evaluations $k_1, \ldots, k_s$ to produce two
approximations of different orders:
\[
y_{n+1}^{(4)} = y_n + \sum_{i} b_i k_i, \qquad y_{n+1}^{(5)} = y_n + \sum_{i} \hat{b}_i k_i.
\]
The difference $e_{n+1} = y_{n+1}^{(5)} - y_{n+1}^{(4)}$ estimates the local error without
extra function evaluations. The step is accepted if $\|e_{n+1}\| \leq \varepsilon_{\mathrm{tol}}$
and rejected (with $h$ reduced) otherwise.
}

\addNote{%
The \emph{Dormand-Prince} method (DOPRI5), used by \texttt{scipy.integrate.solve\_ivp} with
\texttt{method='RK45'}, uses 6 stages to produce 4th and 5th order estimates with the
standard step-size controller
\[
h_{\mathrm{new}} = h_{\mathrm{old}} \cdot \min\!\left(h_{\max},\, \max\!\left(h_{\min},\,
\mathrm{safety} \cdot \left(\frac{\varepsilon_{\mathrm{tol}}}{\|e\|}\right)^{1/5}\right)\right).
\]
}

\subsubsection{Symplectic Integrators}

\textBox{%
For \emph{Hamiltonian systems} $\dot{q} = \partial H/\partial p$, $\dot{p} = -\partial H/\partial q$,
standard Runge-Kutta methods do not conserve the energy $H(q, p)$ over long integrations ---
the numerical energy drifts. Symplectic integrators preserve the geometric structure of the flow.
}

\addDef{Velocity Verlet}{%
For a separable Hamiltonian $H = \frac{1}{2m}p^2 + V(q)$ (so $\dot{q} = p/m$,
$\dot{p} = -\nabla V(q)$), the \emph{velocity Verlet} integrator is
\[
q_{n+1} = q_n + h v_n + \frac{h^2}{2m} F_n, \qquad v_{n+1} = v_n + \frac{h}{2m}(F_n + F_{n+1}),
\]
where $F_n = -\nabla V(q_n)$. This requires one new force evaluation per step.
}

\addThm{Symplectic Structure of Velocity Verlet}{%
Velocity Verlet is a \emph{symplectic} integrator of order 2: it exactly preserves the
symplectic 2-form $dq \wedge dp$. As a consequence, there exists a \emph{modified Hamiltonian}
$\tilde{H}$ that is exactly conserved by the discrete flow. The numerical energy $H(q_n, p_n)$
oscillates but does not drift --- the error remains bounded for arbitrarily long times.
}

\addNote{%
By contrast, Euler's method applied to a Hamiltonian system has monotonically \emph{growing}
energy (it is anti-symplectic), and RK4 also exhibits slow energy drift over very long
integrations.
}

\addDef{Yoshida 4th-Order Symplectic Integrator}{%
The \emph{Yoshida} method composes three Verlet steps with carefully chosen sub-step sizes
to achieve 4th-order accuracy while remaining symplectic:
\[
w_1 = \frac{1}{2 - 2^{1/3}}, \qquad w_0 = 1 - 2w_1,
\]
\[
\Phi_h = \Phi_{w_1 h}^{\mathrm{Verlet}} \circ \Phi_{w_0 h}^{\mathrm{Verlet}} \circ \Phi_{w_1 h}^{\mathrm{Verlet}}.
\]
This requires 3 force evaluations per step and is 4th-order accurate and symplectic.
}

\subsubsection{Exercises}

\addExcr{%
\begin{enumerate}
    \item Apply 4 steps of forward Euler with $h = 0.25$ to solve $y' = -y$, $y(0) = 1$
    on $[0, 1]$. Compare to $y(t) = e^{-t}$. What is the global error?
    \item Verify that RK4 applied to $y' = \lambda y$ with step $h$ gives
    $y_{n+1} = R(h\lambda) y_n$ where $R(z) = 1 + z + z^2/2 + z^3/6 + z^4/24$.
    For which $h\lambda \in \fR$ is $|R(h\lambda)| \leq 1$ (stability)?
    \item Show that forward Euler applied to the harmonic oscillator $\dot{q} = p$,
    $\dot{p} = -q$ gives $H(q_{n+1}, p_{n+1}) = (1 + h^2) H(q_n, p_n)$. Conclude the
    energy grows without bound. What does velocity Verlet give instead?
    \item Derive the velocity Verlet update by applying the Störmer-Verlet splitting:
    decompose the Hamiltonian flow as $\Phi_h \approx \Phi_{h/2}^V \circ \Phi_h^T \circ \Phi_{h/2}^V$
    where $\Phi_h^T$ advances positions (kinetic step) and $\Phi_h^V$ advances momenta
    (potential step).
    \item (Adaptive step control) Implement RK45 for $y' = \cos(t^2)$, $y(0) = 0$ on
    $[0, 10]$ with relative tolerance $10^{-6}$. Plot the step sizes. Where does the
    integrator take small steps? Explain in terms of $|f'(t)|$.
\end{enumerate}
}

\end{document}
