\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Orthogonal Matrices}

Orthogonal matrices represent the class of linear transformations that preserve lengths and
angles --- rotations and reflections. They are numerically ideal: multiplying by an orthogonal
matrix introduces no error amplification ($\kappa_2(\bQ) = 1$). This section develops
orthogonal matrices and their role in the QR factorization, which underpins least-squares
solvers and the QR eigenvalue algorithm.

\addDef{Orthogonal Vectors}{%
Two vectors $\bx, \by \in \fR^n$ are called \emph{orthogonal} if their inner product is
zero:
\[
\bx^T\by = 0.
\]
Geometrically, orthogonal vectors are perpendicular. A set of vectors
$\{\bv_1, \bv_2, \ldots, \bv_k\}$ is \emph{mutually orthogonal} if every pair of distinct
vectors is orthogonal:
\[
\bv_i^T\bv_j = 0 \quad \text{for all } i \neq j.
\]
}
\label{def:orthogonal-vectors}

\addExcr{%
Refer to Def~\ref{def:orthogonal-vectors}.
\begin{enumerate}
    \item Are $(1, 2)^T$ and $(4, -2)^T$ orthogonal? Are $(1, 1)^T$ and $(1, -1)^T$
    orthogonal?
    \item If $\bx \perp \by$ and $\bx \perp \bz$, show that
    $\bx \perp (\alpha\by + \beta\bz)$ for any scalars $\alpha, \beta$.
    \item In $\fR^3$, find a nonzero vector orthogonal to both $(1, 0, 1)^T$ and
    $(0, 1, 1)^T$. (\textit{Hint: solve a homogeneous system or use the cross product.})
\end{enumerate}
}

\addDef{Orthonormal Basis}{%
A set of vectors $\{\bq_1, \bq_2, \ldots, \bq_n\}$ in $\fR^n$ is an \emph{orthonormal
basis} if the vectors are mutually orthogonal and each has unit length:
\[
\bq_i^T\bq_j = \delta_{ij} =
\begin{cases}
1 & \text{if } i = j \\
0 & \text{if } i \neq j.
\end{cases}
\]
Any $n$ such vectors in $\fR^n$ are automatically linearly independent and span $\fR^n$.
If we arrange them as columns of $\bQ = [\bq_1, \ldots, \bq_n]$, then $\bQ^T\bQ = \bI$.
}
\label{def:orthonormal-basis}

\addNote{%
Any vector $\bv \in \fR^n$ can be expressed uniquely as a linear combination of an
orthonormal basis $\{\bq_1, \ldots, \bq_n\}$:
\[
\bv = \sum_{i=1}^n (\bq_i^T\bv)\,\bq_i.
\]
The coefficients $\bq_i^T\bv$ are simple inner products --- no linear system needs to be
solved. This is the key computational advantage of orthonormal bases.
}

\addExcr{%
Refer to Def~\ref{def:orthonormal-basis}.
\begin{enumerate}
    \item Verify that $\bq_1 = (1/\sqrt{2},\,1/\sqrt{2})^T$ and
    $\bq_2 = (1/\sqrt{2},\,-1/\sqrt{2})^T$ form an orthonormal basis for $\fR^2$.
    \item Express $\bv = (3, -1)^T$ as $\sum_i (\bq_i^T\bv)\bq_i$ using the basis above.
    Verify the result equals $\bv$.
    \item A set of $n$ mutually orthogonal nonzero vectors in $\fR^n$ is linearly
    independent. Prove this. (\textit{Hint: assume $\sum c_i\bv_i = \bzero$ and take the
    inner product with each $\bv_j$.})
\end{enumerate}
}

\addDef{Orthogonal Matrix}{%
A square matrix $\bQ \in \fR^{n \times n}$ is \emph{orthogonal} if its columns form an
orthonormal set:
\[
\bQ^T\bQ = \bQ\bQ^T = \bI_n.
\]
Equivalently, $\bQ^T = \bQ^{-1}$, meaning the transpose is its own inverse.
}
\label{def:orthogonal-matrix}

\textBox{%
Orthogonal matrices represent rotations, reflections, or combinations of these in $\fR^n$.
They preserve vector lengths and angles between vectors, making them particularly valuable
in scientific computing where maintaining geometric properties is important.
}

\addExcr{%
Refer to Def~\ref{def:orthogonal-matrix}.
\begin{enumerate}
    \item Verify that $\bQ = \begin{pmatrix}\cos\theta & -\sin\theta \\ \sin\theta &
    \cos\theta\end{pmatrix}$ is orthogonal for any $\theta$. What geometric operation does
    multiplication by $\bQ$ perform?
    \item Is $\bM = \begin{pmatrix}1/\sqrt{2} & 1/\sqrt{2} \\ 1/\sqrt{2} &
    1/\sqrt{2}\end{pmatrix}$ orthogonal? Why or why not?
    \item If $\bQ$ is orthogonal, show that $\bQ^T$ is also orthogonal.
\end{enumerate}
}

\addThm{Orthogonal Matrices Preserve Norms and Inner Products}{%
Let $\bQ \in \fR^{n \times n}$ be an orthogonal matrix. For all $\bx, \by \in \fR^n$:
\begin{enumerate}
    \item $(\bQ\bx)^T(\bQ\by) = \bx^T\by$ \quad (inner-product-preserving)
    \item $\|\bQ\bx\|_2 = \|\bx\|_2$ \quad (isometry)
    \item $\kappa_2(\bQ) = 1$ \quad (optimally conditioned)
\end{enumerate}
}
\label{thm:orthogonal-isometry}

\addPrf{%
(1): $(\bQ\bx)^T(\bQ\by) = \bx^T\bQ^T\bQ\by = \bx^T\bI\by = \bx^T\by$.

(2): Set $\bx = \by$ in (1): $\|\bQ\bx\|_2^2 = \bx^T\bx = \|\bx\|_2^2$.

(3): By (2), $\|\bQ\|_2 = \max_{\|\bx\|=1}\|\bQ\bx\|_2 = 1$. Since $\bQ^{-1} = \bQ^T$ is
also orthogonal, $\|\bQ^{-1}\|_2 = 1$. Thus $\kappa_2(\bQ) = \|\bQ\|_2\|\bQ^{-1}\|_2 = 1$.
}

\addNote{%
The condition number of 1 is the best possible. Multiplying by an orthogonal matrix
introduces no amplification of errors, which is why orthogonal transformations (Householder
reflections, Givens rotations) are preferred over Gaussian elimination when stability is
paramount.
}

\addProp{Rows of an Orthogonal Matrix are Orthonormal}{%
If $\bQ \in \fR^{n \times n}$ satisfies $\bQ^T\bQ = \bI$, then $\bQ\bQ^T = \bI$ as well,
so the rows of $\bQ$ also form an orthonormal basis for $\fR^n$.
}
\label{prop:orthogonal-rows}

\addPrf{%
$\bQ^T\bQ = \bI$ implies $\bQ$ is invertible with $\bQ^{-1} = \bQ^T$. Left-multiplying by
$\bQ$ and right-multiplying by $\bQ^T$ gives $\bQ\bQ^T = \bI$.
}

\addProp{Determinant of an Orthogonal Matrix}{%
For any orthogonal $\bQ$, $\det(\bQ) = \pm 1$. When $\det(\bQ) = +1$, $\bQ$ is a rotation;
when $\det(\bQ) = -1$, it includes a reflection.
}
\label{prop:orthogonal-det}

\addPrf{%
From $\bQ^T\bQ = \bI$: $\det(\bQ^T)\det(\bQ) = 1$, and $\det(\bQ^T) = \det(\bQ)$, so
$\det(\bQ)^2 = 1$.
}

\addProp{Product of Orthogonal Matrices is Orthogonal}{%
If $\bQ_1, \bQ_2 \in \fR^{n \times n}$ are orthogonal, then $\bQ_1\bQ_2$ is also
orthogonal.
}
\label{prop:orthogonal-product}

\addPrf{%
$(\bQ_1\bQ_2)^T(\bQ_1\bQ_2) = \bQ_2^T(\bQ_1^T\bQ_1)\bQ_2 = \bQ_2^T\bI\bQ_2 = \bI$.
}

\addExcr{%
Refer to Props~\ref{prop:orthogonal-rows}--\ref{prop:orthogonal-product}.
\begin{enumerate}
    \item For $\bQ = \begin{pmatrix}0 & -1 \\ 1 & 0\end{pmatrix}$ (rotation by 90°),
    compute $\det(\bQ)$ and verify the rows are orthonormal.
    \item For the reflection $\bQ = \begin{pmatrix}1 & 0 \\ 0 & -1\end{pmatrix}$, verify
    $\det(\bQ) = -1$ and $\bQ\bQ^T = \bI$.
    \item Is $\bQ_1 + \bQ_2$ necessarily orthogonal when $\bQ_1$ and $\bQ_2$ are? Give a
    counterexample.
\end{enumerate}
}

\subsubsection{QR Decomposition}

The QR decomposition factors a matrix into an orthogonal factor $\bQ$ and an upper
triangular factor $\bR$, combining the stability benefits of orthogonal transformations with
the efficiency of triangular solves. It is the factorization of choice for least-squares
problems and forms the basis of the QR eigenvalue algorithm.

\addDef{QR Decomposition}{%
For a matrix $\bA \in \fR^{m \times n}$ with $m \geq n$, the \emph{QR decomposition}
factors $\bA$ as
\[
\bA = \bQ\bR,
\]
where $\bQ \in \fR^{m \times m}$ is orthogonal ($\bQ^T\bQ = \bI$) and $\bR \in \fR^{m \times n}$
is upper triangular with zeros below the main diagonal.
}
\label{def:qr}

\textBox{%
The QR factorization is particularly valuable in scientific computing because it transforms
problems involving arbitrary matrices into problems with triangular matrices, which are much
easier to solve. The orthogonality of $\bQ$ ensures numerical stability in many applications,
as orthogonal transformations preserve Euclidean norms and do not amplify errors.
}

\addExcr{%
Refer to Def~\ref{def:qr}.
\begin{enumerate}
    \item If $\bA = \bQ\bR$, explain how to solve $\bA\bx = \bb$ using $\bQ$ and $\bR$
    separately. What is the cost of each step?
    \item For $\bA = \begin{pmatrix}1 & 0 \\ 0 & 1 \\ 0 & 0\end{pmatrix}$, write a QR
    factorization by inspection. (\textit{Hint: $\bA$ already has orthonormal columns.})
    \item Why does the QR decomposition always exist for any $\bA \in \fR^{m \times n}$ with
    $m \geq n$? (\textit{Hint: consider Gram-Schmidt applied to the columns of $\bA$.})
\end{enumerate}
}

\addProp{QR and the Column Space}{%
If $\bA \in \fR^{m \times n}$ ($m \geq n$) has full column rank, the first $n$ columns of
$\bQ$ (from the full QR factorization $\bA = \bQ\bR$) form an orthonormal basis for
$\text{col}(\bA)$.
}
\label{prop:qr-colspace}

\addProp{Diagonal Entries of $\bR$}{%
In the QR factorization of a full-column-rank matrix $\bA$, the diagonal entries of $\bR$
satisfy $|r_{ii}| > 0$ for each $i$. In particular, $\bR$ is invertible.
}
\label{prop:r-diagonal}

\addExcr{%
Refer to Props~\ref{prop:qr-colspace}--\ref{prop:r-diagonal}.
\begin{enumerate}
    \item If $r_{11} = 0$ in the QR factorization of $\bA$, what does this imply about the
    first column of $\bA$?
    \item For a $3 \times 2$ full-rank matrix $\bA$, identify which columns of the full
    $\bQ \in \fR^{3 \times 3}$ span $\text{col}(\bA)$ and which span
    $\text{col}(\bA)^\perp$.
    \item In NumPy, compute the full and thin QR of a $4 \times 3$ matrix and verify:
    (a) the first 3 columns of $\bQ$ span $\text{col}(\bA)$, (b) the 4th column of $\bQ$
    is orthogonal to each column of $\bA$.
\end{enumerate}
}

\subsubsection{Householder Reflectors}

A key method for computing the QR factorization is via \emph{Householder reflectors}:
orthogonal matrices that reflect a vector across a hyperplane and can zero out all entries
below a chosen position in a column. Applying $n$ such reflectors in sequence reduces $\bA$
to upper triangular form.

\addDef{Householder Reflector}{%
Given a nonzero vector $\bv \in \fR^m$, the \emph{Householder reflector} is the matrix
\[
  \bH = \bI - \frac{2\bv\bv^T}{\bv^T\bv}.
\]
Geometrically, $\bH\bx$ reflects $\bx$ across the hyperplane
$\bv^\perp = \{\bw : \bv^T\bw = 0\}$.
}
\label{def:householder}

\addThm{Properties of Householder Reflectors}{%
For any nonzero $\bv \in \fR^m$, the Householder matrix $\bH = \bI - 2\bv\bv^T/\|\bv\|_2^2$
satisfies:
\begin{enumerate}
  \item $\bH^T = \bH$ \quad (symmetric)
  \item $\bH^T\bH = \bI$ \quad (orthogonal)
  \item $\bH^2 = \bI$, so $\bH^{-1} = \bH$ \quad (involution)
  \item $\det(\bH) = -1$ \quad (reflection, not rotation)
\end{enumerate}
}
\label{thm:householder-props}

\addPrf{%
Let $\tau = \bv^T\bv = \|\bv\|_2^2 > 0$.

(1): $\bH^T = (\bI - 2\bv\bv^T/\tau)^T = \bI - 2\bv\bv^T/\tau = \bH$.

(2): $\bH^T\bH = \bH^2 = (\bI - 2\bv\bv^T/\tau)^2 = \bI - 4\bv\bv^T/\tau +
4\bv(\bv^T\bv)\bv^T/\tau^2 = \bI - 4\bv\bv^T/\tau + 4\bv\bv^T/\tau = \bI$.

(3): Follows from (1) and (2): $\bH^2 = \bH^T\bH = \bI$.

(4): $\bH$ has eigenvalue $-1$ (for any $\bx \parallel \bv$: $\bH\bv = \bv - 2\bv = -\bv$)
with multiplicity 1, and eigenvalue $+1$ (for $\bx \perp \bv$) with multiplicity $m-1$, so
$\det(\bH) = (-1)^1 \cdot 1^{m-1} = -1$.
}

\addThm{Householder Zeroing}{%
For any $\bx \in \fR^m$ with $\bx \neq \pm\|\bx\|_2\be_1$, there exists a Householder
vector $\bv$ such that
\[
  \bH\bx = \pm\|\bx\|_2\be_1,
\]
where $\be_1 = (1,0,\ldots,0)^T$. The choice $\bv = \bx + \|\bx\|_2\be_1$ (taking the
$+$ sign) avoids cancellation and is numerically preferred.
}
\label{thm:householder-zeroing}

\addPrf{%
We want $\bH\bx = \alpha\be_1$ for some scalar $\alpha$. Since $\bH$ is orthogonal,
$|\alpha| = \|\bx\|_2$. Setting $\bv = \bx - \alpha\be_1$:
\[
  \bH\bx = \bx - \frac{2\bv\bv^T\bx}{\bv^T\bv}.
\]
We compute $\bv^T\bx = \|\bx\|_2^2 - \alpha x_1$ and
$\bv^T\bv = 2(\|\bx\|_2^2 - \alpha x_1) = 2\bv^T\bx$, so $\bH\bx = \bx - \bv = \alpha\be_1$.
Choosing $\alpha = -\text{sign}(x_1)\|\bx\|_2$ avoids cancellation when $x_1$ is large.
}

\addNote{%
The key use of Householder reflectors in QR factorization is iterative zeroing: at step $k$,
choose $\bH_k$ to zero all entries below the $(k,k)$ position in column $k$ of the current
matrix. After $n$ such reflections, the result is upper triangular. Since each $\bH_k$ is
orthogonal, the product $\bQ = \bH_1\bH_2\cdots\bH_n$ is orthogonal, giving $\bA = \bQ\bR$.
Each reflector is applied as $\bH\bx = \bx - 2\bv(\bv^T\bx)/\|\bv\|^2$ --- a rank-1 update
costing only $O(m)$, not $O(m^2)$.
}

\addExcr{%
Refer to Def~\ref{def:householder} and Thm~\ref{thm:householder-zeroing}.
\begin{enumerate}
  \item Let $\bx = (3, 4)^T \in \fR^2$. Find the Householder vector $\bv$ such that
  $\bH\bx = -5\be_1$. Write $\bH$ explicitly and verify $\bH\bx = (-5, 0)^T$.
  \item For $\bx = (1,2,2)^T$, find $\bv$ and $\bH$ such that $\bH\bx = -3\be_1$. Verify
  $\bH$ is symmetric and $\bH^2 = \bI$.
  \item In NumPy, apply two successive Householder reflectors to triangularize
  $\bA = \begin{pmatrix} 1 & 2 \\ 2 & 3 \\ 2 & 4 \end{pmatrix}$
  and recover $\bQ$ and $\bR$. Compare with \texttt{np.linalg.qr(A)}.
\end{enumerate}
}

\addLem{QR Factorization Cost}{%
The QR factorization of $\bA \in \fR^{m \times n}$ (via Householder reflections) costs
approximately $2mn^2 - \frac{2}{3}n^3$ flops.
}
\label{lem:qr-cost}

\addExcr{%
Refer to Lem~\ref{lem:qr-cost}.
\begin{enumerate}
    \item For $\bA \in \fR^{n \times n}$ (square), simplify the QR cost formula. Compare it
    to LU factorization cost ($\frac{2}{3}n^3$). Which is cheaper and by what factor?
    \item Why does a zero diagonal entry in $\bR$ imply $\bA$ is rank-deficient?
    (\textit{Hint: $\bA = \bQ\bR$ and $\bQ$ is invertible.})
    \item In NumPy, \texttt{np.linalg.qr(A)} returns the thin QR by default
    (\texttt{mode='reduced'}). What are the shapes of $\bQ$ and $\bR$ for a $5 \times 3$
    matrix? When would you want the full QR (\texttt{mode='complete'})?
\end{enumerate}
}

\subsubsection{Gram-Schmidt}

The Gram-Schmidt process is a constructive algorithm for building an orthonormal basis from
any set of linearly independent vectors. It directly yields the QR factorization: the
orthonormal vectors form the columns of $\bQ$ and the projection coefficients populate
$\bR$. Two variants exist --- classical and modified --- with different numerical stability
properties.

\addDef{Orthogonal Complement}{%
Let $S \subseteq \fR^n$ be a subspace. Its \emph{orthogonal complement} is
\[
    S^\perp = \{\bw \in \fR^n : \bw^T\bv = 0 \text{ for all } \bv \in S\}.
\]
$S^\perp$ is itself a subspace of $\fR^n$.
}
\label{def:orthogonal-complement}

\addProp{Dimension of the Orthogonal Complement}{%
If $S \subseteq \fR^n$ has dimension $k$, then $\dim(S^\perp) = n - k$ and
$S \cap S^\perp = \{\bzero\}$.
}
\label{prop:complement-dim}

\addProp{Orthogonal Decomposition}{%
Every vector $\bx \in \fR^n$ decomposes uniquely as $\bx = \bx_S + \bx_{S^\perp}$ where
$\bx_S \in S$ and $\bx_{S^\perp} \in S^\perp$. The component $\bx_S$ is the orthogonal
projection of $\bx$ onto $S$.
}
\label{prop:orthogonal-decomp}

\addProp{Null Space as an Orthogonal Complement}{%
For $\bA \in \fR^{m \times n}$ with columns spanning $S = \text{col}(\bA)$, the orthogonal
complement satisfies $S^\perp = \ker(\bA^T)$, i.e.,
$S^\perp = \{\bw : \bA^T\bw = \bzero\}$.
}
\label{prop:null-complement}

\addExcr{%
Refer to Def~\ref{def:orthogonal-complement} and
Props~\ref{prop:complement-dim}--\ref{prop:null-complement}.
\begin{enumerate}
    \item Let $S = \text{span}\{(1,1,0)^T, (0,1,1)^T\}$ in $\fR^3$. Find a basis for
    $S^\perp$ by solving $\bA^T\bw = \bzero$ where $\bA$ has the spanning vectors as
    columns.
    \item Verify that your answer to (1) satisfies $\dim(S) + \dim(S^\perp) = 3$.
    \item In NumPy, use the QR factorization of $\bA$ to extract an orthonormal basis for
    $\text{col}(\bA)$ and for $\text{col}(\bA)^\perp = \ker(\bA^T)$.
\end{enumerate}
}

\addDef{Gram-Schmidt Process}{%
The \emph{Gram-Schmidt process} converts a set of linearly independent vectors
$\{\bv_1, \bv_2, \ldots, \bv_n\}$ into an orthonormal basis for the same subspace, by
sequentially removing each vector's projection onto the span of previously orthogonalized
vectors.
}
\label{def:gram-schmidt}

\textBox{%
\textbf{Classical Gram-Schmidt Algorithm.}\ Given linearly independent vectors
$\{\bv_1, \ldots, \bv_n\}$, compute an orthogonal set $\{\bu_1, \ldots, \bu_n\}$ by:
\begin{align}
\bu_1 &= \bv_1, \\
\bu_k &= \bv_k - \sum_{j=1}^{k-1}\frac{\bu_j^T\bv_k}{\bu_j^T\bu_j}\bu_j,
\quad k = 2, \ldots, n.
\end{align}
Normalize each vector to obtain an orthonormal basis: $\bq_i = \bu_i / \|\bu_i\|_2$.
}

\addNote{%
The Gram-Schmidt process can be interpreted geometrically:
\begin{itemize}
\item $\bu_1$ is simply the first vector $\bv_1$.
\item $\bu_2$ is $\bv_2$ minus its projection onto $\bu_1$.
\item $\bu_3$ is $\bv_3$ minus its projections onto both $\bu_1$ and $\bu_2$.
\end{itemize}
Each $\bu_k$ is the component of $\bv_k$ that is orthogonal to the subspace spanned by
$\{\bu_1, \ldots, \bu_{k-1}\}$.
}

\addExmpl{Gram-Schmidt Process}{%
Let $\bv_1 = (1, 1, 1)^T$ and $\bv_2 = (0, 1, 2)^T$ in $\fR^3$.

\textit{Step 1:} $\bu_1 = \bv_1 = (1, 1, 1)^T$.

\textit{Step 2:}
\begin{align}
\bu_2 &= \bv_2 - \frac{\bu_1^T\bv_2}{\bu_1^T\bu_1}\bu_1
       = (0,1,2)^T - \frac{3}{3}(1,1,1)^T = (-1, 0, 1)^T.
\end{align}

Normalizing:
\begin{align}
\bq_1 &= \tfrac{1}{\sqrt{3}}(1, 1, 1)^T, \qquad
\bq_2 = \tfrac{1}{\sqrt{2}}(-1, 0, 1)^T.
\end{align}
}

\addExcr{%
Refer to Def~\ref{def:gram-schmidt}.
\begin{enumerate}
    \item Apply Gram-Schmidt to $\bv_1 = (1, 0, 1)^T$, $\bv_2 = (1, 1, 0)^T$,
    $\bv_3 = (0, 1, 1)^T$. Verify each pair of output vectors is orthogonal.
    \item Show that the Gram-Schmidt process fails (produces $\bu_k = \bzero$) if and only
    if the input vectors are linearly dependent.
    \item In NumPy, implement the classical Gram-Schmidt algorithm for a $4 \times 3$
    matrix $\bA$ and compare your output $\bQ$ with \texttt{np.linalg.qr(A)[0]}.
\end{enumerate}
}

\textBox{%
The Gram-Schmidt process provides a constructive proof of the existence of an orthonormal
basis for any subspace of $\fR^n$. In practice, however, the classical algorithm can suffer
from numerical instability due to accumulated rounding errors, especially for large sets of
nearly dependent vectors. The modified Gram-Schmidt algorithm, which applies
orthogonalization against each previous vector sequentially (using the partially
orthogonalized vector at each step), offers better numerical stability and is preferred in
computational applications.
}

\textBox{%
\textbf{Modified Gram-Schmidt Algorithm.}\ Instead of computing all projections from the
original vector $\bv_k$, modified Gram-Schmidt re-uses the \emph{partially orthogonalized}
vector at each step:
\begin{align}
\bv_k^{(1)} &= \bv_k, \\
\bv_k^{(j+1)} &= \bv_k^{(j)} - \frac{\bu_j^T\,\bv_k^{(j)}}{\bu_j^T\bu_j}\bu_j,
\quad j = 1, \ldots, k-1, \\
\bu_k &= \bv_k^{(k)}.
\end{align}
Normalize as before: $\bq_i = \bu_i / \|\bu_i\|_2$.

In exact arithmetic, classical and modified Gram-Schmidt produce identical results. The
difference is numerical: modified Gram-Schmidt limits error accumulation by immediately
removing components already seen.
}

\addExmpl{Gram-Schmidt and QR Decomposition}{%
Consider the matrix $\bA = \begin{bmatrix} 1 & 0 \\ 1 & 1 \\ 1 & 2 \end{bmatrix}$ whose
columns are $\bv_1 = (1,1,1)^T$ and $\bv_2 = (0,1,2)^T$ from the previous example.

From the Gram-Schmidt process, $\bq_1 = \frac{1}{\sqrt{3}}(1,1,1)^T$ and
$\bq_2 = \frac{1}{\sqrt{2}}(-1,0,1)^T$. The $\bR$ matrix entries are:
\begin{align}
r_{11} &= \bq_1^T\bv_1 = \sqrt{3}, \quad r_{12} = \bq_1^T\bv_2 = \sqrt{3}, \quad
r_{22} = \bq_2^T\bu_2 = \sqrt{2}.
\end{align}

Therefore:
\[
\bR = \begin{bmatrix} \sqrt{3} & \sqrt{3} \\ 0 & \sqrt{2} \end{bmatrix}, \qquad
\bQ\bR = \begin{bmatrix} \frac{1}{\sqrt{3}} & -\frac{1}{\sqrt{2}} \\
\frac{1}{\sqrt{3}} & 0 \\ \frac{1}{\sqrt{3}} & \frac{1}{\sqrt{2}} \end{bmatrix}
\begin{bmatrix} \sqrt{3} & \sqrt{3} \\ 0 & \sqrt{2} \end{bmatrix}
= \begin{bmatrix} 1 & 0 \\ 1 & 1 \\ 1 & 2 \end{bmatrix} = \bA.\;\checkmark
\]
}

\addThm{Classical vs.\ Modified Gram-Schmidt Numerical Stability}{%
Both CGS and MGS produce the QR factorization in exact arithmetic, but differ numerically.
Let $\hat{\bQ}$ denote the computed factor. Then:
\begin{itemize}
    \item \textbf{Classical (CGS):}
    $\|\hat{\bQ}^T\hat{\bQ} - \bI\|_2 = O\!\left(\varepsilon_{\text{mach}}\,\kappa(\bA)^2\right)$
    \item \textbf{Modified (MGS):}
    $\|\hat{\bQ}^T\hat{\bQ} - \bI\|_2 = O\!\left(\varepsilon_{\text{mach}}\,\kappa(\bA)\right)$
\end{itemize}
Both algorithms cost $O(mn^2)$ flops for $\bA \in \fR^{m \times n}$.
}
\label{thm:cgs-mgs}

\addNote{%
The key difference: CGS computes all projections from the \emph{original} vector $\bv_k$,
so rounding errors in early projections contaminate all subsequent ones. MGS re-uses the
\emph{partially orthogonalized} vector at each step, immediately removing components
already seen, which limits error accumulation. For maximum stability (at the same
$O(mn^2)$ cost), Householder QR achieves $O(\varepsilon_{\text{mach}})$ loss of
orthogonality regardless of $\kappa(\bA)$; it is what \texttt{np.linalg.qr} actually uses.
}

\addNote{%
The Gram-Schmidt process is closely related to the QR decomposition. Arranging the original
vectors as columns of $\bA$ and the orthonormal vectors as columns of $\bQ$, we have
$\bA = \bQ\bR$ where $\bR$ is upper triangular with entries:
\[
r_{ij} =
\begin{cases}
\bq_i^T\bv_j & \text{if } i \leq j \\
0 & \text{if } i > j.
\end{cases}
\]
The elements of $\bR$ are precisely the projection coefficients computed during the
Gram-Schmidt process.
}

\addExcr{%
Let $\bA = \begin{pmatrix} 1 & 1 \\ 1 & 2 \\ 0 & 1 \end{pmatrix}$.
\begin{enumerate}
    \item Apply classical Gram-Schmidt by hand to the columns of $\bA$ to find an
    orthonormal basis $\{\bq_1, \bq_2\}$ for $\text{col}(\bA)$.
    \item Write out $\bQ \in \fR^{3 \times 2}$ and $\bR \in \fR^{2 \times 2}$ and verify
    $\bA = \bQ\bR$.
    \item Apply modified Gram-Schmidt to the same columns. Do you get the same $\bQ, \bR$
    (in exact arithmetic)?
    \item In NumPy, compute \texttt{np.linalg.qr(A)} and compare. Verify
    \texttt{Q.T @ Q} is close to $\bI$.
\end{enumerate}
}

\addExcr{%
\textbf{(Orthogonal Projection)} Let $\bQ \in \fR^{m \times n}$ (with $m > n$) have
orthonormal columns (thin QR). The orthogonal projector onto $\text{col}(\bQ)$ is
$\bP = \bQ\bQ^T$.
\begin{enumerate}
    \item Show that $\bP^2 = \bP$ (idempotent) and $\bP^T = \bP$ (symmetric).
    \item Show that $\bP\bx$ is the closest vector in $\text{col}(\bQ)$ to $\bx$, i.e.,
    $\|\bx - \bP\bx\|_2 \leq \|\bx - \by\|_2$ for all $\by \in \text{col}(\bQ)$.
    \item If $\bA = \bQ\bR$ (full QR with $\bQ \in \fR^{m \times m}$ orthogonal), what is
    the projection onto $\text{col}(\bA)$ in terms of the thin $\bQ_1$ (first $n$ columns
    of $\bQ$)?
    \item Write NumPy code to project a vector \texttt{x = np.random.randn(5)} onto the
    column space of a random $5 \times 3$ matrix \texttt{A}. Verify the residual
    \texttt{x - P @ x} is orthogonal to each column of \texttt{A}.
\end{enumerate}
}

\addExcr{%
\textbf{(Stability Experiment)} Generate a nearly linearly dependent set of vectors:
\begin{center}
\texttt{V = np.column\_stack([np.ones(50), np.linspace(0,1,50)**k for k in range(1,6)])}
\end{center}
This is a Vandermonde-like matrix with $\kappa(\bV) \gg 1$.
\begin{enumerate}
    \item Implement classical Gram-Schmidt and compute $\|\hat{\bQ}^T\hat{\bQ} - \bI\|_F$.
    \item Implement modified Gram-Schmidt and compute the same quantity. Compare.
    \item Use \texttt{np.linalg.qr(V)} (Householder) and compute the same quantity.
    What do you observe?
    \item Compute $\kappa(\bV)$ via \texttt{np.linalg.cond(V)}. Does the stability gap
    between CGS and MGS scale as Thm~\ref{thm:cgs-mgs} predicts?
\end{enumerate}
}

\addNote{%
The Gram-Schmidt process and QR factorization are foundational tools in numerical linear
algebra, underlying algorithms for solving linear systems, least-squares problems,
eigenvalue computation (the QR algorithm), and orthogonal basis construction in iterative
solvers.
}

\addExcr{%
For each application below, identify which property of QR (orthogonality of $\bQ$,
triangularity of $\bR$, or both) is the key ingredient and write one sentence explaining
why.
\begin{enumerate}
    \item Solving a square linear system $\bA\bx = \bb$ via $\bA = \bQ\bR$.
    \item Computing the least-squares solution to an overdetermined system
    $\bA\bx \approx \bb$.
    \item One step of the QR eigenvalue algorithm: $\bA_k = \bQ_k\bR_k$,
    $\bA_{k+1} = \bR_k\bQ_k$.
    \item Constructing a coordinate system aligned with a given set of data directions.
\end{enumerate}
}

\end{document}
