\documentclass[../readings.tex]{subfiles}

\begin{document}

\addcontentsline{toc}{section}{Preface}
\section*{Preface}

These notes started as a study guide for CS 111 at UC Santa Barbara and grew through my PhD as teaching and research kept exposing things I thought I understood until I tried to explain them: why LU factorization is numerically stable, why the QR algorithm converges, what the Koopman operator actually is. Writing forces the gaps to surface. The result is a reference written by a graduate student for himself. Treat it accordingly.

\medskip

For a rigorous treatment of this material, see \href{https://math.nyu.edu/~shelley/Classes/SciComp/BindelGoodman.pdf}{Principles of Scientific Computing} by Bindel and Goodman, and G. W. Stewart's \href{https://epubs.siam.org/doi/book/10.1137/1.9781611971491}{Afternotes on Numerical Analysis}. For further reading, the \href{https://www.cs.cornell.edu/courses/cs6241/2025sp/reading.html}{CS 6241 reading list} at Cornell covers the literature well.

\medskip

Computers were built to do numerical computation. The word ``computer'' originally meant a person whose job was to evaluate functions by hand. The U.S. Army employed hundreds of them during the Second World War to compute artillery firing tables, a task so large it directly motivated the construction of \href{https://www.computerhistory.org/revolution/birth-of-the-computer/4/78/323}{ENIAC} (1945). The machine's first classified application was thermonuclear weapon simulation, directed by John von Neumann, whose \href{https://projecteuclid.org/journals/bulletin-of-the-american-mathematical-society-new-series/volume-53/issue-11/Numerical-inverting-of-matrices-of-high-order/bams/1183511222.full}{1947 paper with Goldstine} laid the foundations of numerical stability analysis. In 1950, Jule Charney used ENIAC to run the \href{https://maths.ucd.ie/~plynch/eniac/CFvN-1950.pdf}{first machine weather forecast}: \href{https://archive.org/details/weatherpredictio00richrich}{Lewis Fry Richardson} had estimated the same computation would require 64,000 human computers working in parallel. The \href{https://ntrs.nasa.gov/citations/19760016247}{Apollo guidance computer} navigated to the Moon in real time with 4 kilobytes of memory. The \href{https://www.ams.org/journals/mcom/1965-19-090/S0025-5718-1965-0178586-1/S0025-5718-1965-0178586-1.pdf}{Cooley-Tukey FFT} (1965) reduced a core signal processing computation from $O(n^2)$ to $O(n \log n)$. The demand for numerical computation at scale was the original motivation for programmable machines.

\medskip

Almost every numerical algorithm reduces a nonlinear problem to a sequence of linear ones, because linear systems are the only class of problems we understand completely: their solution structure, their sensitivity to perturbations, their reliable algorithms. Newton's method, finite element discretization, PCA, gradient-based optimization, each is, at its core, a repeated linearization. Linear algebra is the substrate of scientific computing.

\medskip

Much of the perspective here came from \textbf{CS 210A: Matrix Analysis}, taught by Professor Shivkumar Chandrasekaran in the ECE department at UC Santa Barbara. If you have the chance to take that course, take it.

\vspace{1em}
\begin{flushright}
\textit{Aditya Dendukuri} \\
\textit{UC Santa Barbara, 2025}
\end{flushright}

\newpage

\end{document}
