\documentclass[../readings.tex]{subfiles}

\begin{document}
\subsection{Root Finding}
\label{subsec:root-finding}

\textBox{%
Given a continuous function $f : \fR \to \fR$, find $x^*$ such that $f(x^*) = 0$.
Root-finding methods are iterative: starting from an initial guess (or bracket), they
refine the estimate until convergence. The key distinctions are \emph{bracketing} methods
(guaranteed convergence, slower) versus \emph{open} methods (faster convergence, may diverge).
}

\subsubsection{Bisection}

\addThm{Bisection Method}{%
Suppose $f$ is continuous on $[a, b]$ with $f(a)f(b) < 0$. By the Intermediate Value
Theorem, there exists $x^* \in (a, b)$ with $f(x^*) = 0$. The bisection method halves
the bracket at each step: set $m = (a + b)/2$, then replace $[a,b]$ by $[a,m]$ if
$f(a)f(m) < 0$, otherwise by $[m,b]$. After $n$ steps,
\[
|x_n - x^*| \leq \frac{b - a}{2^n}.
\]
}

\addNote{%
Bisection converges \emph{linearly} with rate $1/2$: each step gains exactly one binary
digit of accuracy. It requires only continuity of $f$ (no derivative) and is guaranteed
to converge as long as the initial bracket contains a root.
}

\subsubsection{Newton-Raphson Method}

\addDef{Newton's Method}{%
Starting from an initial guess $x_0$, Newton's method linearizes $f$ at the current
iterate and solves for the root of the linear approximation:
\[
x_{k+1} = x_k - \frac{f(x_k)}{f'(x_k)}.
\]
}

\addThm{Quadratic Convergence of Newton's Method}{%
If $f$ is twice continuously differentiable near a \emph{simple root} $x^*$
(i.e., $f'(x^*) \neq 0$), then Newton's method converges locally with \emph{quadratic rate}:
\[
|x_{k+1} - x^*| \approx \frac{|f''(x^*)|}{2|f'(x^*)|} |x_k - x^*|^2.
\]
The number of correct decimal digits roughly doubles at each step.
}

\addPrf{%
Let $e_k = x_k - x^*$. Taylor-expand $f$ about $x^*$:
\[
f(x_k) = f'(x^*)e_k + \tfrac{1}{2}f''(\xi_k)e_k^2
\]
for some $\xi_k$ between $x_k$ and $x^*$. Similarly, $f'(x_k) = f'(x^*) + O(e_k)$. Then
\[
e_{k+1} = e_k - \frac{f(x_k)}{f'(x_k)}
= e_k - \frac{f'(x^*)e_k + \tfrac{1}{2}f''(\xi_k)e_k^2}{f'(x^*) + O(e_k)}
= \frac{f''(\xi_k)}{2f'(x^*)}e_k^2 + O(e_k^3).
\]
}

\addNote{%
Newton's method can diverge if $x_0$ is far from the root, if $f'(x_k) \approx 0$
(nearly flat), or at a double root $f'(x^*) = 0$ (convergence degrades to linear with
rate $1/2$).
}

\subsubsection{Secant Method}

\addDef{Secant Method}{%
When $f'$ is unavailable or expensive, approximate $f'(x_k)$ by a finite difference
using the previous iterate:
\[
x_{k+1} = x_k - f(x_k)\frac{x_k - x_{k-1}}{f(x_k) - f(x_{k-1})}.
\]
The secant method requires two initial points $x_0, x_1$ and does not evaluate $f'$.
}

\addThm{Superlinear Convergence of the Secant Method}{%
Near a simple root $x^*$ with $f$ twice differentiable, the secant method converges with
order $\phi = \frac{1 + \sqrt{5}}{2} \approx 1.618$ (the golden ratio):
\[
|e_{k+1}| \approx C |e_k|^{\phi}, \qquad C = \left|\frac{f''(x^*)}{2f'(x^*)}\right|^{\phi - 1}.
\]
}

\addNote{%
The secant method is \emph{superlinear} --- faster than bisection's linear rate but slower
than Newton's quadratic rate. Each step costs one function evaluation (versus two for
finite-difference Newton), making it often more efficient per digit gained.
}

\subsubsection{Brent's Method}

\textBox{%
Brent's method combines three ideas to get the best of bracketing and open methods:
(1) \emph{bisection} as a fallback, always maintaining a bracket;
(2) the \emph{secant method} for fast superlinear convergence near the root;
(3) \emph{inverse quadratic interpolation (IQI)}, fitting a quadratic through the last
three points $(a, f(a))$, $(b, f(b))$, $(c, f(c))$ and evaluating it at $y = 0$:
\[
x = \frac{a f(b)f(c)}{(f(a)-f(b))(f(a)-f(c))} + \frac{b f(a)f(c)}{(f(b)-f(a))(f(b)-f(c))} + \frac{c f(a)f(b)}{(f(c)-f(a))(f(c)-f(b))}.
\]
}

\addThm{Brent's Method Guarantees}{%
Brent's method is \emph{guaranteed to converge} (like bisection) and achieves
\emph{superlinear convergence} near a root (like the secant method). It accepts an IQI
or secant step only when it falls inside the bracket and improves sufficiently over
bisection; otherwise it takes a bisection step.
}

\addNote{%
Brent's method is the default solver in \texttt{scipy.optimize.brentq}, the GNU Scientific
Library, and MATLAB's \texttt{fzero}.
}

\subsubsection{Exercises}

\addExcr{%
\begin{enumerate}
    \item Apply 4 steps of bisection to find a root of $f(x) = x^3 - 2$ on $[1, 2]$.
    What is the guaranteed error bound after these steps? Compare to $2^{1/3}$.
    \item Apply 3 steps of Newton's method to $f(x) = x^2 - 2$ starting from $x_0 = 1$.
    Compute $|x_k - \sqrt{2}|$ at each step and verify quadratic convergence by checking
    the ratio $|e_{k+1}|/|e_k|^2$.
    \item Show that Newton's method applied to $f(x) = x^2$ converges only \emph{linearly}
    to $x^* = 0$ (double root). What is the linear convergence rate?
    \item Derive the convergence order $\phi = (1+\sqrt{5})/2$ of the secant method.
    (\emph{Hint:} assume $|e_{k+1}| \approx C|e_k|^\phi$, use $e_{k+1} \approx
    \frac{f''}{2f'} e_k e_{k-1}$, and equate exponents.)
    \item Find the positive root of $f(x) = e^x - 3x = 0$ near $x = 1$ using bisection
    on $[0, 2]$, Newton's method from $x_0 = 1$, and the secant method from $x_0 = 1$,
    $x_1 = 1.5$. Compare the number of function evaluations to reach $|f(x_k)| < 10^{-10}$.
\end{enumerate}
}

\end{document}
