\documentclass[../readings.tex]{subfiles}

\begin{document}

\subsection{Symmetric Positive Definite (SPD) Matrices}

\addDef{Symmetric Matrix}{%
A matrix \(\bA \in \mathbb{R}^{n \times n}\) is called \emph{symmetric} if 
\[
\bA = \bA^T,
\]
meaning that the entry in the \(i\)th row and \(j\)th column is the same as the entry in the \(j\)th row and \(i\)th column for all \(1 \le i,j \le n\).
}

\addDef{Positive Definite Matrix}{%
A symmetric matrix \(\bA \in \mathbb{R}^{n \times n}\) is \emph{positive definite} (PD) if
\[
\bx^T\bA\bx > 0 \quad \text{for every nonzero } \bx \in \mathbb{R}^n.
\]
In computational terms, when we transform any nonzero vector \(\bx\) using the quadratic form \(\bx^T\bA\bx\), the result is always a positive number.
}

\addThm{SPD Matrices Induce a Metric}{%
Let $\bA \in \fR^{n \times n}$ be an SPD matrix. The function
\[
    d_\bA(\bx, \by) = \sqrt{(\bx - \by)^T \bA (\bx - \by)}
\]
defines a metric on $\fR^n$, satisfying non-negativity, identity of indiscernibles, symmetry, and the triangle inequality.
}

\addPrf{%
Non-negativity and identity of indiscernibles follow directly from the positive-definite property: $(\bx-\by)^T\bA(\bx-\by) \geq 0$ with equality iff $\bx - \by = \bzero$. Symmetry holds since $(\bx-\by)^T\bA(\bx-\by) = (\by-\bx)^T\bA(\by-\bx)$. The triangle inequality follows from the Cauchy-Schwarz inequality applied to the inner product $\langle \bu, \bv\rangle_\bA = \bu^T\bA\bv$.
}

\addNote{%
When $\bA = \bI$, $d_\bA$ reduces to the standard Euclidean distance. The general $\bA$-metric stretches space according to the eigenvectors and eigenvalues of $\bA$: directions with large eigenvalues are "expensive" to move in.
}

\addExcr{%
\begin{enumerate}
    \item For $\bA = \begin{pmatrix} 4 & 0 \\ 0 & 1 \end{pmatrix}$, compute $d_\bA\!\left((1,0)^T,\,(0,0)^T\right)$ and $d_\bA\!\left((0,1)^T,\,(0,0)^T\right)$. Compare to the Euclidean distances. In what direction is the $\bA$-metric "larger"?
    \item Explain geometrically: the set $\{\bx : d_\bA(\bx, \bzero) = 1\}$ is an ellipse. What are its axes and their lengths in terms of the eigenvalues of $\bA$?
    \item Where does this metric appear in machine learning? (\textit{Hint: Mahalanobis distance.})
\end{enumerate}
}

\addDef{SPD Interpretation}{%
An SPD matrix transforms vector spaces by rescaling their components in a way that preserves a consistent notion of size. In standard Euclidean space, the norm \(\|\bx\|\) returns a positive length for any nonzero vector \(\bx\). Similarly, an SPD matrix \(\bA\) creates a new norm 
\[
\|\bx\|_{\bA} = \sqrt{\bx^T\bA\bx},
\]
which remains strictly positive for all nonzero \(\bx\). This property ensures that the transformation never collapses vectors to zero length or produces negative squared lengths, which is crucial for numerical stability in computational algorithms.
}


\addNote{%
Another way to visualize SPD matrices is through their geometric interpretation. When you use an SPD matrix \(\bA\) to compute distances via \(\bx^T\bA\bx\), the set of all vectors with constant "size" forms an ellipsoid (a stretched or squashed sphere). This is important in scientific computing applications like principal component analysis (PCA), where we often need to understand how data varies in different directions. The axes of this ellipsoid correspond to the eigenvectors of \(\bA\), and their lengths are determined by the eigenvalues.
}
\addDef{$\bA$-Inner Product}{%
For an SPD matrix $\bA \in \fR^{n \times n}$, the function
\[
    \langle \bx, \by \rangle_\bA = \bx^T\bA\by
\]
defines a valid inner product on $\fR^n$, inducing the norm $\|\bx\|_\bA = \sqrt{\bx^T\bA\bx}$. This is the natural inner product for problems where $\bA$ encodes the geometry (e.g., solving $\bA\bx = \bb$ via conjugate gradients).
}

\addThm{SPD Characterization via Eigenvalues}{%
A symmetric matrix $\bA \in \fR^{n \times n}$ is positive definite if and only if all its eigenvalues are strictly positive. (By the Spectral Theorem for real symmetric matrices, \S\ref{subsec:eigenvalues}, such a matrix is orthogonally diagonalizable with real eigenvalues.)
}

\addPrf{%
By the Spectral Theorem, $\bA = \bQ\bD\bQ^T$ where $\bQ$ is orthogonal and $\bD = \text{diag}(\lambda_1, \ldots, \lambda_n)$. For any nonzero $\bx$, let $\by = \bQ^T\bx$. Since $\bQ$ is invertible, $\by \neq \bzero$. Then:
\[
    \bx^T\bA\bx = \by^T\bD\by = \sum_{i=1}^n \lambda_i y_i^2.
\]
$(\Rightarrow)$: If all $\lambda_i > 0$ and $\by \neq \bzero$, then at least one $y_i \neq 0$, so the sum is $> 0$.

$(\Leftarrow)$: If some $\lambda_k \leq 0$, choose $\bx = \bQ\mathbf{e}_k$ (the $k$th column of $\bQ$), giving $\bx^T\bA\bx = \lambda_k \leq 0$.
}

\addCor{SPD Matrices are Invertible}{%
Every SPD matrix is invertible, since all eigenvalues are positive and thus $\det(\bA) = \prod_i \lambda_i > 0$.
}

\textBox{%
From a computational perspective, the condition \(\bx^T\bA\bx > 0\) for all nonzero \(\bx\) ensures that the matrix \(\bA\) never maps a vector to a zero result through the quadratic form, nor does it produce negative values. This is critical in numerical algorithms because it guarantees the stability and uniqueness of solutions. For example, when solving linear systems like \(\bA\bx = \bb\) where \(\bA\) is SPD, efficient methods like Cholesky decomposition can be used, and the solution is guaranteed to exist and be unique.
}

\addNote{%
SPD matrices are fundamental in scientific computing applications. For example:
\begin{itemize}
    \item In machine learning, covariance matrices are SPD and used to understand data distributions
    \item In numerical optimization, the Hessian matrix of a convex function is SPD near the minimum
    \item In finite element methods, the stiffness matrix is typically SPD
    \item In numerical linear algebra, SPD matrices allow for specialized algorithms like Cholesky factorization, which is twice as efficient as LU decomposition
    \item In image processing, SPD matrices can represent local structure tensors that capture directional information
\end{itemize}
Their computational properties make algorithms more stable, efficient, and reliable.
}

\subsubsection{Cholesky Decomposition}

\addDef{Cholesky Decomposition}{%
For a symmetric positive definite (SPD) matrix \(\bA \in \mathbb{R}^{n \times n}\), the Cholesky Decomposition expresses \(\bA\) as 
\[
\bA = \bL\bL^T,
\]
where \(\bL\) is a unique lower triangular matrix with strictly positive diagonal entries.
}

\textBox{%
SPD matrices, which satisfy \(\bx^T\bA\bx > 0\) for all nonzero \(\bx\), are prevalent in optimization, statistics, and finite element methods. The Cholesky factorization is both computationally efficient and numerically stable since it avoids the need for pivoting.
}

\addDef{Algorithm for Cholesky Decomposition}{%
For an SPD matrix \(\bA \in \mathbb{R}^{n \times n}\), the Cholesky factorization \(\bA = \bL\bL^T\) can be computed entry by entry as follows:

For each entry \(l_{ij}\) of \(\bL\) where \(1 \leq i,j \leq n\):
\begin{align}
l_{ij} &= 
\begin{cases}
0, & \text{if } i < j \\
\sqrt{a_{ii} - \sum_{k=1}^{i-1} l_{ik}^2}, & \text{if } i = j \\
\frac{1}{l_{jj}} \left( a_{ij} - \sum_{k=1}^{j-1} l_{ik}l_{jk} \right), & \text{if } i > j
\end{cases}
\end{align}

This algorithm proceeds row by row, using previously computed elements of \(\bL\) to determine subsequent entries.
}

\textBox{%
\textbf{Derivation of the Cholesky Algorithm (Matrix Form)}

To derive the Cholesky algorithm, we start with the relationship $\bA = \bL\bL^T$ where $\bA$ is an SPD matrix and $\bL$ is lower triangular. Let's express both matrices explicitly:

$$
\bA = \begin{pmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{n1} & a_{n2} & \cdots & a_{nn}
\end{pmatrix}, \quad
\bL = \begin{pmatrix}
l_{11} & 0 & \cdots & 0 \\
l_{21} & l_{22} & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
l_{n1} & l_{n2} & \cdots & l_{nn}
\end{pmatrix}
$$

The transpose of $\bL$ is:

$$
\bL^T = \begin{pmatrix}
l_{11} & l_{21} & \cdots & l_{n1} \\
0 & l_{22} & \cdots & l_{n2} \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & l_{nn}
\end{pmatrix}
$$

Now, the matrix product $\bL\bL^T$ gives us:

$$
\bL\bL^T = \begin{pmatrix}
l_{11}^2 & l_{11}l_{21} & \cdots & l_{11}l_{n1} \\
l_{21}l_{11} & l_{21}^2 + l_{22}^2 & \cdots & l_{21}l_{n1} + l_{22}l_{n2} \\
\vdots & \vdots & \ddots & \vdots \\
l_{n1}l_{11} & l_{n1}l_{21} + l_{n2}l_{22} & \cdots & \sum_{k=1}^{n} l_{nk}^2
\end{pmatrix}
$$

Equating this with $\bA$, we can derive the formulas for different elements of $\bL$:

\textbf{Case 1: Diagonal elements (position $(i,i)$)}

From the diagonal entries of $\bA = \bL\bL^T$, we have:
$$
a_{ii} = \sum_{k=1}^{i} l_{ik}^2 = l_{i1}^2 + l_{i2}^2 + \cdots + l_{ii}^2
$$

Solving for $l_{ii}$:
$$
l_{ii} = \sqrt{a_{ii} - \sum_{k=1}^{i-1} l_{ik}^2}
$$

\textbf{Case 2: Below diagonal (position $(i,j)$ where $i > j$)}

From the below-diagonal entries of $\bA = \bL\bL^T$, we have:
$$
a_{ij} = \sum_{k=1}^{j} l_{ik}l_{jk} = l_{i1}l_{j1} + l_{i2}l_{j2} + \cdots + l_{ij}l_{jj}
$$

Since $l_{jk} = 0$ for $k > j$ (as $\bL$ is lower triangular), the sum stops at $k = j$. 
Solving for $l_{ij}$:
$$
l_{ij} = \frac{1}{l_{jj}}\left(a_{ij} - \sum_{k=1}^{j-1} l_{ik}l_{jk}\right)
$$

\textbf{Case 3: Above diagonal (position $(i,j)$ where $i < j$)}

Since $\bA$ is symmetric, $a_{ij} = a_{ji}$. The elements of $\bL$ above the diagonal are defined to be zero: $l_{ij} = 0$ for $i < j$.

This gives us the complete algorithm for computing the Cholesky decomposition:
$$
l_{ij} = 
\begin{cases}
0, & \text{if } i < j \\
\sqrt{a_{ii} - \sum_{k=1}^{i-1} l_{ik}^2}, & \text{if } i = j \\
\frac{1}{l_{jj}} \left( a_{ij} - \sum_{k=1}^{j-1} l_{ik}l_{jk} \right), & \text{if } i > j
\end{cases}
$$

The algorithm proceeds by computing $\bL$ one column at a time, from left to right, and within each column from top to bottom.
}

\addDef{Computational Complexity (Cholesky)}{%
The Cholesky decomposition algorithm requires:
\begin{itemize}
    \item Approximately \(\frac{1}{3}n^3\) floating-point operations (flops) for an \(n \times n\) matrix
    \item This is roughly half the cost of LU factorization (\(\frac{2}{3}n^3\) flops), which is the standard method for non-symmetric matrices
    \item Storage requirement of \(\frac{n(n+1)}{2}\) elements (only the lower triangular part is stored)
\end{itemize}
The efficiency comes from exploiting both the symmetry of \(\bA\) and its positive definiteness, which eliminates the need for pivoting strategies.
}

\addExmpl{%
Let's compute the Cholesky decomposition of:
\[
\bA = \begin{pmatrix} 4 & 2 \\ 2 & 3 \end{pmatrix}
\]

Step 1: Compute the first column of \(\bL\).
\begin{align}
l_{11} &= \sqrt{a_{11}} = \sqrt{4} = 2 \\
l_{21} &= \frac{a_{21}}{l_{11}} = \frac{2}{2} = 1
\end{align}

Step 2: Compute the second column of \(\bL\).
\begin{align}
l_{22} &= \sqrt{a_{22} - l_{21}^2} = \sqrt{3 - 1^2} = \sqrt{2} \approx 1.414
\end{align}

Therefore:
\[
\bL = \begin{pmatrix} 2 & 0 \\ 1 & \sqrt{2} \end{pmatrix}
\]

Verification: 
\[
\bL\bL^T = \begin{pmatrix} 2 & 0 \\ 1 & \sqrt{2} \end{pmatrix} \begin{pmatrix} 2 & 1 \\ 0 & \sqrt{2} \end{pmatrix} = \begin{pmatrix} 4 & 2 \\ 2 & 3 \end{pmatrix} = \bA
\]
}

\addNote{%
The uniqueness of the Cholesky Decomposition stems from the positivity of the leading principal minors of an SPD matrix, ensuring that each pivot in the elimination process is nonzero. For an SPD matrix \(\bA\), the leading principal minor of order \(k\), denoted as \(\det(\bA_k)\), is the determinant of the submatrix formed by the first \(k\) rows and columns. These determinants are all positive, which ensures that the diagonal entries of \(\bL\) are well-defined and unique.
}

\addThm{Cholesky Existence and Uniqueness}{%
A matrix $\bA \in \fR^{n \times n}$ admits a Cholesky factorization $\bA = \bL\bL^T$ (with $\bL$ lower triangular and strictly positive diagonal) if and only if $\bA$ is SPD. The factorization is unique.
}

\addPrf{%
$(\Rightarrow)$: Suppose $\bA = \bL\bL^T$. Then $\bA^T = (\bL\bL^T)^T = \bL\bL^T = \bA$, so $\bA$ is symmetric. For any nonzero $\bx$:
\[
    \bx^T\bA\bx = \bx^T\bL\bL^T\bx = \|\bL^T\bx\|_2^2 \geq 0.
\]
Equality holds iff $\bL^T\bx = \bzero$. Since the diagonal of $\bL$ is strictly positive, $\bL$ is invertible, so $\bx = \bzero$. Hence $\bA$ is SPD.

$(\Leftarrow)$: If $\bA$ is SPD, all leading principal minors are positive (each submatrix is also SPD), so the Cholesky algorithm never encounters a zero or negative under a square root.

Uniqueness: If $\bL_1\bL_1^T = \bL_2\bL_2^T$, set $\bM = \bL_1^{-1}\bL_2$. Then $\bM\bM^T = \bI$, so $\bM$ is lower triangular and orthogonal. The only lower triangular orthogonal matrix with positive diagonal is $\bI$, so $\bL_1 = \bL_2$.
}

\addThm{Cholesky Factorization Cost}{%
The Cholesky factorization of an SPD matrix $\bA \in \fR^{n \times n}$ costs $\frac{1}{3}n^3 + O(n^2)$ flops — exactly half the cost of LU factorization — and requires storing only $\frac{n(n+1)}{2}$ entries.
}

\addPrf{%
At step $j$ of the algorithm, we compute $l_{jj}$ (1 square root) and $n - j$ below-diagonal entries $l_{ij}$ for $i > j$, each requiring $j$ multiplications and 1 division. The dominant cost is:
\[
    \sum_{j=1}^{n} (n-j) \cdot j \approx \int_0^n (n-t)\,t\,dt = \frac{n^3}{6} + \frac{n^3}{6} = \frac{n^3}{3}.
\]
}

\addExcr{%
\begin{enumerate}
    \item Which of the following matrices are SPD? For each, either prove it or find a nonzero $\bx$ with $\bx^T\bA\bx \leq 0$.
    \[
        \bA_1 = \begin{pmatrix} 2 & 1 \\ 1 & 2 \end{pmatrix}, \quad
        \bA_2 = \begin{pmatrix} 1 & 2 \\ 2 & 1 \end{pmatrix}, \quad
        \bA_3 = \begin{pmatrix} 4 & 2 & 0 \\ 2 & 3 & 1 \\ 0 & 1 & 1 \end{pmatrix}.
    \]
    \item Compute the Cholesky factorization of $\bA_1$ and $\bA_3$ by hand. Verify $\bL\bL^T = \bA$.
    \item Use the Cholesky factorization of $\bA_1$ to solve $\bA_1\bx = (3, 4)^T$ via forward then backward substitution.
    \item In NumPy, \texttt{np.linalg.cholesky(A)} raises \texttt{LinAlgError} if $\bA$ is not SPD. Write code to attempt Cholesky on each matrix above and handle the exception. What does a \texttt{LinAlgError} tell you about the matrix?
\end{enumerate}
}

\addExcr{%
\textbf{(Condition Number of SPD Matrices)} For an SPD matrix with eigenvalues $\lambda_1 \leq \lambda_2 \leq \cdots \leq \lambda_n$, the 2-norm condition number is $\kappa_2(\bA) = \lambda_n / \lambda_1$.
\begin{enumerate}
    \item Prove this formula. (\textit{Hint: $\|\bA\|_2 = \max_i |\lambda_i|$ for symmetric $\bA$; and $\bA^{-1}$ has eigenvalues $1/\lambda_i$.})
    \item For $\bA = \begin{pmatrix} 10 & 0 \\ 0 & 0.01 \end{pmatrix}$, what is $\kappa_2(\bA)$? What does this imply for solving $\bA\bx = \bb$?
    \item The Cholesky factor $\bL$ satisfies $\kappa_2(\bL) = \sqrt{\kappa_2(\bA)}$. Why is this numerically significant? (\textit{Think: solving $\bL\by = \bc$ and $\bL^T\bx = \by$ each introduces what error?})
    \item In NumPy, compute \texttt{np.linalg.cond(A)} for an SPD matrix and verify it matches $\lambda_{\max}/\lambda_{\min}$ computed from \texttt{np.linalg.eigvalsh(A)}.
\end{enumerate}
}




\end{document}