\documentclass[../readings.tex]{subfiles}
\begin{document}
\subsection{Stability and Condition Number}
\label{subsec:stability-condition}

\textBox{%
In an idealized mathematical setting, the solution to $\bA\bx = \bb$ is simply
$\bx = \bA^{-1}\bb$. In practice, two separate issues corrupt the computed answer:
the \emph{inherent sensitivity} of the problem to perturbations in the data, measured
by the condition number, and the \emph{algorithm's ability to avoid amplifying errors}
beyond this inherent sensitivity, measured by numerical stability. Understanding both
is essential for reliable scientific computing. This section develops the theory
behind each and shows how they interact in the matrix factorizations studied so far.
}

\subsubsection{Why Not Invert Directly?}

\textBox{%
The formula $\bx = \bA^{-1}\bb$ is mathematically correct but computationally
inadvisable. Computing the full inverse $\bA^{-1}$ is unnecessary for solving a
single system, wastes memory, and is less numerically stable than factorization-based
approaches. The arguments below quantify exactly what is lost.
}

\addNote{%
\textbf{Cost of explicit inversion.}
Computing $\bA^{-1}$ requires solving $n$ systems $\bA\mathbf{x}_i = \be_i$. After
the LU factorization (cost $\tfrac{2}{3}n^3$), each solve costs $2n^2$, giving a
total of $\tfrac{2}{3}n^3 + 2n^3 \approx \tfrac{8}{3}n^3$ flops. Compare this to
solving $\bA\bx = \bb$ for a single $\bb$: factorize once at $\tfrac{2}{3}n^3$, then
one triangular solve at $2n^2$. Explicitly forming $\bA^{-1}$ costs four times as much
as necessary.
}

\addNote{%
\textbf{Stability of explicit inversion.}
When $\bA$ is nearly singular, explicitly forming $\bA^{-1}$ amplifies round-off
errors severely. A single erroneous column of $\bA^{-1}$ propagates into the entire
product $\bA^{-1}\bb$, whereas solving via a stable factorization confines errors
to individual triangular solve steps. The practical rule is: never explicitly form
$\bA^{-1}$. Use \texttt{np.linalg.solve(A, b)} for a single system, or
\texttt{scipy.linalg.lu\_factor} and \texttt{scipy.linalg.lu\_solve} for multiple
right-hand sides.
}

\addNote{%
\textbf{Factorize once, solve many.}
Given the LU factorization $\bA = \bP\bL\bU$ computed once at cost $\tfrac{2}{3}n^3$,
each new right-hand side $\bb$ requires only two triangular solves at cost $2n^2$.
For $k$ right-hand sides the total is $\tfrac{2}{3}n^3 + 2kn^2$, far better than
$k$ separate $O(n^3)$ solves or storing and applying $\bA^{-1}$.
}

\addExcr{%
\begin{enumerate}
  \item For $n = 500$, estimate the flop counts for (a) computing $\bA^{-1}$ and
  then applying it to 10 right-hand sides, versus (b) computing the LU factorization
  once and solving 10 triangular systems. Which is faster, and by what factor?
  \item In Python, time \texttt{np.linalg.inv(A) @ b} vs.\
  \texttt{np.linalg.solve(A, b)} for a random $500 \times 500$ matrix over 20 solves.
  Which is faster? Does the timing match your flop count estimate?
  \item Construct a nearly singular matrix \texttt{A = np.diag([1, 1, 1e-12])} and
  a right-hand side \texttt{b = [1, 1, 1]}. Solve via \texttt{np.linalg.inv(A) @ b}
  and \texttt{np.linalg.solve(A, b)}. Compare the residuals $\|\bA\hat{\bx} - \bb\|$.
\end{enumerate}
}

\subsubsection{Numerical Stability}

\textBox{%
Even after choosing a factorization-based approach, there are good and bad ways to
implement the factorization. The concept of backward stability gives a precise
criterion for what ``good'' means: a backward stable algorithm is one whose computed
output is the exact output for a slightly perturbed input. This is the best one can
hope for in finite-precision arithmetic.
}

\addDef{Backward Stability}{%
An algorithm for computing $f(\bx)$ is \emph{backward stable} if for every input
$\bx$, the computed output $\hat{f}(\bx)$ satisfies
\[
  \hat{f}(\bx) = f(\bx + \delta\bx)
\]
for some perturbation $\|\delta\bx\| / \|\bx\| = O(\varepsilon_{\mathrm{mach}})$.
Applied to $\bA\bx = \bb$: an algorithm is backward stable if the computed solution
$\hat{\bx}$ satisfies $(\bA + \delta\bA)\hat{\bx} = \bb + \delta\bb$ with
$\|\delta\bA\|/\|\bA\| = O(\varepsilon_{\mathrm{mach}})$ and
$\|\delta\bb\|/\|\bb\| = O(\varepsilon_{\mathrm{mach}})$.
}
\label{def:backward-stable}

\addNote{%
Backward stability is a property of the \emph{algorithm}, not the problem.
Gaussian elimination without pivoting can be unstable; with partial pivoting it is
backward stable for virtually all practical matrices.
}

\addExcr{%
Refer to Def~\ref{def:backward-stable}.
\begin{enumerate}
  \item Explain in your own words the difference between a backward stable algorithm
  and one that simply produces a small residual $\|\bA\hat{\bx} - \bb\|$. Are the
  two equivalent? (\textit{Hint: think about the relationship between backward error
  and residual.})
  \item Consider computing $e^x$ for large negative $x$ directly versus via
  $1/e^{-x}$. Which is backward stable? (\textit{Hint: evaluate both at
  $x = -30$ in Python and compare to the true value.})
\end{enumerate}
}

\addProp{Gaussian Elimination with Partial Pivoting is Backward Stable}{%
Gaussian elimination with partial pivoting (GEPP) produces a computed solution
$\hat{\bx}$ satisfying $(\bA + \delta\bA)\hat{\bx} = \bb$ with
\[
  \|\delta\bA\|_\infty \leq g_n \cdot \varepsilon_{\mathrm{mach}} \cdot \|\bA\|_\infty,
\]
where $g_n$ is the \emph{growth factor}. In the worst case $g_n \leq 2^{n-1}$, but
for random matrices in practice $g_n = O(n^{2/3})$.
}
\label{prop:gepp-stable}

\addPrf{%
At each elimination step, partial pivoting selects the largest entry in the current
column as pivot, ensuring all multipliers satisfy $|m_{ij}| \leq 1$. This prevents
exponential amplification of round-off at each stage. A detailed rounding error
analysis over the $O(n^3)$ scalar operations shows the accumulated backward error
satisfies the stated bound. For a complete proof see Higham, \textit{Accuracy and
Stability of Numerical Algorithms}, Ch.~9.
}

\addExcr{%
Refer to Prop~\ref{prop:gepp-stable}.
\begin{enumerate}
  \item The growth factor for a random $n \times n$ matrix is typically much smaller
  than $2^{n-1}$. In Python, construct 100 random $50 \times 50$ matrices and compute
  the LU factorization via \texttt{scipy.linalg.lu}. For each, compute the growth
  factor $g_n = \|\bU\|_\infty / \|\bA\|_\infty$. Plot a histogram. Does $g_n$
  stay small in practice?
  \item Construct the $n \times n$ Kahan matrix (a known worst-case for GEPP) and
  verify the growth factor can be large. Use \texttt{scipy.linalg.khatri\_rao} or
  build it manually.
\end{enumerate}
}

\subsubsection{Condition Number}

\textBox{%
Backward stability tells us about the algorithm; the condition number tells us about
the problem. Even a perfectly backward stable algorithm cannot produce an accurate
solution when the problem itself is ill-conditioned: small perturbations in the input
lead to large changes in the output. The condition number quantifies exactly how bad
this amplification can be.
}

\addDef{Condition Number}{%
The \emph{condition number} of an invertible matrix $\bA \in \fR^{n\times n}$ with
respect to a norm $\|\cdot\|$ is
\[
  \kappa(\bA) = \|\bA\|\,\|\bA^{-1}\|.
\]
For the 2-norm, $\kappa_2(\bA) = \sigma_{\max}/\sigma_{\min}$ (ratio of largest to
smallest singular values). For a symmetric matrix, $\kappa_2(\bA) = |\lambda_{\max}|/|\lambda_{\min}|$.
}
\label{def:condition-number}

\addExcr{%
Refer to Def~\ref{def:condition-number}.
\begin{enumerate}
  \item Compute $\kappa_2(\bA)$ for $\bA = \begin{pmatrix}3&1\\1&3\end{pmatrix}$ by
  hand using the singular values. Verify with \texttt{np.linalg.cond(A)}.
  \item For $\bA = \mathrm{diag}(1, 10^{-3}, 10^{-6})$, compute $\kappa_2(\bA)$,
  $\kappa_1(\bA)$, and $\kappa_\infty(\bA)$. How do they compare?
  \item The Hilbert matrix $H_{ij} = 1/(i+j-1)$ is notoriously ill-conditioned.
  Compute $\kappa_2(H_n)$ for $n = 3, 5, 8, 10$ using \texttt{scipy.linalg.hilbert}.
  How fast does the condition number grow with $n$?
\end{enumerate}
}

\addThm{Perturbation Bound for Linear Systems}{%
Let $\bA\bx = \bb$ with $\bA$ invertible and $\bb \neq \bzero$.
If $\hat{\bx}$ satisfies $(\bA + \delta\bA)\hat{\bx} = \bb + \delta\bb$ and
$\|\bA^{-1}\|\,\|\delta\bA\| < 1$, then
\[
  \frac{\|\hat{\bx} - \bx\|}{\|\bx\|}
  \leq \frac{\kappa(\bA)}{1 - \kappa(\bA)\,\|\delta\bA\|/\|\bA\|}
  \left(\frac{\|\delta\bA\|}{\|\bA\|} + \frac{\|\delta\bb\|}{\|\bb\|}\right).
\]
}
\label{thm:perturbation-bound}

\addPrf{%
\textbf{Step 1.}
Subtract $\bA\bx = \bb$ from $(\bA + \delta\bA)\hat{\bx} = \bb + \delta\bb$:
\[
  (\bA + \delta\bA)(\hat{\bx} - \bx) = \delta\bb - \delta\bA\,\bx.
\]

\textbf{Step 2.}
Since $\|\bA^{-1}\delta\bA\| \leq \|\bA^{-1}\|\,\|\delta\bA\| < 1$, the Neumann
series gives $(\bI + \bA^{-1}\delta\bA)^{-1} = \sum_{k=0}^\infty (-\bA^{-1}\delta\bA)^k$,
so $\bA + \delta\bA$ is invertible and
\[
  \|(\bA + \delta\bA)^{-1}\| \leq \frac{\|\bA^{-1}\|}{1 - \|\bA^{-1}\|\,\|\delta\bA\|}.
\]

\textbf{Step 3.}
Applying this bound to the error from Step 1:
\[
  \|\hat{\bx} - \bx\|
  \leq \frac{\|\bA^{-1}\|}{1 - \|\bA^{-1}\|\,\|\delta\bA\|}
  \bigl(\|\delta\bb\| + \|\delta\bA\|\,\|\bx\|\bigr).
\]
Divide by $\|\bx\|$ and use $\|\bb\| \leq \|\bA\|\,\|\bx\|$ to convert
$\|\bx\|$ terms to $\|\bb\|$ terms. This yields the stated bound.
}

\addCor{Perturbation in the Right-Hand Side Only}{%
If only $\bb$ is perturbed ($\delta\bA = \bzero$), then
\[
  \frac{\|\hat{\bx} - \bx\|}{\|\bx\|} \leq \kappa(\bA)\,\frac{\|\delta\bb\|}{\|\bb\|}.
\]
}
\label{cor:rhs-perturbation}

\addPrf{%
Set $\delta\bA = \bzero$ in Thm~\ref{thm:perturbation-bound}. Then
$\hat{\bx} - \bx = \bA^{-1}\delta\bb$, so $\|\hat{\bx}-\bx\| \leq \|\bA^{-1}\|\,\|\delta\bb\|$.
Dividing by $\|\bx\| \geq \|\bb\|/\|\bA\|$ gives $\kappa(\bA)\,\|\delta\bb\|/\|\bb\|$.
}

\addNote{%
The condition number is the worst-case amplification factor: a relative perturbation
of size $\varepsilon$ in the data can produce a relative error up to $\kappa(\bA)\cdot\varepsilon$
in the solution. In double precision $\varepsilon_{\mathrm{mach}} \approx 10^{-16}$, so
we can expect at most $\lfloor 16 - \log_{10}\kappa(\bA)\rfloor$ correct decimal digits
in the computed solution.
}

\addExmpl{%
Let $\bA = \begin{pmatrix} 1 & 0 \\ 0 & 10^{-6} \end{pmatrix}$, $\bb = (1,1)^T$.
The exact solution is $\bx^* = (1, 10^6)^T$ and $\kappa_2(\bA) = 10^6$.
Perturbing $\bb$ to $\tilde{\bb} = (1 + 10^{-4}, 1)^T$ has
$\|\delta\bb\|/\|\bb\| \approx 10^{-4}$, and the bound predicts relative error up to
$10^6 \cdot 10^{-4} = 100$. The actual perturbed solution is $(1+10^{-4}, 10^6)^T$,
giving relative error $\approx 10^{-10}$: much smaller than the bound because the
perturbation was in the well-conditioned direction. The bound is worst-case.
}

\addExcr{%
Refer to Thm~\ref{thm:perturbation-bound} and Cor~\ref{cor:rhs-perturbation}.
\begin{enumerate}
  \item For $\bA_\epsilon = \begin{pmatrix}1 & 0 \\ 0 & \epsilon\end{pmatrix}$
  and $\bb = (1,1)^T$, compute $\kappa_2(\bA_\epsilon)$ for
  $\epsilon = 10^{-4}, 10^{-8}, 10^{-12}$.
  Perturb $\bb$ to $\tilde{\bb} = (1+10^{-4},1)^T$, solve both systems, and compare
  the actual relative error to the bound $\kappa_2(\bA_\epsilon)\cdot\|\delta\bb\|/\|\bb\|$.
  \item At which $\epsilon$ does \texttt{np.linalg.solve} give an unreliable answer
  in double precision? Explain in terms of the number of correct digits formula.
  \item Show that for a symmetric matrix $\bA$ with eigenvalues $\lambda_1, \ldots, \lambda_n$,
  $\kappa_2(\bA) = |\lambda_{\max}|/|\lambda_{\min}|$.
  (\textit{Hint: use the Spectral Theorem and the formula $\kappa_2 = \sigma_{\max}/\sigma_{\min}$.})
\end{enumerate}
}

\subsubsection{Conditioning vs.\ Stability}

\textBox{%
Conditioning and stability are often conflated but are fundamentally different.
Conditioning is a property of the mathematical problem; stability is a property of the
algorithm. The two interact through the forward error bound: for a backward stable
algorithm applied to a problem with condition number $\kappa$, the forward error is
at most $O(\kappa \cdot \varepsilon_{\mathrm{mach}})$. Neither property alone is
sufficient: good conditioning with an unstable algorithm, or bad conditioning with
a stable one, both lead to inaccurate results.
}

\addDef{Conditioning}{%
\emph{Conditioning} is an intrinsic property of the mathematical problem. It measures
the sensitivity of the exact solution to perturbations in the data. A matrix is
\emph{ill-conditioned} if $\kappa(\bA) \gg 1$ and \emph{well-conditioned} if $\kappa(\bA)$
is modest. Conditioning cannot be improved by choosing a better algorithm; it can only
be improved by reformulating the problem or preconditioning.
}
\label{def:conditioning}

\addDef{Numerical Stability}{%
\emph{Numerical stability} is a property of the algorithm. It measures how well the
algorithm avoids introducing errors beyond those forced by the problem's conditioning.
A backward stable algorithm introduces no more error than is intrinsic to the problem.
}
\label{def:numerical-stability}

\addNote{%
A well-conditioned problem solved by an unstable algorithm may yield inaccurate results.
An ill-conditioned problem solved by a stable algorithm is still inherently sensitive:
no algorithm can overcome poor conditioning. The best strategy is to use a numerically
stable algorithm, and if $\kappa(\bA)$ is large, reformulate or precondition the problem.
}

\addExcr{%
Refer to Defs~\ref{def:conditioning} and~\ref{def:numerical-stability}.
\begin{enumerate}
  \item Give an example of a well-conditioned problem solved inaccurately due to an
  unstable algorithm. (\textit{Hint: evaluate $e^x - 1$ for small $x$ in Python
  directly vs.\ via \texttt{np.expm1(x)}.})
  \item The Hilbert matrix $H_n$ is ill-conditioned. Compute $\kappa_2(H_{12})$ and
  estimate how many correct digits to expect from double-precision arithmetic.
  Verify by solving $H_{12}\bx = H_{12}\cdot\mathbf{1}$ and checking the error.
  \item Can you make an ill-conditioned problem well-conditioned by scaling? Construct
  a diagonal scaling matrix $\bD = \mathrm{diag}(d_1, \ldots, d_n)$ such that
  $\kappa_2(\bD\bA\bD)$ is minimized, and apply it to a $3\times 3$ example.
\end{enumerate}
}

\subsubsection{Stability of Matrix Factorizations}

\textBox{%
The three main factorizations studied in this course have different stability properties,
and the right choice depends on the structure of the problem.
}

\addProp{Stability of Cholesky Factorization}{%
For symmetric positive definite $\bA$, Cholesky factorization $\bA = \bL\bL^T$ is
backward stable without pivoting. All diagonal entries $\ell_{kk}$ remain positive and
bounded by $\sqrt{\|\bA\|_2}$, so no element growth occurs.
}
\label{prop:cholesky-stable}

\addPrf{%
For an SPD matrix, all principal submatrices are SPD, so each diagonal entry
encountered during Cholesky is positive and the square root is well-defined.
Since no cancellation or sign change can occur, the factorization is inherently stable.
A rounding error analysis shows the backward error satisfies
$\|\bA + \delta\bA - \hat{\bL}\hat{\bL}^T\|_F \leq c_n\varepsilon_{\mathrm{mach}}\|\bA\|_F$
for a modest constant $c_n$.
}

\addProp{Stability of QR via Householder Reflections}{%
QR decomposition using Householder reflections is backward stable with no growth
factor. Each Householder reflector $\bH = \bI - 2\bv\bv^T/\|\bv\|_2^2$ is orthogonal
with $\kappa_2(\bH) = 1$, so orthogonal transformations cannot amplify errors.
}
\label{prop:qr-stable}

\addPrf{%
Each Householder reflection is orthogonal, so $\|\bH\|_2 = \|\bH^{-1}\|_2 = 1$.
Round-off errors at each step are not amplified by subsequent orthogonal transformations.
A detailed analysis shows the overall backward error satisfies
$\|(\hat{\bQ} + \delta\bQ)(\hat{\bR} + \delta\bR) - \bA\|_F = O(\varepsilon_{\mathrm{mach}}\|\bA\|_F)$.
}

\addNote{%
\textbf{Hierarchy of stability.}
SVD $\geq$ QR (Householder) $\geq$ Cholesky (SPD only) $\approx$ LU with partial
pivoting $\gg$ LU without pivoting. SVD is the most stable but most expensive. For
general systems, GEPP is standard; for SPD systems, Cholesky is preferred; for least
squares or ill-conditioned problems, QR is recommended.
}

\addDef{Iterative Refinement}{%
After computing $\hat{\bx}$ via a factorization, \emph{iterative refinement} improves
accuracy by three steps: (1) compute the residual $\br = \bb - \bA\hat{\bx}$; (2) solve
$\bA\boldsymbol{\delta} = \br$ using the same factorization; (3) update
$\hat{\bx} \leftarrow \hat{\bx} + \boldsymbol{\delta}$. Each cycle recovers roughly
$\lfloor\log_{10}(1/\kappa(\bA))\rfloor$ additional correct digits, up to the precision
limit.
}
\label{def:iterative-refinement}

\addExcr{%
Refer to Def~\ref{def:iterative-refinement}.
\begin{enumerate}
  \item Implement one step of iterative refinement for $H_{10}\bx = H_{10}\cdot\mathbf{1}$
  (Hilbert matrix, $n=10$). Compare the error before and after refinement.
  How many digits did you recover?
  \item Apply two steps of refinement. Does a second step help? Is there a limit
  to how many digits can be recovered?
\end{enumerate}
}

\addExcr{%
\textbf{(Comparing factorization stability.)}
Build the $n \times n$ Hilbert matrix via \texttt{scipy.linalg.hilbert(n)}.
\begin{enumerate}
  \item For $n = 10$: compute $\kappa_2(H)$ and predict how many correct digits to
  expect. Solve $H\bx = H\cdot\mathbf{1}$ via \texttt{np.linalg.solve} (LU) and
  \texttt{np.linalg.lstsq} (QR-based). Compare the errors
  $\|\hat{\bx} - \mathbf{1}\|_2$.
  \item Use \texttt{mpmath} with \texttt{mp.dps = 50} to solve the same system in
  high precision. Does the error decrease significantly compared to double precision?
  What does this tell you about conditioning vs.\ precision?
  \item Apply one step of iterative refinement to the LU solution. How much
  improvement do you observe?
\end{enumerate}
}

\addExcr{%
\textbf{(Backward error analysis in practice.)}
For $\bA\bx = \bb$ with a random $10 \times 10$ matrix scaled so $\kappa(\bA) \approx 10^4$:
\begin{enumerate}
  \item Compute $\hat{\bx}$ via \texttt{np.linalg.solve}. Compute the backward error
  $\|\bA\hat{\bx} - \bb\|_2/\|\bb\|_2$. Is it consistent with $O(\varepsilon_{\mathrm{mach}})$?
  \item Compute the forward error $\|\hat{\bx} - \bx^*\|_2/\|\bx^*\|_2$ (using the
  true solution). Is the ratio (forward error)/(backward error) consistent with $\kappa(\bA)$?
  \item Repeat with $\kappa \approx 10^{12}$ and interpret the results.
\end{enumerate}
}

\end{document}
