\documentclass[../readings.tex]{subfiles}
\begin{document}


\subsection{Vectors and Matrices}

Scientific computing is built on two fundamental data structures: vectors and matrices.
Before introducing these formally, we fix notation for the standard number systems that
provide the raw material for all numerical computation.

\textBox{%
\textbf{Standard Number Systems.}\enspace Throughout these notes, $\mathbb{R}$ denotes
the set of all real numbers, $\mathbb{Z}$ the set of all integers, and $\mathbb{C}$ the
set of all complex numbers.  Our primary focus is real-valued computation; complex numbers
arise naturally in eigenvalue analysis and signal processing.
}

\addDef{Scalar}{%
An element of a number system ($\mathbb{R}$, $\mathbb{Z}$, or $\mathbb{C}$) is called a
\textbf{scalar}.  Scalars represent individual numerical quantities and serve as the
atomic building blocks from which vectors and matrices are constructed.
}
\label{def:scalar}

\addExcr{%
Refer to Def~\ref{def:scalar}.
\begin{enumerate}
    \item Classify each of the following as a real scalar, an integer, a complex (non-real)
    scalar, or \emph{not} a scalar: $\;3.14$;\; $-7$;\; $2+3i$;\; $\begin{pmatrix}1\\0\end{pmatrix}$.
    \item In Python, evaluate \texttt{type(3.14)}, \texttt{type(3)}, and \texttt{type(3+2j)}.
    Which NumPy \texttt{dtype} would you use to store each?
\end{enumerate}
}

\addDef{Vectors}{%
A \textbf{vector} is a one-dimensional ordered array of scalars.  Vectors can represent
points, directions, or more abstract quantities within a vector space, and are fundamental
in encoding data and linear relationships.
}
\label{def:vector}

\addNote{%
Vectors may be written as row vectors or column vectors.  Although both forms carry the
same information, the distinction matters when applying linear transformations and
computing inner products.
}

\addExmpl{%
\begin{equation*}
\begin{pmatrix} 1 & 2 & 3 \end{pmatrix}, \quad
\begin{pmatrix} 1 \\ 2 \\ 3 \end{pmatrix}, \quad
\begin{pmatrix} 1+2i \\ 7i \\ 3 \end{pmatrix}, \ldots
\end{equation*}
}

\addExcr{%
Refer to Def~\ref{def:vector}.
\begin{enumerate}
    \item Write a column vector $\bx \in \fR^4$ whose $i$-th entry equals $2^i$ for
    $i = 1, 2, 3, 4$.
    \item In NumPy, construct this vector with \texttt{np.array([...])} and confirm
    \texttt{x.shape == (4,)}.  Why does NumPy report shape \texttt{(4,)} rather than
    \texttt{(4,1)}?
    \item Can a vector in $\fR^3$ be added to a vector in $\fR^4$?  Justify in one sentence.
\end{enumerate}
}

\textBox{%
Just as scalars belong to number systems like $\mathbb{R}$ or $\mathbb{C}$, vectors inhabit
\textbf{vector spaces}.  While these notes treat vector spaces informally, subsequent
chapters provide more rigorous definitions and properties.
}

\addDef{$\fR^n$}{%
The set of all real vectors of size $n$.  When we write $\bx \in \fR^n$, it means $\bx$
has $n$ real entries.  Vectors in $\fR^n$ take one of the following forms:
\begin{align*}
\boldsymbol{x} \in \fR^{n \times 1} &= \begin{pmatrix} x_1 \\ x_2 \\ \vdots \\ x_n \end{pmatrix}
\quad \text{(column vector, default convention)}\\[1mm]
\boldsymbol{x} \in \fR^{1 \times n} &= \begin{pmatrix} x_1 & x_2 & \cdots & x_n \end{pmatrix}
\quad \text{(row vector)}
\end{align*}
}
\label{def:Rn}

\addNote{%
These notes focus on finite-dimensional spaces.  Infinite-dimensional vector spaces, central
to functional analysis, arise in differential equations and quantum mechanics but are beyond
our scope here.
}

\addExcr{%
Refer to Def~\ref{def:Rn}.
\begin{enumerate}
    \item Let $\bx = (1, -2, 3)^T \in \fR^3$.  Identify the value of entry $x_2$.  What
    is the dimension of $\fR^3$?
    \item In NumPy, create a column vector (shape \texttt{(3,1)}) from the list
    \texttt{[1, -2, 3]} in two ways: using \texttt{np.reshape} and using
    \texttt{[:, np.newaxis]}.
    \item What does \texttt{x.T} return when \texttt{x.shape == (3,)}?  How does the result
    differ when \texttt{x.shape == (3,1)}?
\end{enumerate}
}

\addDef{Matrices}{%
A \textbf{matrix} is a two-dimensional array of scalars.  A matrix $\bA \in \fR^{m \times n}$
is written as
\begin{equation*}
\bA = \begin{pmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{m1} & a_{m2} & \cdots & a_{mn}
\end{pmatrix}.
\end{equation*}
Matrices are used to represent linear transformations, systems of linear equations, and
more complex data structures.
}
\label{def:matrix}

\addExmpl{%
Two-dimensional matrices include:
\begin{equation*}
\begin{pmatrix}
1 & 2 & 3 \\[1mm]
4 & 5 & 6
\end{pmatrix} \in \fR^{2 \times 3}, \quad
\begin{pmatrix}
1 & 2 \\[1mm]
4 & 5 \\[1mm]
6 & 7
\end{pmatrix} \in \fR^{3 \times 2}.
\end{equation*}
}

\addNote{%
Matrices can be interpreted as vectors whose entries are themselves vectors:
\begin{equation*}
\bA = \printmatrow{r} = \printmatcol{c},
\end{equation*}
where $r_i = \bA[i,:]$ is the $i$-th row and $c_i = \bA[:,i]$ is the $i$-th column.
This viewpoint is useful when studying block matrices and matrix factorizations.
}

\addExcr{%
Refer to Def~\ref{def:matrix}.
\begin{enumerate}
    \item State the dimensions $(m, n)$ of $\bA = \begin{pmatrix}1 & 2 & 3 \\ 4 & 5 & 6\end{pmatrix}$.
    What is the shape of its second column as a vector?
    \item In NumPy (0-based indexing), write the expression to access entry $a_{12}$
    (first row, second column in 1-based notation).  Then extract the second column of $\bA$
    and verify its shape is \texttt{(2,)}.
    \item True or False: a vector in $\fR^n$ is a special case of a matrix in
    $\fR^{n \times 1}$.  How does NumPy handle this distinction in practice?
\end{enumerate}
}

\addDef{Matrix Transpose}{%
The \textbf{transpose} of $\bA \in \fR^{m \times n}$, denoted $\bA^T$, is the $n \times m$
matrix obtained by interchanging rows and columns:
\[
    (\bA^T)_{ij} = a_{ji}, \quad 1 \le i \le n,\ 1 \le j \le m.
\]
For a column vector $\bx \in \fR^n$, the transpose $\bx^T$ is its corresponding row vector
and vice versa.
}
\label{def:transpose}

\addDef{Symmetric and Skew-Symmetric Matrices}{%
A square matrix $\bA \in \fR^{n \times n}$ is:
\begin{itemize}
    \item \textbf{symmetric} if $\bA^T = \bA$, equivalently $a_{ij} = a_{ji}$ for all $i,j$.
    \item \textbf{skew-symmetric} if $\bA^T = -\bA$, equivalently $a_{ij} = -a_{ji}$ for all $i,j$.
\end{itemize}
}
\label{def:symmetric}

\addNote{%
Every square matrix $\bA$ decomposes uniquely into a symmetric part and a skew-symmetric part:
\[
    \bA = \underbrace{\tfrac{1}{2}(\bA + \bA^T)}_{\text{symmetric}} +
          \underbrace{\tfrac{1}{2}(\bA - \bA^T)}_{\text{skew-symmetric}}.
\]
Symmetric matrices arise constantly in scientific computing: covariance matrices in statistics,
graph Laplacians in network problems, and Hessians of quadratic objectives in optimization
are all symmetric.
}

\addThm{Properties of the Transpose}{%
Let $\bA \in \fR^{m \times n}$, $\bB \in \fR^{n \times p}$, and $\alpha \in \fR$. Then:
\begin{enumerate}
    \item $(\bA^T)^T = \bA$
    \item $(\bA + \bB)^T = \bA^T + \bB^T$ \quad (for conformable $\bA, \bB$)
    \item $(\alpha\bA)^T = \alpha\bA^T$
    \item $(\bA\bB)^T = \bB^T\bA^T$ \quad \textbf{(Reversal Law)}
\end{enumerate}
}
\label{thm:transpose-props}

\addPrf{%
Properties (1)--(3) follow directly from the definition $(\bA^T)_{ij} = a_{ji}$ applied
entry-wise. For example, for (2):
\[
    (\bA+\bB)^T_{ij} = (\bA+\bB)_{ji} = a_{ji} + b_{ji} = (\bA^T)_{ij} + (\bB^T)_{ij}
    = (\bA^T + \bB^T)_{ij}.
\]

For (4), let $\mathbf{C} = \bA\bB \in \fR^{m \times p}$. By definition of transpose and
matrix product:
\[
    (\mathbf{C}^T)_{ij} = c_{ji} = \sum_{k=1}^{n} a_{jk}\,b_{ki}.
\]
Expanding $\bB^T\bA^T$ using $(\bB^T)_{ik} = b_{ki}$ and $(\bA^T)_{kj} = a_{jk}$:
\[
    (\bB^T\bA^T)_{ij} = \sum_{k=1}^{n} (\bB^T)_{ik}\,(\bA^T)_{kj} = \sum_{k=1}^{n} b_{ki}\,a_{jk}.
\]
Both sums are identical, so $(\bA\bB)^T = \bB^T\bA^T$.
}

\addCor{Reversal Law for Multiple Factors}{%
By induction on Thm~\ref{thm:transpose-props}(4):
\[
    (\bA_1 \bA_2 \cdots \bA_k)^T = \bA_k^T \cdots \bA_2^T \bA_1^T.
\]
}
\label{cor:reversal-law}

\addExcr{%
Refer to Def~\ref{def:transpose} and Thm~\ref{thm:transpose-props}.
Let $\bA = \begin{pmatrix} 1 & 2 & 3 \\ 4 & 5 & 6 \end{pmatrix}$.
\begin{enumerate}
    \item State the dimensions of $\bA$, $\bA^T$, $\bA\bA^T$, and $\bA^T\bA$.
    \item Compute $\bA^T$ explicitly.  Then compute $\bA\bA^T$ and verify it is symmetric
    using Def~\ref{def:symmetric}.
    \item In NumPy (0-based indexing), write the expression to retrieve element $a_{12}$
    (first row, second column in 1-based notation).  Then write a one-liner for $\bA\bA^T$.
    \item Is $\bA\bA^T = \bA^T\bA$ in general?  What can you say about each product
    individually?
\end{enumerate}
}

\addCor{Outer Product is Symmetric of Rank 1}{%
For any nonzero $\bx \in \fR^n$, the outer product $\bx\bx^T \in \fR^{n \times n}$ is
symmetric and has rank 1.  The inner product $\bx^T\bx \in \fR$ satisfies
$\bx^T\bx \ge 0$ with equality iff $\bx = \bzero$.
}
\label{cor:outer-product}

\addExcr{%
In this exercise, we will prove Cor~\ref{cor:outer-product}.
Refer to Def~\ref{def:transpose} and Thm~\ref{thm:transpose-props}.

Let $\bx = (x_1, \ldots, x_n)^T \in \fR^n$ be a nonzero column vector.
\begin{enumerate}
    \item The \textbf{outer product} is $\bx\bx^T \in \fR^{n \times n}$.  Write out entry
    $(i,j)$ explicitly.
    \item Use the Reversal Law (Thm~\ref{thm:transpose-props}(4)) to show $\bx\bx^T$ is
    symmetric.
    \item Show that $\bx\bx^T$ has rank 1 by arguing about the column space.
    (\textit{Hint: every column of $\bx\bx^T$ is a scalar multiple of $\bx$.})
    \item Express $\bx^T\bx$ as a sum of squares.  Conclude $\bx^T\bx \ge 0$ with equality
    iff $\bx = \bzero$, and identify the geometric quantity $\sqrt{\bx^T\bx}$.
\end{enumerate}
}

\end{document}
